### 2013-07-16 paper/manuscript.tex
Added annotations, manuscript & scripts.

\NeedsTeXFormat{LaTeX2e}[1995/12/01]
\documentclass[10pt]{bmc_article}

% Load packages
%\usepackage{hyperref}
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage{xspace}
%\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\urlstyle{rm}

\usepackage{rotating}
\usepackage{colortbl}
\usepackage{color}
\usepackage{comment}

\usepackage{subfigure}

\newcommand {\pg}[1]{\textcolor{blue}{#1}}
\newcommand {\ppg}[1]{\textcolor{blue}{#1}}

\newcommand{\minitab}[2][l]{\begin{tabular}{#1}#2\end{tabular}}

%\newcommand{}   {\mbox{}\xspace}

\def\squeezetable{\def\tabular@font{\tiny}}%

% Change useable area of a page to be slightly larger 
\setlength{\topmargin}{0.0cm}
\setlength{\textheight}{21.5cm}
\setlength{\oddsidemargin}{0cm} 
\setlength{\textwidth}{16.5cm}
\setlength{\columnsep}{0.6cm}

\newboolean{publ}

%Settings: comment\uncomment bmcformat definition to get format required

%Review style settings
\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings
%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}

% Begin ...
\begin{document}
\begin{bmcformat}

\title{Non-coding RNAs of bird genomes.}

\author{
Paul P. Gardner\correspondingauthor$^{1,2}$
\email{Paul P. Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
Sarah W. Burge$^3$
\email{sb30@sanger.ac.uk},
Maria Ninova$^4$
\email{Maria.Ninova@postgrad.manchester.ac.uk},
Stephanie Kehr$^5$
\email{steffi@bierdepot.bioinf.uni-leipzig.de},
Anne Nitsche$^5$
\email{anne@bierdepot.bioinf.uni-leipzig.de},
Mario Fasold$^5$
\email{mario@bierdepot.bioinf.uni-leipzig.de},
Tammy E. Steeves$^1$
\email{tammy.steeves@canterbury.ac.nz},
Sam Griffiths-Jones$^4$
\email{sam.griffiths-jones@manchester.ac.uk}
and
Peter Stadler\correspondingauthor$^5$
\email{Peter Stadler\correspondingauthor - studla@bioinf.uni-leipzig.de}
}
\address{
\iid(1) School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(2) Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(3) European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
\iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, Härtelstrasse 16-18, D-04107 Leipzig, Germany
}

\maketitle

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
        %% \paragraph*{Background:} Text for this section of the abstract.       
        %% \paragraph*{Results:} Text for this section of the abstract \ldots
        %% \paragraph*{Conclusions:} Text for this section of the abstract \ldots

Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase. Although we show
extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and
more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds,
we also demonstrate apparent ``losses'' in several RNA families. These
include the divergence of some classical ncRNAs and the loss of
several snoRNAs and microRNAs. In contrast, we show a significant
number of lncRNAs are surprisingly well conserved between birds and
mammals including several intriguing cases where the reported
mammalian lncRNA function is not conserved in birds. These combined
results illustrate the utility of applying homology based methods for
annotating vertebrate genomes and illustrate many complex evolutionary
patterns within the avian ncRNA cohort.
\end{abstract}

%% Non-coding RNAs in bird genomes

%% Paul P. Gardner1,2, Sarah W. Burge3, Jana Hertel5, Maria Ninova4, Stephanie Kehr5, Anne Nitsche5, Mario Fasold5, Tammy E. Steeves1, Sam Griths-Jones4, Peter Stadler5

%% 1 School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand. 2 Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand. 3 European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK. 4Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom. 5 Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, Härtelstrasse 16-18, D-04107 Leipzig, Germany

%% Talking points:
%% 1.     Although the majority of classical eukaryotic ncRNAs are conserved in birds, several notable classical ncRNAs are not. Homology searches in associated genes indicate these apparent "losses" are most likely due to divergence.
%% 2.     Of the 59 snoRNA conserved between humans and yeast, we find 45 are conserved in birds. Homology searches in associated genes for the remaining non-conserved snoRNAs indicate these apparent losses are indeed due to multiple losses in several bird lineages.
%% 3.     We find several miRNA families are specific to the avian lineage.

%% %% Draft Abstract
%% Here we present the results of a large-scale bioinformatic annotation of non-coding RNAs in 48 avian genomes. Our approach uses probabilistic models of hand-curated families from the Rfam database to infer conserved RNA families within each avian genome. We supplement these annotations with predictions from the tRNA annotation tool, tRNAscan-SE and microRNAs from miRBase. Although we show extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds, we also demonstrate apparent ``losses'' in several RNA families. These include the divergence of some classical ncRNAs and the loss of several snoRNAs and microRNAs. These combined results illustrate the utility of applying homology based methods for annotating vertebrate genomes and illustrate many complex evolutionary patterns within the avian ncRNA cohort.

%%%%%LNCRNA:

%% 3.     Several well-characterised lncRNA families of known function are conserved between mammals and birds.
%% 4.     We discuss the conservation of HOX associated lncRNAs and several lncRNAs that have been implicated in processes associated with cancer.
%% In contrast, we show a signi_cant number of lncRNAs are surprisingly well conserved between birds and mammals including several intriguing cases where the reported mammalian lncRNA function is not conserved in birds.

\ifthenelse{\boolean{publ}}{\begin{multicols}{2}}{}

\section*{Introduction}

Non-coding RNAs (ncRNAs) are an important class of genes. However,
they pose serious research challenges, particularly for the field of
genomics. They lack the strong statistical signals associated with
protein coding genes, e.g. open reading frames, G+C content and
codon-usage biases \cite{Rivas:2000}.

Non-coding RNAs are classified into hierarchical groupings by the Rfam
database. The basic units are ``families'' which are groups of
homologous, alignable sequences; ``clans'' which are groups of
un-alignable (or functionally distinct), homologous families; and
``classes'' which are groups of clans and families with related
biological functions e.g. spliceosomal RNAs, miRNAs and snoRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
The major RNA families include the classical, highly conserved RNAs,
sometimes called ``molecular fossils'', such as the transfer RNAs,
ribosomal RNAs, RNA components of RNase P and the signal recognition
particle \cite{Jeffares:1998}.  Other classes appear to have have
evolved more recently, e.g. the small nucleolar RNAs (snoRNAs),
microRNAs (miRNAs) and the long non-coding RNAs (lncRNAs)
\cite{Hoeppner:2012}. The publication of avian genomes provides
exciting opportunities to to explore ncRNA conservation in
unprecedented detail.

Despite promising techniques such as RNA-seq \cite{Croucher:2010},
homology based methods namely covariance models (CMs) are state of the
art for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families. These are used to train
probabilistic models for each family. These models can be used to
generate sequences with similar properties, score the likelihood that
a sequence is was generated by the same evolutionary processes as the
training sequences and to build alignments based upon sequence and
structural information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The tRNAscan-SE software package uses CMs to accurately predict
transfer RNAs \cite{Lowe:1997,Chan:2009}. The Rfam database contains
thousands of curated alignments and consensus structures for diverse
classes of ncRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches dramatically out-perform alternative methods
\cite{Freyhult:2007}.

The CM based approach works well for almost all classes of ncRNA, but
the long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. Recent technological advances have led to dramatic
speed and memory-usage enhancements for CM analyses
\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
cannot model the exon-intron structures of spliced lncRNAs, nor can they
deal simply with the repeats that many lncRNAs host. Consequently in
the latest release of Rfam the lncRNA families that were added were
composed of local conserved and structured elements within lncRNAs,
analogous to the ``domains'' housed within protein sequences
\cite{Burge:2013}. The functions determined to date for lncRNAs
range from regulating chromatin status to chromosomal inactivation
\cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
genes is a lengthy and expensive process \cite{Guttman:2009}.

%Still needs a {\bf what have we done} paragraph!

\section*{Results}

In the following we explore the conservation patterns of the major
classes of ncRNA in further detail. The collection of ncRNA sequences
is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}. Consequently the bulk of the
families discussed below are either conserved across the vertebrates
or have been apparently lost or diversified in the avian
lineage(s).

\subsection*{Classic RNAs: LUCA and LECA}

Many RNA families constitute the most evolutionarily conserved genes
across all life on this planet \cite{Jeffares:1998}. Examples of RNAs
derived from the Last Universal Common Ancestor (LUCA) include, the
transfer RNAs (tRNA), ribosomal RNAs (rRNA), RNA components of RNase P
(RNase P RNA), RNase MRP (RNase MRP RNA) and the signal recognition
particle (SRP RNA). Other classes of RNA are likely to have been
components of the Last Eukaryotic Common Ancestor (LECA) include the
telomerase RNA, major spliceosomal RNAs (U1, U2, U4, U5, and U6) and
the minor spliceosomal RNAs (U11, U12, U4atac, and U6atac)
\cite{Hoeppner:2012}.

%\subsubsection*{Losses}

Unsurprisingly, the bulk of these classes of RNAs are well represented
across the bird genomes (See Figure~\ref{fig:1}). However, there
appear to have been ``losses'' of a few of these RNAs in certain bird
species. Some of these may be due to sequence divergence, of which
there are several notable examples e.g.
\cite{Leonardi:2008,Webb:2008,Mao:,Lai:2010,Chan:2011}. Other losses
may be due to incomplete genome coverage.

A number of the classic RNAs are incorporated into RNA-protein
complexes (RNPs) involved in core cellular processes. An example of
this are the spliceosomal RNAs. Based upon the presence/absence
patterns of the major spliceosomal RNAs they are all well represented
in these genome sequences. Except for the U4 RNA in cormorant and the
U5 RNA in the bee eater. Again these two genomes are low coverage
suggesting these genes weren't captured in the current assembly. The
minor spliceosomal RNAs are more interesting, the U4atac and U11
snRNAs show widespread patterns of loss, even in some of the high
coverage genomes. These RNAs are frequently missed in bioinformatic
screens. Indicating either frequent loss \cite{Davila_Lopez:2008} or
sequences that have diverged beyond the ability of detection by
covariance models \cite{Marz:2008}.

The telomerase RNA is also largely missing from the avian
annotations. This RNA acts as a template for the telomerase enzyme
that extends the telomeres found on chromosome ends. It is only found
in the chicken, bald eagle, kea, budgerigar, crow and
zebrafinch. Homology searches searches with the telomerase reverse
transcriptase (TERT) protein show that the protein component of the
telomerase RNP is conserved across all the bird genomes (data not
shown). This pattern of presumably divergent telomerase RNA and
conserved telomerase protein has been noted previously, most notably
in the fungi \cite{Leonardi:2008,Webb:2008}.

The the RNA components of RNase P and RNase MRP also appear to have
undergone dramatic losses within the bird lineage. RNase P is required
for the maturation of tRNA, the paralogous enzyme, RNase MRP is
required for the maturation of rRNA. Each RNP cleaves smaller RNAs
from larger transcripts \cite{Lopez:2009}. It is unlikely that the
these genes have been lost in any of the birds. Homology searches with
the RNase associated protein coding genes (POP1, POP4, POP5, POP7,
RPP1, RPP14, RPP25, RPP38, RPP40 and RPR2), identified viable homologs
of each in all of the bird genomes \cite{Rosenblad:2006} (data not
shown). This suggests that the bird RNase P and MRP RNAs may have
diverged slightly from the canonical models.

The 5.8S component of the ribosome in the turtle, turkey bustard,
hoatzin, flamingo, tropicbird, seriema, owl, cuckoo roller, trogon,
bee eater and falcon appears to have been lost (See
Figure~\ref{fig:1}). However, the genomes for these species are also
low-coverage. Therefore gaps in these families are more likely to be
due to incomplete genome assemblies.

\subsection*{Small nucleolar RNAs}

Small nucleolar RNAs (snoRNAs) are important ncRNAs that participate
in the maturation of other functional RNAs \cite{Gardner:2010}. The
bulk of the characterised snoRNAs guide either methylation or
pseudouridylation modifications, primarily of rRNAs but also
spliceosomal RNAs. The two types of modifications are guided by two
different types of RNA, the C/D and the H/ACA box snoRNAs respectively,
each with a characteristic cohort of motifs and secondary structures
\cite{Marz:2011}.

There are 66 ribosomal modification sites, guided by 59 snoRNA
families, that are preserved between \emph{H. sapiens} and
\emph{S. cerevisiae} \cite{Lestrade:2006}. Of these, 45 snoRNA
families are conserved in the bird dataset.

Over a third of the apparent losses of the yeast-human conserved
snoRNA families appear to cluster on 2 locii of the ancestral
vertebrate genome. We investigated these losses further.

The first cluster is found at chr11:62620797-62622484 on the human
genome (hg19) and contains SNORD27, SNORD29 and SNORD31 of the
human-yeast conserved snoRNAs. These snoRNAs are located in the
inside-out gene SNHG1 which hosts a total of eight C/D box snoRNAs:
SNORD25, SNORD26, SNORD27, SNORD28, SNORD29, SNORD22, SNORD30 and
SNORD31 \cite{Tycowski:1996}. Each of which are also found in the
alligator and turtle genomes within a 3-4 KB locus, yet these have
largely been lost in the birds. Notable exceptions are five of the
eight are located in the tinamou genome, these are located on the same
scaffold and are within 2 KB of each other. Implying that SNHG1 is
conserved in the tinamou. Three of the eight are located in the
ostrich and zebrafinch genomes, again within 2 KB of each other on the
same scaffolds. This complex pattern of loss could be attributed to
many different models, e.g. multiple losses in birds, poor homology
modelling or incomplete genome sequences.

%%GAS5?
%SNORD81, SNORD47, SNORD80, SNORD79, SNORD78, SNORD44, SNORD77, SNORD76, SNORD75, SNORD74
%egrep 'SNORD81;|SNORD47;|SNORD80;|SNORD79;|SNORD78;|SNORD44;|SNORD77;|SNORD76;|SNORD75;|SNORD74' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'

The second cluster is located at chr19:49993222-49994231 on the human
genome (hg19) and contains two copies of SNORD33 and one SNORD34 all
within a 1 KB genomic region. The turtle and alligator genomes retain
the two copies of SNORD33 yet don't have an obvious SNORD34 gene on
the same scaffold. Within the bird genomes, the crow and rifleman each
retain a single SNORD33 and SNORD34 gene on the same scaffold. While
the ground-finch and zebrafinch retain a single SNORD33 and the bald
eagle and seriema retain a single SNORD34 (See Figure~\ref{fig:3}).
In human these snoRNAs are intronic to the host gene, ribosomal
protein L13a (RPL13A). Based on BLASTP (version 2.2.18) homology
searches for the RPL13A gene, the protein is conserved in the human
and turtle genomes and in the bald eagle, crow, rifleman and
zebrafinch avian genomes (data not shown). Therefore the conservation
patterns of the RPL13A host gene and corresponding intronic snoRNAs
largely mimic each other, each independently supports a pattern of
loss of the RPL13A gene and the intronic snoRNAs that it hosts in the
bird genomes.

\subsection*{MicroRNAs}

MicroRNAs are an important class of non-coding RNA. They have been
found in the genomes of Chromalveolata \cite{Cock:2010,Huang:2011},
Metazoa \cite{Lee:1993,Lau:2001,Hertel:2006}, Mycetozoa
\cite{Hinas:2007,Avesson:2012}, Viridiplantae
\cite{Reinhart:2002,Fattash:2007,Axtell:2007,Molnar:2007} and Viruses
\cite{Pfeffer:2004,Ouellet:2008,Pfeffer:2005,Landgraf:2007}.  The
miRNAs have been shown to regulate the expression of large numbers of
messenger RNAs \cite{Lim:2005}. The mature miRNA product is generally
22 nucleotides long which is usually processed from a larger RNA that
is characterised by a stable hairpin-shaped secondary structure.

Chicken and zebrafinch are the only birds with previously annotated
microRNAs.  We searched for homologs of these and other vertebrate
microRNAs in the genomes of the 48 birds, American alligator and green
turtle. Overall, we annotate a total of 16617 putative microRNA loci,
homologous to 543 known microRNA genes, of which 487 are annotated in
chicken and/or zebra finch, while 56 have been so far known only in
non-avian vertebrates. The numbers of annotated loci in the individual
species are approximately equal - 300-400 per species, except for the
turkey (Meleagris gallopavo) where we identified 543 sequences
homologous to known microRNAs.

Nearly all the microRNAs that are broadly conserved in fish,
amphibians and mammals are also conserved in the birds. Nevertheless,
there are obvious instances of microRNAs lost in all birds. For
example, mammalian, amphibian and fish genomes contain three loci of
clustered microRNAs from the mir-17 and mir-92 families.  One of these
clusters (mir-93~25) was not found in turtles, crocodiles and
birds. In addition, we can confidently identify a further 3 microRNA
families that are present in mammals, and turtle and/or crocodile, but
not in any avian genome (mir-150, mir-208, mir-590). This suggests
that these sequences were lost in the last common ancestor of
archosaurs or birds. There are also a number of microRNAs that are
predicted to be present in turtles and/or crocodiles, and only a small
number of bird genomes.  Indeed, there are many missing annotations,
species-specific and otherwise, that are not consistent with the
consensus phylogeny, and could be due to either incomplete genomes or
widespread microRNA loss.

The turkey genome contains a high number (190) of microRNAs so far
found only in chicken, which account for the higher number of
annotated sequences in this genome compared with other birds. This is
consistent with its phylogenetic position as the closest chicken
relative among the examined birds.  However, 101 chicken microRNAs
have no homolog in the turkey or other bird genomes, suggesting that
these genes are chicken-specific. This is consistent with previous
reports of large number of speciesspecific microRNAs in all animals,
and supports the view of fast microRNA turnover during animal
evolution.

\subsection*{Long non-coding RNAs}

As mentioned in the introduction, lncRNAs are a diverse group of RNAs
that have been implicated in a multitude of functional processes
\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have mainly been
characterised in mammalian species, particularly human and mouse.

In general Rfam cannot include the entire length of any large, spliced
RNAs. This is a limitation of the covariance-models used for the
homology-searches Rfam runs \cite{Nawrocki:2009}. Consequently, only
short, well-conserved regions with evolutionarily conserved secondary
structures are included in Rfam, we will refer to these as RNA-domains
\cite{Burge:2013}.

When analysing the RNA-domain annotations it is striking that many of
the lncRNAs with multiple RNA-domains appear to be consistently
preserved in the birds. The annotations of these domains are all in
the same genomic region supporting a high degree of evolutionary
conservation for the entire lncRNA. In particular the HOXA11-AS1,
PCA3, RMST, Six3os1, SOX2OT and ST7-OT3 lncRNAs have multiple well
conserved RNA-domains (See Figure~\ref{fig:5}).

The HOX cluster lncRNAs, HOTAIRM1, HOTTIP and HOXA11-AS1, are
remarkably well conserved. In the human genome the corresponding
RNA-domains are located at chr7:27135743-27245922 (hg19) and are
modelled by 15 RNA-domains in total, the % (5, 4 and 6 respectively)
lncRNAs are ordered HOTAIRM1 (5 RNA-domains) $\rightarrow$ HOXA11-AS1
(6 RNA-domains) $\rightarrow$ HOTTIP (4 RNA-domains). In the alligator
and turtle genomes five and six of the HOX RNA-domains are conserved,
each supporting an ancestral gene order of HOTAIRM1 $\rightarrow$
HOXA11-AS1 $\rightarrow$ HOTTIP. In the birds, where two or more of
the HOX cluster lncRNA RNA-domains were predicted on the same
scaffold, this gene order was preserved.

%egrep 'HOXA11-AS1|HOTAIR|HOTTIP' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'PCA3' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'RMST' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'

Many of the lncRNAs have been associated with cancer, sparking a minor
review industry \cite{Prensner:2011,Spizzo:2012}. Three examples of
these that are also conserved in the birds are described below.

The RMST (Rhabdomyosarcoma 2 associated transcript) RNA-domains 6, 7,
8, and 9 are conserved across the birds. In each bird the gene order
was also consistent with the human ordering. In the alligator and
turtle an additional RNA-domain was predicted in each, these were
RNA-domains 2 and 4 respectively, again the ordering of the domains
was consistent with human. This suggests that the RMAST lncRNA is
highly conserved. However, little is known about the function of this
RNA.  It was originally identified in a screen for differentially
expressed genes in two Rhabdomyosarcoma tumor types \cite{Chan:2002}.

In addition, the lncRNA DLEU2, is well conserved conserved across the
vertebrates, it is a host gene for two miRNA genes, miR-15 and miR-16,
both of which are also well conserved across the vertebrates (See
Figure~\ref{fig:3}). DLEU2 is thought to be a tumor-suppressor
gene as it is frequently deleted in malignant tumours \cite{Lerner:2009,Klein:2010}.

%(MIR1251 is in the same locus)
%PCA3 RNA-domains inverted relative to human...

The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
\cite{Xu:1997}. Both are expressed in a broad ranch of tissues. Extensive
research on BRCA1 has shown that it is involved in DNA repair
\cite{}. The function of NBR2 remains unknown, yet its conservation
across the vertebrates certainly implies a function (See
Figure~\ref{fig:5}).

%(neighbor of BRCA1 gene 2) 
%MHM results!!!!

%% Another of the functions associated with lncRNAs is chromosomal dosage
%% compensation. In mammals this is carried out by the lncRNAs Xist and
%% Tsix \cite{}. In birds it has been proposed that the lncRNA MHM might
%% play a related role \cite{}.

%DLEU2, frequently deleted in malignancy, functions as a critical host
%gene of the cell cycle inhibitory microRNAs miR-15a and miR-16-1

%% TUMOR-SUPPRESSOR:
%% DLEU2, part of a tumor-suppressor cluster (with two microRNAs), most recent refs here:
%% http://www.ncbi.nlm.nih.gov/pubmed/19591824
%% http://www.ncbi.nlm.nih.gov/pubmed/20060366
%% SMAD5, tumor suppressor candidate:
%% http://www.ncbi.nlm.nih.gov/pubmed/10049768?dopt=Abstract

%% DEVELOPMENT:
%% HOXA11, most complete reference list here:
%% http://www.lncrnadb.org/Detail.aspx?TKeyID=171
%% Six3OS, retinal development:
%% http://www.ncbi.nlm.nih.gov/pubmed/21936910
%% ZEB2, brain development:
%% http://www.ncbi.nlm.nih.gov/pubmed/18347095?dopt=Abstract

%% MAT2A:
%% http://www.ncbi.nlm.nih.gov/pubmed/1426236%2C9703951?dopt=Abstract

%% RMST, doesn't look we know it does (also shows up in review paper at top):
%% http://www.ncbi.nlm.nih.gov/pubmed/20062813

%% ST7-OT, another candidate:
%% http://www.ncbi.nlm.nih.gov/pubmed/12213198?dopt=Abstract

%% PART1, meant to be human prostate-specific, birds don't even have a prostate:
%% http://www.ncbi.nlm.nih.gov/pubmed/10706094?dopt=Abstract
%% PCA3, prostate/androgen receptor signalling (curious there's heaps of it in male and female birds? same for PART1?):
%% http://www.ncbi.nlm.nih.gov/pubmed/?term=ferreira+and+pca3

\subsection*{Cis-regulatory elements}
%% Histone3, IRE, IRES, K chan RES, SECIS, Antizyme FSE, Vimentin3, GABA3,
%% CAESAR,

The cis-regulatory RNAs are a group of RNA structures encoded on
mRNAs. Generally they are involved in regulating the expression of the
mRNA they are encoded within. Others may recode the translated protein
product into an alternate sequence.

This group includes the iron response element (IRE) \cite{Stevens:}
and the histone $3^\prime$ UTR (histone3)
\cite{Davila_Lopez:2008a}. These are structured motifs bound by
regulatory proteins. The selenocysteine insertion sequence (SECIS) is
a structured motif that recodes UGA stop codons to selenocysteines
\cite{Lambert:2002} and the GABRA3 stem-loop is a structure recognised
by the ADAR enzyme family. This enzyme edits adenine nucleotides to
inosine, in this case recoding an isoleucine codon to methionine in
exon 9 of the GABRA3 gene \cite{Ohlson:2007}.

These regulatory elements and others, including an internal ribosome
entry site (IRES), potassium channel RNA editing signal (K chan RES),
Antizyme RNA frameshifting stimulation element (Antizyme FSE),
vimentin $3^\prime$ UTR protein-binding region (Vimentin3) and a
connective tissue growth factor (CTGF) $3^\prime$ UTR element (CAESAR)
are conserved across a diverse group of vertebrates, including the
bird lineages explored here (See Figure~\ref{fig:1}).

\subsection*{Exceptional RNAs}

%Background to this group...

A number of other ncRNAs can also be found in Eukaryotic
genomes. These do not fit into the main classes of RNA but still
perform vital roles in the function and evolution of Eukaryotes.
Their functions are diverse and many have not yet been characterised.

An example of an uncharacterised RNA is the ultraconserved element,
uc.338 (also known as TUC338)
\cite{Bejerano:2004,Bejerano:2006,Braconi:2011}.  The uc.338 element
is derived from a short interspersed element (SINE) called the
lobe-finned fishes SINE (LF-SINE) as it is conserved between the
coelacanth and mammals \cite{Bejerano:2006}.  Analysis of the
expression of uc.338 implies that it plays a role in the progression
of hepatocellular carcinoma, possibly by influencing cell growth
\cite{Braconi:2011}. This RNA is conserved in the birds and appears to
have been duplicated in several lineages.

The Y RNA is an enigmatic ncRNA where we know very little about the
function. It was discovered in the 1980s in ribonucleoprotein
complexes \cite{Lerner:1981}. The function of the Y RNAs remain
unknown, but evidence is emerging that they may be associated with DNA
replication \cite{Christov:2006}.  There are 4 functional Y RNAs
encoded in the human genome, Y1, Y3, Y4 and Y5. However, there are
hundreds of pseudogenised copies of the Y RNA scattered throughout the
human genome \cite{Mosig:2007}. In the birds and other lizards, we
identify between two and seven Y RNA paralogs (See
Figure~\ref{fig:1}).

The Vault RNA forms a major component of the vault ribonucleoprotein
complex, this is one of the largest particles found in the vertebrate
cell, it could envelope the ribosome \cite{Kong:1999}. As yet

\cite{Stadler:2009}

 Vault, 7SK, HAR1A

\subsection*{Pseudogenes}

The human genome is littered with decaying, pseudogenised copies of
ncRNAs. The Dfam database of human repeats estimates that
approximately $54\%$ of the human genome is composed of repeat
elements \cite{Wheeler:2013}. Of these annotations XXX\% are derived
from ncRNAs, including the large Alu family of repeats, which is
derived from a retransposed SRP RNA pseudogene \cite{}.

tRNA, RSV RNA, 
 \cite{International_Chicken_Genome_Sequencing_Consortium:2004}

\emph{Endogenous retroviruses:} RSV RNA, Alfamo CPB

\subsection*{Contamination}

Bacterial families can be used to identify problematic sequences that
are likely to be the result of contamination from non-avian sources.

\section*{Conclusions}

%Talking points: 
%1. several classical RNAs (e.g. spliceosomal, telomerase, RNase MRP
%RNAs) have diversified in birds relative to other Eukaryotes.
%2. several lncRNA families are conserved between mammals and
%birds. Several of these are characterised by expression in tissues
%that are not present in birds (e.g. the prostate).
%3. several regulatory RNAs, snoRNAs and microRNAs, have been lost
%since the birds diverged from the other vertebrates. Indicating many
%avian-specific alterations in gene regulation.

\section*{Methods}

Bird genomes were searched using the cmsearch program from INFERNAL
1.1 and the covariance models from the Rfam database
v11.0\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}. All
matches above the curated GA threshold were included. Subsequently,
all hits with an E value greater than 0.0005 were discarded, so only
matches which passed the model-specific GA threshold and had an E-value smaller than 0.0005 were retained.

tRNA-scan version 1.3.1, ... \cite{Lowe:1997,Chan:2009}

Rfam matches and the tRNA-scan results for families belonging to the
same clan were then competed so that only the best match was retained.
Matches were further sorted into taxonomic groups {\bf ONLY MATCHES IN 5 OR MORE ...}

999 microRNA sequence families, previously annotated in at least one
vertebrate, were retrieved from miRBase (v19). Individual sequences or
multiple sequence alignments were used to build covariance models with
INFERNAL (v1.1rc3), and these models were searched against the 48 bird
genomes, and the genomes of the american alligator and the green
turtle as outgroups.  Hits with e-value <10 realigned with the query
sequences and the resultant multiple sequence alignments manually
inspected and edited using RALEE.

\section*{Acknowledgements}

Erich Jarvis, Guojie Zhang and Tom Gilbert. Magnus Alm Rosenblad.

%%%%%%%%%%%%%%%%%%%%%%
%% The Bibliography %%
%%
%% ---------------------------------
%% BioMedCentral bibtex .BST file will be used to
%% create a .BBL file which includes the BMC XML.
%% Note that the displayed Bibliography will not be 
%% exactly as specified in the online Instructions for Authors

{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
\bibliographystyle{bmc_article}  % Style BST file
 \bibliography{bird} }     % Bibliography file (usually '*.bib' )

%%%%%%%%%%%

\clearpage
\newpage

\begin{figure}[ht]
  \includegraphics[width=0.4\textwidth]{figures/RNA.pdf}  
  \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
  \includegraphics[width=0.4\textwidth]{figures/snoRNA-human-yeast-correspondences.pdf}
  \caption[]{This figure illustrates the number of copies of Rfam and
    tRNAscan annotations for each RNA family in each avian species and
    outgroups. The families are ordered alphabetically vertically and
    the species are in phylogenetic order horizontally. Darker shades
    indicate high copy numbers, lighter shades indicate fewer, white
    indicates no predictions were made.}\label{fig:1}
\end{figure}

%% \begin{figure}[ht]
%%   \includegraphics[width=0.65\textwidth]{figures/snoRNA-human-yeast-correspondences.pdf}
%%   \caption[]{.}\label{fig:2}
%% \end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.5\textwidth]{figures/miRNA.pdf}
  \includegraphics[width=0.45\textwidth]{figures/snoRNA.pdf}
  \caption[]{.}\label{fig:3}
\end{figure}

\begin{figure}[ht]
  \includegraphics[width=0.45\textwidth]{figures/lncRNA.pdf}
  \includegraphics[width=0.45\textwidth]{figures/tRNA.pdf}
  \caption[]{.}\label{fig:5}
\end{figure}

\end{bmcformat}
\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2013-07-16 scripts/compete_clans.pl

@@ -1,0 +1,308 @@
+#!/usr/bin/perl 
+
+#Given multiple gff files, merge them together and combine entries with the same coordinates. 
+
+use warnings;
+use strict;
+use Getopt::Long;
+use IO::File;
+
+my (@gffs, $clanInfo, $verbose, $help);
+my $overlapThreshold=10.0;
+my $overlapDir=".";
+&GetOptions( 
+    "ot|overlapthresh=s"  => \$overlapThreshold,
+    "g|gff=s@"            => \@gffs,
+    "cl|claninfo=s"       => \$clanInfo,
+    "od|overlapdir=s"     => \$overlapDir,
+    "v|verbose"           => \$verbose,
+    "h|help"              => \$help
+    );
+
+if( $help ) {
+    &help();
+    exit(1);
+}
+elsif (@gffs == 0){
+    print "FATAL: no gff files given\n";
+    &help();
+    exit(1);
+}
+
+my $gffs = join(' ', @gffs);
+
+my %clans;
+if(defined($clanInfo) && (-s $clanInfo)){
+    open(CL, "< $clanInfo") or die "FATAL: failed to open [$clanInfo]\n[$!]"; 
+    while(my $cl=<CL>){
+	my @cl=split(/\t/, $cl);
+	#rfamacc        clanacc
+	$clans{$cl[2]}=$cl[0];
+    }
+    close(CL);
+}
+else{
+    print "WARNING: no Clan information! (which annotations should be merged?)\n"
+}
+
+my $fh = IO::File->new();
+# sort features so they can be merged in coordinate
+# order
+$fh->open( "sortGffs.pl $gffs | uniq | " );
+print "#running [sortGffs.pl $gffs |]\n" if(defined $verbose);
+#$gffs = join('_', @gffs);
+open(OV, "> $overlapDir/$$\_unclassified_overlaps_file.overlaps");
+print OV "#GFFS: [$gffs]\n"; 
+my ($cur, $prev, $ext, $rfid, $prfid, %uniq);
+my $cnt=0;
+while($cur = <$fh>) {
+    chomp($cur); 
+    next if ($cur =~ /^#/);
+    if(defined($prev)){
+	my @cur  = split(/\t/,  $cur);
+	my @prev = split(/\t/, $prev);
+	next if (not defined $cur[3] || not defined $cur[4] || not defined $cur[6]); #not a GFF
+	$rfid="undef";	
+	if($cur[8]=~/rfam-acc=(RF\d+)/){
[…]

### 2013-07-16 scripts/gffs2heatmaps.pl

@@ -1,0 +1,456 @@
+#!/usr/bin/perl 
+
+#1. Filter contaminants
+#2. Select families conserved in >10% of birds
+##   --print to GFF
+#3. Create Heatmaps
+#4. Analyse repeated/pseudogene families
+#     --"calibrate" with tRNAscan predictions
+#     
+
+#split the tRNAs out. 
+
+use warnings;
+use strict;
+use Getopt::Long;
+#use Statistics::Descriptive;
+
+
+my ($gffsDir, $outDir, $rDir) = ("data/merged-annotations","data/conserved-merged-annotations", "data/R");
+
+my ($verbose, $help);
+&GetOptions( 
+    "gd|gffDir=s"         => \$gffsDir,
+    "od|outDir"           => \$outDir,
+    "rd|rDir"             => \$rDir,
+    "v|verbose"           => \$verbose,
+    "h|help"              => \$help
+    );
+
+my @gffs = glob("$gffsDir/*gff"); 
+
+if( $help ) {
+    &help();
+    exit(1);
+}
+elsif ((not defined($gffsDir)) or (@gffs)==0){
+    print "FATAL: no gff files given\n";
+    &help();
+    exit(1);
+}
+
+my %whitelist; 
+if (-s "$rDir/allRNA.dat"){
+    open(W, "< $rDir/allRNA.dat");
+    while(my $w=<W>){
+	chomp($w);
+	my @w=split(/\t/, $w);
+	$whitelist{$w[50]}=1;
+    }
+    close(W);
+}
+
+my %lncRNAlist;
+if (-s "data/rfam11_lncRNAs.txt"){
+    open(W, "< data/rfam11_lncRNAs.txt");    
+    while(my $w=<W>){
+	chomp($w);
+	$lncRNAlist{$w}=1;
+    }
+    close(W);
+}
+
+my %blacklist=(
+    mraW=>1,    
+    );
+
+print "Reading GFFs\n" if (defined($verbose));
+my (%HoHspeciesRfamCount, %familiesSpeciesCounts, %familiesTotalCounts, %statistics, @bitscores, %tRNA, %tRNAfamilies); 
[…]

### 2013-07-16 scripts/heatmaps.R

@@ -1,0 +1,104 @@
+#R CMD BATCH --no-save heatmaps.R
+
+library(gplots)
+breaks<-c(-1000,0,1,5,10,50,100,500,1000)
+lb<-length(breaks)-2
+
+lmat = rbind(c(1,2),c(3,4))
+lwid = rbind(c(4,.5))
+lhei = rbind(c(4,.25))
+
+key<-c(0,1,5,10,50,100,500,1000)
+key<-matrix(c(key,key),ncol=2,nrow=8,dimnames = list(key,c("","")), byrow = FALSE)
+pdf(file="paper/figures/key.pdf",width=9, height=9)
+heatmap.2(key,trace="none",col=gray(lb:0/lb),
+          dendrogram = "none", scale = "none",
+          breaks=breaks,
+          sepcolor="black", Rowv=F, Colv=F, key = F,
+          lmat=lmat,
+          cexCol=1.1, cexRow=1.1
+          )
+dev.off()
+
+rna<-read.table("data/R/RNA.dat",header = T, sep = "\t",row.names=52)
+pdf(file="paper/figures/RNA.pdf",width=8, height=16)
+heatmap.2(as.matrix(rna),trace="none",col=gray(lb:0/lb),
+          dendrogram = "none", scale = "none",
+          breaks=breaks,
+          sepcolor="black", Rowv=F, Colv=F, key = F,
+          lmat=lmat, lwid=lwid, lhei=lhei,
+          cexCol=1.1, cexRow=1.1
+          )
+dev.off()
+
+
+
+mirna<-read.table("data/R/miRNA.dat",header = T, sep = "\t",row.names=52)
+pdf(file="paper/figures/miRNA.pdf",width=8, height=50)
+heatmap.2(as.matrix(mirna),trace="none",col=gray(lb:0/lb),
+          dendrogram = "none", scale = "none",
+          breaks=breaks,
+          sepcolor="black", Rowv=F, Colv=F, key = F,
+          lmat=lmat, lwid=lwid, lhei=lhei,
+          cexCol=1.1, cexRow=1.1
+          )
+dev.off()
+
+
+snorna<-read.table("data/R/snoRNA.dat",header = T, sep = "\t",row.names=52)
+pdf(file="paper/figures/snoRNA.pdf",width=8, height=25)
+heatmap.2(as.matrix(snorna),trace="none",col=gray(lb:0/lb),
+          dendrogram = "none", scale = "none",
+          breaks=breaks,
[…]

### 2013-07-16 scripts/sortGffs.pl

@@ -1,0 +1,100 @@
+#!/usr/bin/perl -w
+
+use strict;
+use File::Basename;
+
+my $prog = basename($0);
+my $usage .= "$prog - sort one or many GFF files, using a specified sort method.";
+$usage .= "  Default sort is by Name, Start, and End fields, in ascending order.\n";
+$usage .= "\n";
+$usage .= "Usage: $prog [options] <file1> <file2> ...";
+$usage .= "\n";
+$usage .= "Options:\n";
+$usage .= "   -maxSc: sort by maximum score, in descending order\n";
+$usage .= "   -minSc: sort by minimum score, in ascending order\n";
+$usage .= "   -flen: sort by feature length, in ascending order\n";
+$usage .= "   -userdef: sort by user defined expression; use single quotes\n"; 
+
+my $sortMethod = "byNSE";
+my $sortExpr = undef;
+
+while (@ARGV) 
+{
+    last unless $ARGV[0] =~ /^-./;      # Loop thru all the command line options.
+    my $opt = shift;
+    if ($opt eq "-maxSc") 
+    { 
+      $sortMethod = "byMaxScore"; 
+    }
+    elsif ($opt eq "-minSc") 
+    { 
+      $sortMethod = "byMinScore"; 
+    }
+    elsif ($opt eq "-flen") 
+    { 
+      $sortMethod = "byFeatureLen"; 
+    }
+    elsif ($opt eq "-userdef") 
+    { 
+      $sortMethod = "byexpr";
+      $sortExpr = shift || die "$usage\nMissing sort expression.\n";
+    }
+    else 
+    { 
+      die "$usage\nUnknown option: $opt\n" 
+    }
+}
+ 
+if (@ARGV==0)
+{
+  die $usage;
+}
+
+my (@line, @name, @type, @start, @end, @score);
+
+while (<>) { # Read lines from file(s) specified on command line. Store in $_.
+    s/#.*//; # Remove comments from $_.
+    next unless /\S/; # \S matches non-whitespace.  If not found in $_, skip to next line.
+    my @f = split /\t/; # split $_ at tabs separating fields.
+    push @line, $_;    # complete line
+    push @name, $f[0]; # name field
+    push @type, $f[2]; # type field
+    push @start, $f[3]; # start field
[…]

### 2013-07-16 scripts/gffs2heatmaps.pl
Updated README.

@@ -131,113 +131,120 @@ foreach my $f (@gffs){
 #WHERE IS ANTROSTOMUS (CHUCK WILLS WIDOW) & MELOPSITTACUS (BUDGIE)?!!!
 #THIS IS CALLED Caprimugus_Carolinensis BELOW FOR SOME REASON?!
 #####renamed to Antrostomus_carolinensis
-my @speciesPhyloOrder = qw(
-Homo_sapiens
-Alligator_mississippiensis
-Chelonia_mydas
-Struthio_camelus
-Tinamus_guttatus
-Anas_platyrhynchos
-Meleagris_gallopavo
-Gallus_gallus
-Antrostomus_carolinensis
-Chaetura_pelagica
-Calypte_anna
-Columba_livia
-Cuculus_canorus
-Mesitornis_unicolor
-Pterocles_guturalis
-Tauraco_erythrolophus
-Chlamydotis_undulata
-Opisthocomus_hoazin
-Charadrius_vociferus
-Balearica_regulorum
-Podiceps_cristatus
-Phoenicopterus_ruber
-Eurypyga_helias
-Phaethon_lepturus
-Gavia_stellata
-Egretta_garzetta
-Nipponia_nippon
-Pelecanus_crispus
-Phalacrocorax_carbo
-Fulmarus_glacialis
-Pygoscelis_adeliae
-Aptenodytes_forsteri
-Cariama_cristata
-Tyto_alba
-Cathartes_aura
-Haliaeetus_albicilla
-Haliaeetus_leucocephalus
-Colius_striatus
-Leptosomus_discolor
-Apaloderma_vittatum
-Buceros_rhinoceros
-Merops_nubicus
-Picoides_pubescens
-Falco_peregrinus
-Nestor_notabilis
-Melopsittacus_undulatus
-Acanthisitta_chloris
-Manacus_vitellinus
-Corvus_brachyrhynchos
-Taeniopygia_guttata
-Geospiza_fortis
-);
+my (@speciesPhyloOrder,@speciesCommonPhyloOrder); 
 
-my @speciesCommonPhyloOrder = qw(
-human
-alligator
-turtle
-ostrich_85X
-tinamou_100X
-duck_50X
-turkey
-chicken
-nightjar_30X
-swift_106X
-hummingbird_110X
-pigeon_64X
-cuckoo_100X
-mesite_29X
-sandgrouse_25X
-turaco_30X
-bustard_27X
-hoatzin_100X
[…]

### 2013-07-16 scripts/gffs2heatmaps.pl
Updated gffs2heatmaps.pl. Added overlaps, conserved-merged-annotations & misc accessory files.

@@ -45,11 +45,13 @@ if (-s "$rDir/allRNA.dat"){
     while(my $w=<W>){
 	chomp($w);
 	my @w=split(/\t/, $w);
-	$whitelist{$w[50]}=1;
+	my $w=pop(@w);
+	$whitelist{$w}=1;
     }
     close(W);
 }
 
+#replace with a family2type hash:
 my %lncRNAlist;
 if (-s "data/rfam11_lncRNAs.txt"){
     open(W, "< data/rfam11_lncRNAs.txt");    
@@ -68,183 +70,75 @@ print "Reading GFFs\n" if (defined($verbose));
 my (%HoHspeciesRfamCount, %familiesSpeciesCounts, %familiesTotalCounts, %statistics, @bitscores, %tRNA, %tRNAfamilies); 
 foreach my $f (@gffs){
 
-    #print "$f\n";
+    print "$f\n" if(defined($verbose));
     my $species;
     if ($f=~/$gffsDir\/(\S+?_\S+?)\.gff/){
 	$species=$1; 
     }
+    else{
+	print "WARNING: no species name parsed from [$f], skipping!\n";
+	next;
+    }
     
     open(GFF, "< $f");
+    open(OUT, "> $outDir/$species\.gff");
     while(my $g=<GFF>){
 	chomp($g);
 	my @g=split(/\t/, $g);
-	my $type; 
-	if($g[8]=~/rfam-id=(\S+);evalue/ or $g[8]=~/Alias=(\S+);Note/){
-	    $type=$1;
-	    #push(@bitscores, $g[5]) if (defined($whitelist{$type}) && $g[5]=~/^\d+\.\d+$/); 
+	my $family; 
+	if($g[8]=~/rfam-id=(\S+);evalue/ or $g[8]=~/Alias=(\S+);Note/){#Rfam
+	    $family=$1;
+	    #push(@bitscores, $g[5]) if (defined($whitelist{$family}) && $g[5]=~/^\d+\.\d+$/); 
 	}
-	elsif($g[8]=~/ID=(\S+)\_\d+/){
-	    $type=$1;
+	elsif($g[8]=~/ID=(\S+)\_\d+/){#miRBase
+	    $family=$1;
 	}
-	elsif($g[8]=~/type=Pseudo;/){
-	    $type="tRNA-pseudogene";
+	elsif($g[8]=~/type=Pseudo;/){#tRNAscan
[…]

### 2013-07-19 paper/manuscript.tex
Writing, writing.

@@
 
-\title{Non-coding RNAs of bird genomes.}
+\title{Non-coding RNAs of bird genomes}
 
@@
 
+{\bf MAY BE MOVED INTO THE lncRNA PAPER!}
+
 As mentioned in the introduction, lncRNAs are a diverse group of RNAs
@@
 complex, this is one of the largest particles found in the vertebrate
-cell, it could envelope the ribosome \cite{Kong:1999}. As yet
+cell; In fact, it is larger than the ribosome \cite{Kong:1999}. As yet
+not much is known about the function of Vault. The Vault RNA has been
+shown to be broadly conserved in metazoans \cite{Stadler:2009}.
 
-\cite{Stadler:2009}
-
- Vault, 7SK, HAR1A
+% Vault, 7SK, HAR1A
 
@@
 \begin{figure}[ht]
-  \includegraphics[width=0.5\textwidth]{figures/miRNA.pdf}
-  \includegraphics[width=0.45\textwidth]{figures/snoRNA.pdf}
+  \includegraphics[height=0.95\textheight]{figures/miRNA.pdf}
+  \includegraphics[height=0.95\textheight]{figures/snoRNA.pdf}
   \caption[]{.}\label{fig:3}

### 2013-07-29 paper/manuscript.tex
Updated Gallus annotations to galGal4.
Updated manuscript.

@@
 \email{Paul P. Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
+Jana Hertel$^5$
+\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
 Sarah W. Burge$^3$
@@
 \email{steffi@bierdepot.bioinf.uni-leipzig.de},
-Anne Nitsche$^5$
-\email{anne@bierdepot.bioinf.uni-leipzig.de},
 Mario Fasold$^5$
@@
 
-%% Paul P. Gardner1,2, Sarah W. Burge3, Jana Hertel5, Maria Ninova4, Stephanie Kehr5, Anne Nitsche5, Mario Fasold5, Tammy E. Steeves1, Sam Griths-Jones4, Peter Stadler5
+%% Paul P. Gardner1,2, Sarah W. Burge3, Jana Hertel5, Maria Ninova4, Stephanie Kehr5, Anne Nitsche5, Mario Fasold5, Tammy E. Steeves1, Sam Griffiths-Jones4, Peter Stadler5
 
@@
 Homology searches in associated genes for the remaining non-conserved snoRNAs indicate these apparent losses are indeed due to multiple losses in several bird lineages.
-%% 3.     We find several miRNA families are specific to the avian lineage.
+%% 3.     We find that 3 families of broadly conserved vertebrate microRNA families are lost in the avian lineage.  We also annotate a number of bird-specific microRNAs.
 
@@
 
-%%%%%LNCRNA:
+%% LNCRNA:
 
@@
 modelling or incomplete genome sequences.
+
+%Furthermore our results contradict the text in the paper. We find
+%SNORD26 stated to be lost in sauropsids in the 5 species:
+%C.vociferus, M.unicolor, H.leucocephalus, L.discolor and T.guttata.
 
@@
 not much is known about the function of Vault. The Vault RNA has been
-shown to be broadly conserved in metazoans \cite{Stadler:2009}.
+shown to be broadly conserved in metazoans
+\cite{Stadler:2009}. However, in the bird lineages
 

### 2013-08-12 paper/manuscript.tex
Updated manuscript.

@@
 %% 3.     We find that 3 families of broadly conserved vertebrate microRNA families are lost in the avian lineage.  We also annotate a number of bird-specific microRNAs.
+%% 4.     Several well-characterised lncRNA families of known function are conserved between mammals and birds.
+%% 5.     We discuss the conservation of HOX associated lncRNAs and several lncRNAs that have been implicated in processes associated with cancer.
 
 %% %% Draft Abstract
-%% Here we present the results of a large-scale bioinformatic annotation of non-coding RNAs in 48 avian genomes.
-Our approach uses probabilistic models of hand-curated families from the Rfam database to infer conserved RNA families within each avian genome.
-We supplement these annotations with predictions from the tRNA annotation tool, tRNAscan-SE and microRNAs from miRBase.
+%% We present the results of a large-scale bioinformatic annotation of non-coding RNAs in 48 avian genomes.
+We use probabilistic models of hand-curated families from the Rfam database to infer conserved RNA families within each avian genome.
+We supplement these annotations with predictions from the tRNA annotation tool, tRNAscan-SE, microRNAs from miRBase and selected sequences from analyses of the chicken transcriptome.
 Although we show extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds, we also demonstrate apparent ``losses'' in several RNA families.
-These include the divergence of some classical ncRNAs and the loss of several snoRNAs and microRNAs.
+These include the divergence of some classical ncRNAs and the genuine loss of several snoRNAs and microRNAs.
+In contrast, we show a signicant number of lncRNAs are surprisingly well conserved between birds and mammals including several intriguing cases where the reported mammalian lncRNA function is not conserved in birds.
 These combined results illustrate the utility of applying homology based methods for annotating vertebrate genomes and illustrate many complex evolutionary patterns within the avian ncRNA cohort.
@@
 %% LNCRNA:
-
-%% 3.     Several well-characterised lncRNA families of known function are conserved between mammals and birds.
-%% 4.     We discuss the conservation of HOX associated lncRNAs and several lncRNAs that have been implicated in processes associated with cancer.
-%% In contrast, we show a signi_cant number of lncRNAs are surprisingly well conserved between birds and mammals including several intriguing cases where the reported mammalian lncRNA function is not conserved in birds.
 
@@
 
-The 5.8S component of the ribosome in the turtle, turkey bustard,
+The 5.8 S component of the ribosome in the turtle, turkey bustard,
 hoatzin, flamingo, tropicbird, seriema, owl, cuckoo roller, trogon,
 bee eater and falcon appears to have been lost (See
-Figure~\ref{fig:1}). However, the genomes for these species are also
-low-coverage. Therefore gaps in these families are more likely to be
-due to incomplete genome assemblies.
+Figure~\ref{fig:1}). The rRNA repeats are frequently not assembled,
+consequently it is not surprising to see ``losses'' in these \cite{Floutsakou:2013}.
+Furthermore, the genomes for these species are also
+low-coverage.
 
@@
 shown to be broadly conserved in metazoans
-\cite{Stadler:2009}. However, in the bird lineages
+\cite{Stadler:2009}. However, in the bird lineages it appears to have
+either been lost or diversified.
 
@@
 
-The human genome is littered with decaying, pseudogenised copies of
-ncRNAs. The Dfam database of human repeats estimates that
-approximately $54\%$ of the human genome is composed of repeat
-elements \cite{Wheeler:2013}. Of these annotations XXX\% are derived
-from ncRNAs, including the large Alu family of repeats, which is
-derived from a retransposed SRP RNA pseudogene \cite{}.
+The human genome is littered with pseudogenised copies of ncRNAs. The
+Dfam database of human repeats estimates that approximately $54\%$ of
+the human genome is composed of repeat elements
+\cite{Wheeler:2013}. Of these annotations XXX\% are derived from
+ncRNAs, including the large Alu family of repeats, which is derived
+from a retransposed SRP RNA pseudogene \cite{}.
 
@@
   \includegraphics[height=0.95\textheight]{figures/snoRNA.pdf}
-  \caption[]{.}\label{fig:3}
+  \caption[]{Heatmaps showing the prescence/abscence and approximate
+    copy number of {\bf miRNA} families on the right and {\bf snoRNA} families on
+    the left.}\label{fig:3}
 \end{figure}
@@
   \includegraphics[width=0.45\textwidth]{figures/tRNA.pdf}
-  \caption[]{.}\label{fig:5}
+  \caption[]{Heatmaps showing the prescence/abscence and approximate
+    copy number of {\bf lncRNA} families on the right and {\bf tRNA}
+    families on the left.}\label{fig:5}
 \end{figure}

### 2013-08-27 paper/manuscript.tex
Added edits by Jana.

@@
 \usepackage{xspace}
-%\usepackage[utf8]{inputenc} %unicode support
+\usepackage[utf8]{inputenc} %unicode support
 %\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
+%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
 \urlstyle{rm}
+\usepackage[OT1]{fontenc}
 
@@
 \newcommand {\ppg}[1]{\textcolor{blue}{#1}}
+\newcommand {\new}[1]{\textcolor{green}{#1}}
 
@@
 \iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
-\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, Härtelstrasse 16-18, D-04107 Leipzig, Germany
+\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany
 }
@@
 \begin{abstract}
-        % Do not use inserted blank lines (ie \\) until main body of text.
-        %% \paragraph*{Background:} Text for this section of the abstract.       
-        %% \paragraph*{Results:} Text for this section of the abstract \ldots
-        %% \paragraph*{Conclusions:} Text for this section of the abstract \ldots
-
 Here we present the results of a large-scale bioinformatic annotation
@@
 4Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
-5 Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, Härtelstrasse 16-18, D-04107 Leipzig, Germany
+5 Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, Hartelstrasse 16-18, D-04107 Leipzig, Germany
 
@@
 \section*{Results}
-
 In the following we explore the conservation patterns of the major
@@
 spliceosomal RNAs. The two types of modifications are guided by two
-different types of RNA, the C/D and the H/ACA box snoRNAs respectively,
-each with a characteristic cohort of motifs and secondary structures
-\cite{Marz:2011}.
+different types of RNA, the box C/D and the H/ACA snoRNAs
+respectively, each with a characteristic cohort of motifs and
+secondary structures \cite{Marz:2011}.
 
@@
 families are conserved in the bird dataset.
-
 Over a third of the apparent losses of the yeast-human conserved
@@
 scaffold and are within 2 KB of each other. Implying that SNHG1 is
-conserved in the tinamou. Three of the eight are located in the
-ostrich and zebrafinch genomes, again within 2 KB of each other on the
-same scaffolds. This complex pattern of loss could be attributed to
-many different models, e.g. multiple losses in birds, poor homology
-modelling or incomplete genome sequences.
+conserved in the tinamou. Loci with four of the eight snoRNAs can be
+found in zebrafinch, ground-finch, and bald eagle. Still, three of the
+eight are located in the ostrich, crow, and cockoo genomes, again
+within 2 KB of each other on the same scaffolds. This complex pattern
+of loss could be attributed to many different models, e.g. multiple
+losses in birds, poor homology modelling or incomplete genome
+sequences.
 
@@
 retain a single SNORD33 and SNORD34 gene on the same scaffold. While
-the ground-finch and zebrafinch retain a single SNORD33 and the bald
-eagle and seriema retain a single SNORD34 (See Figure~\ref{fig:3}).
-In human these snoRNAs are intronic to the host gene, ribosomal
-protein L13a (RPL13A). Based on BLASTP (version 2.2.18) homology
-searches for the RPL13A gene, the protein is conserved in the human
-and turtle genomes and in the bald eagle, crow, rifleman and
-zebrafinch avian genomes (data not shown). Therefore the conservation
-patterns of the RPL13A host gene and corresponding intronic snoRNAs
-largely mimic each other, each independently supports a pattern of
-loss of the RPL13A gene and the intronic snoRNAs that it hosts in the
-bird genomes.
+the ground-finch and bald eagle retain a single SNORD33 and the
+zebrafinch and seriema retain a single SNORD34 (See
+Figure~\ref{fig:3}).  In human these snoRNAs are intronic to the host
+gene, ribosomal protein L13a (RPL13A). Based on BLASTP (version
+2.2.18) homology searches for the RPL13A gene, the protein is
+conserved in the human and turtle genomes and in the bald eagle, crow,
+rifleman and zebrafinch avian genomes (data not shown). Therefore the
+conservation patterns of the RPL13A host gene and corresponding
+intronic snoRNAs largely mimic each other, each independently supports
+a pattern of loss of the RPL13A gene and the intronic snoRNAs that it
+hosts in the bird genomes.
 
 \subsection*{MicroRNAs}
-
 MicroRNAs are an important class of non-coding RNA. They have been
@@
 
-{\bf MAY BE MOVED INTO THE lncRNA PAPER!}
+%%{\bf MAY BE MOVED INTO THE lncRNA PAPER!}
 
@@
 scaffold, this gene order was preserved.
-
 %egrep 'HOXA11-AS1|HOTAIR|HOTTIP' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
@@
 %MHM results!!!!
-
 %% Another of the functions associated with lncRNAs is chromosomal dosage
@@
 
-\subsection*{Pseudogenes}
+%% \subsection*{Pseudogenes}
 
-The human genome is littered with pseudogenised copies of ncRNAs. The
-Dfam database of human repeats estimates that approximately $54\%$ of
-the human genome is composed of repeat elements
-\cite{Wheeler:2013}. Of these annotations XXX\% are derived from
-ncRNAs, including the large Alu family of repeats, which is derived
-from a retransposed SRP RNA pseudogene \cite{}.
+%% The human genome is littered with pseudogenised copies of ncRNAs. The
+%% Dfam database of human repeats estimates that approximately $54\%$ of
+%% the human genome is composed of repeat elements
+%% \cite{Wheeler:2013}. Of these annotations XXX\% are derived from
+%% ncRNAs, including the large Alu family of repeats, which is derived
+%% from a retransposed SRP RNA pseudogene \cite{}.
 
-tRNA, RSV RNA, 
- \cite{International_Chicken_Genome_Sequencing_Consortium:2004}
+%% tRNA, RSV RNA,
+%%  \cite{International_Chicken_Genome_Sequencing_Consortium:2004}
 
-\emph{Endogenous retroviruses:} RSV RNA, Alfamo CPB
+%% \emph{Endogenous retroviruses:} RSV RNA, Alfamo CPB
 
@@
 
+An additonal snoRNA homology search was performed with snoStrip
+\cite{Bartschat:2013}. As initial queries we used deutorostomian
+snoRNA families from human \cite{Lestrade:2006}, platypus
+\cite{Schmitz:2008}, and chicken \cite{Shao:2009}.
+
+SnoRNA annotations overlapping miRNAs annotations where manually
+inspected \ppg{and?...}.
+
 \section*{Acknowledgements}
@@
 Erich Jarvis, Guojie Zhang and Tom Gilbert. Magnus Alm Rosenblad. 
-
 %%%%%%%%%%%%%%%%%%%%%%
@@
 %% \end{figure}
-
 \begin{figure}[ht]

### 2013-08-27 scripts/gffs2heatmaps.pl

@@ -50,7 +50,7 @@ if (-s "$rDir/allRNA.dat"){
     }
     close(W);
 }
-
+$whitelist{'SNORD34'}=1;
 #replace with a family2type hash:
 my %lncRNAlist;
 if (-s "data/rfam11_lncRNAs.txt"){
@@ -217,6 +217,7 @@ my @tRNAfamily = sort {
 }(keys %tRNAfamilies);
 
 print "Printing R-dat files\n" if (defined($verbose));
+my %exceptions = (SNORD34 => 1);
 foreach my $family ( @family ) {
     
     #print "$family\n";# if ($family=~/Pseudo/i);
@@ -227,7 +228,7 @@ foreach my $family ( @family ) {
     next if(defined($blacklist{$family}));
     open(AL, ">> $rDir/allRNA.dat");
     if ($family=~/^sno/i or $family=~/^SCA/ or $family=~/^ACA/){
-	next if( ($familiesSpeciesCounts{$family}/scalar(@speciesPhyloOrder)) < 0.10 or $familiesTotalCounts{$family}<5);
+	next if( (($familiesSpeciesCounts{$family}/scalar(@speciesPhyloOrder)) < 0.10 or $familiesTotalCounts{$family}<5) and not defined($exceptions{$family}) );
 	open(UT, ">> $rDir/snoRNA.dat");
     }
     elsif ($family=~/^mir/i or $family=~/^let-7/ or $family=~/^lin-4/ ){

### 2013-09-09 paper/supplementary-material.tex
Added further analysis from Peter Stadler, created supp-mat.

--- 2013-08-27 paper/manuscript.tex
@@
 
-\title{Non-coding RNAs of bird genomes}
+\title{Supplementary material for: non-coding RNAs of bird genomes}
 
@@
 \begin{abstract}
-Here we present the results of a large-scale bioinformatic annotation
-of non-coding RNAs in 48 avian genomes. Our approach uses
-probabilistic models of hand-curated families from the Rfam database
-to infer conserved RNA families within each avian genome. We
-supplement these annotations with predictions from the tRNA annotation
-tool, tRNAscan-SE and microRNAs from miRBase. Although we show
-extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and
-more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds,
-we also demonstrate apparent ``losses'' in several RNA families. These
-include the divergence of some classical ncRNAs and the loss of
-several snoRNAs and microRNAs. In contrast, we show a significant
-number of lncRNAs are surprisingly well conserved between birds and
-mammals including several intriguing cases where the reported
-mammalian lncRNA function is not conserved in birds. These combined
-results illustrate the utility of applying homology based methods for
-annotating vertebrate genomes and illustrate many complex evolutionary
-patterns within the avian ncRNA cohort.
+...
 \end{abstract}
@@
 
-\section*{Introduction}
+\section*{Supplementary introduction}
+...
 
-Non-coding RNAs (ncRNAs) are an important class of genes. However,
-they pose serious research challenges, particularly for the field of
-genomics. They lack the strong statistical signals associated with
-protein coding genes, e.g. open reading frames, G+C content and
-codon-usage biases \cite{Rivas:2000}.
+\section*{Supplementary results}
+In the following we explore ...
 
-Non-coding RNAs are classified into hierarchical groupings by the Rfam
-database. The basic units are ``families'' which are groups of
-homologous, alignable sequences; ``clans'' which are groups of
-un-alignable (or functionally distinct), homologous families; and
-``classes'' which are groups of clans and families with related
-biological functions e.g. spliceosomal RNAs, miRNAs and snoRNAs
-\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
-The major RNA families include the classical, highly conserved RNAs,
-sometimes called ``molecular fossils'', such as the transfer RNAs,
-ribosomal RNAs, RNA components of RNase P and the signal recognition
-particle \cite{Jeffares:1998}. Other classes appear to have have
-evolved more recently, e.g. the small nucleolar RNAs (snoRNAs),
-microRNAs (miRNAs) and the long non-coding RNAs (lncRNAs)
-\cite{Hoeppner:2012}. The publication of avian genomes provides
-exciting opportunities to to explore ncRNA conservation in
-unprecedented detail.
-
-Despite promising techniques such as RNA-seq \cite{Croucher:2010},
-homology based methods namely covariance models (CMs) are state of the
-art for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
-The CM based approach for annotating ncRNAs in genomes requires
-reliable alignments and consensus secondary structures of
-representative sequences of RNA families. These are used to train
-probabilistic models for each family. These models can be used to
-generate sequences with similar properties, score the likelihood that
-a sequence is was generated by the same evolutionary processes as the
-training sequences and to build alignments based upon sequence and
-structural information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
-The tRNAscan-SE software package uses CMs to accurately predict
-transfer RNAs \cite{Lowe:1997,Chan:2009}. The Rfam database contains
-thousands of curated alignments and consensus structures for diverse
-classes of ncRNAs
-\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
-Independent benchmarks of bioinformatic annotation tools have shown
-that the CM approaches dramatically out-perform alternative methods
-\cite{Freyhult:2007}.
-
-The CM based approach works well for almost all classes of ncRNA, but
-the long non-coding RNAs (lncRNAs) are a particular challenge
-\cite{Guttman:2009}. Recent technological advances have led to dramatic
-speed and memory-usage enhancements for CM analyses
-\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
-cannot model the exon-intron structures of spliced lncRNAs, nor can they
-deal simply with the repeats that many lncRNAs host. Consequently in
-the latest release of Rfam the lncRNA families that were added were
-composed of local conserved and structured elements within lncRNAs,
-analogous to the ``domains'' housed within protein sequences
-\cite{Burge:2013}. The functions determined to date for lncRNAs
-range from regulating chromatin status to chromosomal inactivation
-\cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
-genes is a lengthy and expensive process \cite{Guttman:2009}.
-
-%Still needs a {\bf what have we done} paragraph!
-
-\section*{Results}
-In the following we explore the conservation patterns of the major
-classes of ncRNA in further detail. The collection of ncRNA sequences
-is generally biased towards model organisms
-\cite{Gardner:2010,Hoeppner:2012}. Consequently the bulk of the
-families discussed below are either conserved across the vertebrates
-or have been apparently lost or diversified in the avian
-lineage(s).
-
-\subsection*{Classic RNAs: LUCA and LECA}
-
-Many RNA families constitute the most evolutionarily conserved genes
-across all life on this planet \cite{Jeffares:1998}. Examples of RNAs
-derived from the Last Universal Common Ancestor (LUCA) include, the
-transfer RNAs (tRNA), ribosomal RNAs (rRNA), RNA components of RNase P
-(RNase P RNA), RNase MRP (RNase MRP RNA) and the signal recognition
-particle (SRP RNA). Other classes of RNA are likely to have been
-components of the Last Eukaryotic Common Ancestor (LECA) include the
-telomerase RNA, major spliceosomal RNAs (U1, U2, U4, U5, and U6) and
-the minor spliceosomal RNAs (U11, U12, U4atac, and U6atac)
-\cite{Hoeppner:2012}.
-
-%\subsubsection*{Losses}
-
-Unsurprisingly, the bulk of these classes of RNAs are well represented
-across the bird genomes (See Figure~\ref{fig:1}). However, there
-appear to have been ``losses'' of a few of these RNAs in certain bird
-species. Some of these may be due to sequence divergence, of which
-there are several notable examples e.g.
-\cite{Leonardi:2008,Webb:2008,Mao:,Lai:2010,Chan:2011}. Other losses
-may be due to incomplete genome coverage.
-
-A number of the classic RNAs are incorporated into RNA-protein
-complexes (RNPs) involved in core cellular processes. An example of
-this are the spliceosomal RNAs. Based upon the presence/absence
-patterns of the major spliceosomal RNAs they are all well represented
-in these genome sequences. Except for the U4 RNA in cormorant and the
-U5 RNA in the bee eater. Again these two genomes are low coverage
-suggesting these genes weren't captured in the current assembly. The
-minor spliceosomal RNAs are more interesting, the U4atac and U11
-snRNAs show widespread patterns of loss, even in some of the high
-coverage genomes. These RNAs are frequently missed in bioinformatic
-screens. Indicating either frequent loss \cite{Davila_Lopez:2008} or
-sequences that have diverged beyond the ability of detection by
-covariance models \cite{Marz:2008}.
-
-The telomerase RNA is also largely missing from the avian
-annotations. This RNA acts as a template for the telomerase enzyme
-that extends the telomeres found on chromosome ends. It is only found
-in the chicken, bald eagle, kea, budgerigar, crow and
-zebrafinch. Homology searches searches with the telomerase reverse
-transcriptase (TERT) protein show that the protein component of the
-telomerase RNP is conserved across all the bird genomes (data not
-shown). This pattern of presumably divergent telomerase RNA and
-conserved telomerase protein has been noted previously, most notably
-in the fungi \cite{Leonardi:2008,Webb:2008}.
-
-The the RNA components of RNase P and RNase MRP also appear to have
-undergone dramatic losses within the bird lineage. RNase P is required
-for the maturation of tRNA, the paralogous enzyme, RNase MRP is
-required for the maturation of rRNA. Each RNP cleaves smaller RNAs
-from larger transcripts \cite{Lopez:2009}. It is unlikely that the
-these genes have been lost in any of the birds. Homology searches with
-the RNase associated protein coding genes (POP1, POP4, POP5, POP7,
-RPP1, RPP14, RPP25, RPP38, RPP40 and RPR2), identified viable homologs
-of each in all of the bird genomes \cite{Rosenblad:2006} (data not
-shown). This suggests that the bird RNase P and MRP RNAs may have
-diverged slightly from the canonical models.
-
-The 5.8 S component of the ribosome in the turtle, turkey bustard,
-hoatzin, flamingo, tropicbird, seriema, owl, cuckoo roller, trogon,
-bee eater and falcon appears to have been lost (See
-Figure~\ref{fig:1}). The rRNA repeats are frequently not assembled,
-consequently it is not surprising to see ``losses'' in these \cite{Floutsakou:2013}.
-Furthermore, the genomes for these species are also
-low-coverage.
-
-\subsection*{Small nucleolar RNAs}
-
-Small nucleolar RNAs (snoRNAs) are important ncRNAs that participate
-in the maturation of other functional RNAs \cite{Gardner:2010}. The
-bulk of the characterised snoRNAs guide either methylation or
-pseudouridylation modifications, primarily of rRNAs but also
-spliceosomal RNAs. The two types of modifications are guided by two
-different types of RNA, the box C/D and the H/ACA snoRNAs
-respectively, each with a characteristic cohort of motifs and
-secondary structures \cite{Marz:2011}.
-
-There are 66 ribosomal modification sites, guided by 59 snoRNA
-families, that are preserved between \emph{H. sapiens} and
-\emph{S. cerevisiae} \cite{Lestrade:2006}. Of these, 45 snoRNA
-families are conserved in the bird dataset.
-Over a third of the apparent losses of the yeast-human conserved
-snoRNA families appear to cluster on 2 locii of the ancestral
-vertebrate genome. We investigated these losses further.
-
-The first cluster is found at chr11:62620797-62622484 on the human
-genome (hg19) and contains SNORD27, SNORD29 and SNORD31 of the
-human-yeast conserved snoRNAs. These snoRNAs are located in the
-inside-out gene SNHG1 which hosts a total of eight C/D box snoRNAs:
-SNORD25, SNORD26, SNORD27, SNORD28, SNORD29, SNORD22, SNORD30 and
-SNORD31 \cite{Tycowski:1996}. Each of which are also found in the
-alligator and turtle genomes within a 3-4 KB locus, yet these have
-largely been lost in the birds. Notable exceptions are five of the
-eight are located in the tinamou genome, these are located on the same
-scaffold and are within 2 KB of each other. Implying that SNHG1 is
-conserved in the tinamou. Loci with four of the eight snoRNAs can be
-found in zebrafinch, ground-finch, and bald eagle. Still, three of the
-eight are located in the ostrich, crow, and cockoo genomes, again
-within 2 KB of each other on the same scaffolds. This complex pattern
-of loss could be attributed to many different models, e.g. multiple
-losses in birds, poor homology modelling or incomplete genome
-sequences.
-
-%Furthermore our results contradict the text in the paper. We find
-%SNORD26 stated to be lost in sauropsids in the 5 species:
-%C.vociferus, M.unicolor, H.leucocephalus, L.discolor and T.guttata.
-
-%%GAS5?
-%SNORD81, SNORD47, SNORD80, SNORD79, SNORD78, SNORD44, SNORD77, SNORD76, SNORD75, SNORD74
-%egrep 'SNORD81;|SNORD47;|SNORD80;|SNORD79;|SNORD78;|SNORD44;|SNORD77;|SNORD76;|SNORD75;|SNORD74' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
-
-The second cluster is located at chr19:49993222-49994231 on the human
-genome (hg19) and contains two copies of SNORD33 and one SNORD34 all
-within a 1 KB genomic region. The turtle and alligator genomes retain
-the two copies of SNORD33 yet don't have an obvious SNORD34 gene on
-the same scaffold. Within the bird genomes, the crow and rifleman each
-retain a single SNORD33 and SNORD34 gene on the same scaffold. While
-the ground-finch and bald eagle retain a single SNORD33 and the
-zebrafinch and seriema retain a single SNORD34 (See
-Figure~\ref{fig:3}).  In human these snoRNAs are intronic to the host
-gene, ribosomal protein L13a (RPL13A). Based on BLASTP (version
-2.2.18) homology searches for the RPL13A gene, the protein is
-conserved in the human and turtle genomes and in the bald eagle, crow,
-rifleman and zebrafinch avian genomes (data not shown). Therefore the
-conservation patterns of the RPL13A host gene and corresponding
-intronic snoRNAs largely mimic each other, each independently supports
-a pattern of loss of the RPL13A gene and the intronic snoRNAs that it
-hosts in the bird genomes.
-
-\subsection*{MicroRNAs}
-MicroRNAs are an important class of non-coding RNA. They have been
-found in the genomes of Chromalveolata \cite{Cock:2010,Huang:2011},
-Metazoa \cite{Lee:1993,Lau:2001,Hertel:2006}, Mycetozoa
-\cite{Hinas:2007,Avesson:2012}, Viridiplantae
-\cite{Reinhart:2002,Fattash:2007,Axtell:2007,Molnar:2007} and Viruses
-\cite{Pfeffer:2004,Ouellet:2008,Pfeffer:2005,Landgraf:2007}. The
-miRNAs have been shown to regulate the expression of large numbers of
-messenger RNAs \cite{Lim:2005}. The mature miRNA product is generally
-22 nucleotides long which is usually processed from a larger RNA that
-is characterised by a stable hairpin-shaped secondary structure.
-
-Chicken and zebrafinch are the only birds with previously annotated
-microRNAs. We searched for homologs of these and other vertebrate
-microRNAs in the genomes of the 48 birds, American alligator and green
-turtle. Overall, we annotate a total of 16617 putative microRNA loci,
-homologous to 543 known microRNA genes, of which 487 are annotated in
-chicken and/or zebra finch, while 56 have been so far known only in
-non-avian vertebrates. The numbers of annotated loci in the individual
-species are approximately equal - 300-400 per species, except for the
-turkey (Meleagris gallopavo) where we identified 543 sequences
-homologous to known microRNAs.
-
-Nearly all the microRNAs that are broadly conserved in fish,
-amphibians and mammals are also conserved in the birds. Nevertheless,
-there are obvious instances of microRNAs lost in all birds. For
-example, mammalian, amphibian and fish genomes contain three loci of
-clustered microRNAs from the mir-17 and mir-92 families. One of these
-clusters (mir-93~25) was not found in turtles, crocodiles and
-birds. In addition, we can confidently identify a further 3 microRNA
-families that are present in mammals, and turtle and/or crocodile, but
-not in any avian genome (mir-150, mir-208, mir-590). This suggests
-that these sequences were lost in the last common ancestor of
-archosaurs or birds. There are also a number of microRNAs that are
-predicted to be present in turtles and/or crocodiles, and only a small
-number of bird genomes. Indeed, there are many missing annotations,
-species-specific and otherwise, that are not consistent with the
-consensus phylogeny, and could be due to either incomplete genomes or
-widespread microRNA loss.
-
-The turkey genome contains a high number (190) of microRNAs so far
-found only in chicken, which account for the higher number of
-annotated sequences in this genome compared with other birds. This is
-consistent with its phylogenetic position as the closest chicken
-relative among the examined birds. However, 101 chicken microRNAs
-have no homolog in the turkey or other bird genomes, suggesting that
-these genes are chicken-specific. This is consistent with previous
-reports of large number of speciesspecific microRNAs in all animals,
-and supports the view of fast microRNA turnover during animal
-evolution.
-
-\subsection*{Long non-coding RNAs}
-
-%%{\bf MAY BE MOVED INTO THE lncRNA PAPER!}
-
-As mentioned in the introduction, lncRNAs are a diverse group of RNAs
-that have been implicated in a multitude of functional processes
-\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have mainly been
-characterised in mammalian species, particularly human and mouse.
-
-In general Rfam cannot include the entire length of any large, spliced
-RNAs. This is a limitation of the covariance-models used for the
-homology-searches Rfam runs \cite{Nawrocki:2009}. Consequently, only
-short, well-conserved regions with evolutionarily conserved secondary
-structures are included in Rfam, we will refer to these as RNA-domains
-\cite{Burge:2013}.
-
-When analysing the RNA-domain annotations it is striking that many of
-the lncRNAs with multiple RNA-domains appear to be consistently
-preserved in the birds. The annotations of these domains are all in
-the same genomic region supporting a high degree of evolutionary
-conservation for the entire lncRNA. In particular the HOXA11-AS1,
-PCA3, RMST, Six3os1, SOX2OT and ST7-OT3 lncRNAs have multiple well
-conserved RNA-domains (See Figure~\ref{fig:5}).
-
-The HOX cluster lncRNAs, HOTAIRM1, HOTTIP and HOXA11-AS1, are
-remarkably well conserved. In the human genome the corresponding
-RNA-domains are located at chr7:27135743-27245922 (hg19) and are
-modelled by 15 RNA-domains in total, the % (5, 4 and 6 respectively)
-lncRNAs are ordered HOTAIRM1 (5 RNA-domains) $\rightarrow$ HOXA11-AS1
-(6 RNA-domains) $\rightarrow$ HOTTIP (4 RNA-domains). In the alligator
-and turtle genomes five and six of the HOX RNA-domains are conserved,
-each supporting an ancestral gene order of HOTAIRM1 $\rightarrow$
-HOXA11-AS1 $\rightarrow$ HOTTIP. In the birds, where two or more of
-the HOX cluster lncRNA RNA-domains were predicted on the same
-scaffold, this gene order was preserved.
-%egrep 'HOXA11-AS1|HOTAIR|HOTTIP' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
-%egrep 'PCA3' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
-%egrep 'RMST' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
-
-Many of the lncRNAs have been associated with cancer, sparking a minor
-review industry \cite{Prensner:2011,Spizzo:2012}. Three examples of
-these that are also conserved in the birds are described below.
-
-The RMST (Rhabdomyosarcoma 2 associated transcript) RNA-domains 6, 7,
-8, and 9 are conserved across the birds. In each bird the gene order
-was also consistent with the human ordering. In the alligator and
-turtle an additional RNA-domain was predicted in each, these were
-RNA-domains 2 and 4 respectively, again the ordering of the domains
-was consistent with human. This suggests that the RMAST lncRNA is
-highly conserved. However, little is known about the function of this
-RNA. It was originally identified in a screen for differentially
-expressed genes in two Rhabdomyosarcoma tumor types \cite{Chan:2002}.
-
-In addition, the lncRNA DLEU2, is well conserved conserved across the
-vertebrates, it is a host gene for two miRNA genes, miR-15 and miR-16,
-both of which are also well conserved across the vertebrates (See
-Figure~\ref{fig:3}). DLEU2 is thought to be a tumor-suppressor
-gene as it is frequently deleted in malignant tumours \cite{Lerner:2009,Klein:2010}.
-
-%(MIR1251 is in the same locus)
-%PCA3 RNA-domains inverted relative to human...
-
-The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
-\cite{Xu:1997}. Both are expressed in a broad ranch of tissues. Extensive
-research on BRCA1 has shown that it is involved in DNA repair
-\cite{}. The function of NBR2 remains unknown, yet its conservation
-across the vertebrates certainly implies a function (See
-Figure~\ref{fig:5}).
-
-%(neighbor of BRCA1 gene 2)
-%MHM results!!!!
-%% Another of the functions associated with lncRNAs is chromosomal dosage
-%% compensation. In mammals this is carried out by the lncRNAs Xist and
-%% Tsix \cite{}. In birds it has been proposed that the lncRNA MHM might
-%% play a related role \cite{}.
-
-%DLEU2, frequently deleted in malignancy, functions as a critical host
-%gene of the cell cycle inhibitory microRNAs miR-15a and miR-16-1
-
-%% TUMOR-SUPPRESSOR:
-%% DLEU2, part of a tumor-suppressor cluster (with two microRNAs), most recent refs here:
-%% http://www.ncbi.nlm.nih.gov/pubmed/19591824
-%% http://www.ncbi.nlm.nih.gov/pubmed/20060366
-%% SMAD5, tumor suppressor candidate:
-%% http://www.ncbi.nlm.nih.gov/pubmed/10049768?dopt=Abstract
-
-%% DEVELOPMENT:
-%% HOXA11, most complete reference list here:
-%% http://www.lncrnadb.org/Detail.aspx?TKeyID=171
-%% Six3OS, retinal development:
-%% http://www.ncbi.nlm.nih.gov/pubmed/21936910
-%% ZEB2, brain development:
-%% http://www.ncbi.nlm.nih.gov/pubmed/18347095?dopt=Abstract
-
-%% MAT2A:
-%% http://www.ncbi.nlm.nih.gov/pubmed/1426236%2C9703951?dopt=Abstract
-
-%% RMST, doesn't look we know it does (also shows up in review paper at top):
-%% http://www.ncbi.nlm.nih.gov/pubmed/20062813
-
-%% ST7-OT, another candidate:
-%% http://www.ncbi.nlm.nih.gov/pubmed/12213198?dopt=Abstract
-
-%% PART1, meant to be human prostate-specific, birds don't even have a prostate:
-%% http://www.ncbi.nlm.nih.gov/pubmed/10706094?dopt=Abstract
-%% PCA3, prostate/androgen receptor signalling (curious there's heaps of it in male and female birds? same for PART1?):
-%% http://www.ncbi.nlm.nih.gov/pubmed/?term=ferreira+and+pca3
-
-\subsection*{Cis-regulatory elements}
-%% Histone3, IRE, IRES, K chan RES, SECIS, Antizyme FSE, Vimentin3, GABA3,
-%% CAESAR,
-
-The cis-regulatory RNAs are a group of RNA structures encoded on
-mRNAs. Generally they are involved in regulating the expression of the
-mRNA they are encoded within. Others may recode the translated protein
-product into an alternate sequence.
-
-This group includes the iron response element (IRE) \cite{Stevens:}
-and the histone $3^\prime$ UTR (histone3)
-\cite{Davila_Lopez:2008a}. These are structured motifs bound by
-regulatory proteins. The selenocysteine insertion sequence (SECIS) is
-a structured motif that recodes UGA stop codons to selenocysteines
-\cite{Lambert:2002} and the GABRA3 stem-loop is a structure recognised
-by the ADAR enzyme family. This enzyme edits adenine nucleotides to
-inosine, in this case recoding an isoleucine codon to methionine in
-exon 9 of the GABRA3 gene \cite{Ohlson:2007}.
-
-These regulatory elements and others, including an internal ribosome
-entry site (IRES), potassium channel RNA editing signal (K chan RES),
-Antizyme RNA frameshifting stimulation element (Antizyme FSE),
-vimentin $3^\prime$ UTR protein-binding region (Vimentin3) and a
-connective tissue growth factor (CTGF) $3^\prime$ UTR element (CAESAR)
-are conserved across a diverse group of vertebrates, including the
-bird lineages explored here (See Figure~\ref{fig:1}).
-
-\subsection*{Exceptional RNAs}
-
-%Background to this group...
-
-A number of other ncRNAs can also be found in Eukaryotic
-genomes. These do not fit into the main classes of RNA but still
-perform vital roles in the function and evolution of Eukaryotes.
-Their functions are diverse and many have not yet been characterised.
-
-An example of an uncharacterised RNA is the ultraconserved element,
-uc.338 (also known as TUC338)
-\cite{Bejerano:2004,Bejerano:2006,Braconi:2011}. The uc.338 element
-is derived from a short interspersed element (SINE) called the
-lobe-finned fishes SINE (LF-SINE) as it is conserved between the
-coelacanth and mammals \cite{Bejerano:2006}. Analysis of the
-expression of uc.338 implies that it plays a role in the progression
-of hepatocellular carcinoma, possibly by influencing cell growth
-\cite{Braconi:2011}. This RNA is conserved in the birds and appears to
-have been duplicated in several lineages.
-
-The Y RNA is an enigmatic ncRNA where we know very little about the
-function. It was discovered in the 1980s in ribonucleoprotein
-complexes \cite{Lerner:1981}. The function of the Y RNAs remain
-unknown, but evidence is emerging that they may be associated with DNA
-replication \cite{Christov:2006}. There are 4 functional Y RNAs
-encoded in the human genome, Y1, Y3, Y4 and Y5. However, there are
-hundreds of pseudogenised copies of the Y RNA scattered throughout the
-human genome \cite{Mosig:2007}. In the birds and other lizards, we
-identify between two and seven Y RNA paralogs (See
-Figure~\ref{fig:1}).
-
-The Vault RNA forms a major component of the vault ribonucleoprotein
-complex, this is one of the largest particles found in the vertebrate
-cell; In fact, it is larger than the ribosome \cite{Kong:1999}. As yet
-not much is known about the function of Vault. The Vault RNA has been
-shown to be broadly conserved in metazoans
-\cite{Stadler:2009}. However, in the bird lineages it appears to have
-either been lost or diversified.
-
-% Vault, 7SK, HAR1A
-
-%% \subsection*{Pseudogenes}
-
-%% The human genome is littered with pseudogenised copies of ncRNAs. The
-%% Dfam database of human repeats estimates that approximately $54\%$ of
-%% the human genome is composed of repeat elements
-%% \cite{Wheeler:2013}. Of these annotations XXX\% are derived from
-%% ncRNAs, including the large Alu family of repeats, which is derived
-%% from a retransposed SRP RNA pseudogene \cite{}.
-
-%% tRNA, RSV RNA,
-%%  \cite{International_Chicken_Genome_Sequencing_Consortium:2004}
-
-%% \emph{Endogenous retroviruses:} RSV RNA, Alfamo CPB
-
-\subsection*{Contamination}
-
-Bacterial families can be used to identify problematic sequences that
-are likely to be the result of contamination from non-avian sources.
-
-\section*{Conclusions}
-
-%Talking points:
-%1. several classical RNAs (e.g. spliceosomal, telomerase, RNase MRP
-%RNAs) have diversified in birds relative to other Eukaryotes.
-%2. several lncRNA families are conserved between mammals and
-%birds. Several of these are characterised by expression in tissues
-%that are not present in birds (e.g. the prostate).
-%3. several regulatory RNAs, snoRNAs and microRNAs, have been lost
-%since the birds diverged from the other vertebrates. Indicating many
-%avian-specific alterations in gene regulation.
-
-\section*{Methods}
-
-Bird genomes were searched using the cmsearch program from INFERNAL
-1.1 and the covariance models from the Rfam database
-v11.0\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}. All
-matches above the curated GA threshold were included. Subsequently,
-all hits with an E value greater than 0.0005 were discarded, so only
-matches which passed the model-specific GA threshold and had an E-value smaller than 0.0005 were retained.
-
-tRNA-scan version 1.3.1, ... \cite{Lowe:1997,Chan:2009}
-
-Rfam matches and the tRNA-scan results for families belonging to the
-same clan were then competed so that only the best match was retained.
-Matches were further sorted into taxonomic groups {\bf ONLY MATCHES IN 5 OR MORE ...}
-
-999 microRNA sequence families, previously annotated in at least one
-vertebrate, were retrieved from miRBase (v19). Individual sequences or
-multiple sequence alignments were used to build covariance models with
-INFERNAL (v1.1rc3), and these models were searched against the 48 bird
-genomes, and the genomes of the american alligator and the green
-turtle as outgroups. Hits with e-value <10 realigned with the query
-sequences and the resultant multiple sequence alignments manually
-inspected and edited using RALEE.
-
-An additonal snoRNA homology search was performed with snoStrip
-\cite{Bartschat:2013}. As initial queries we used deutorostomian
-snoRNA families from human \cite{Lestrade:2006}, platypus
-\cite{Schmitz:2008}, and chicken \cite{Shao:2009}.
-
-SnoRNA annotations overlapping miRNAs annotations where manually
-inspected \ppg{and?...}.
-
-\section*{Acknowledgements}
-
-Erich Jarvis, Guojie Zhang and Tom Gilbert. Magnus Alm Rosenblad. 
-%%%%%%%%%%%%%%%%%%%%%%
-%% The Bibliography %%
-%%
-%% ---------------------------------
-%% BioMedCentral bibtex .BST file will be used to
-%% create a .BBL file which includes the BMC XML.
-%% Note that the displayed Bibliography will not be
-%% exactly as specified in the online Instructions for Authors
+\section*{Supplementary methods}
+...
 
@@
 
-\begin{figure}[ht]
-  \includegraphics[width=0.4\textwidth]{figures/RNA.pdf}
-  \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
-  \includegraphics[width=0.4\textwidth]{figures/snoRNA-human-yeast-correspondences.pdf}
-  \caption[]{This figure illustrates the number of copies of Rfam and
-    tRNAscan annotations for each RNA family in each avian species and
-    outgroups. The families are ordered alphabetically vertically and
-    the species are in phylogenetic order horizontally. Darker shades
-    indicate high copy numbers, lighter shades indicate fewer, white
-    indicates no predictions were made.}\label{fig:1}
-\end{figure}
-
-%% \begin{figure}[ht]
-%% \includegraphics[width=0.65\textwidth]{figures/snoRNA-human-yeast-correspondences.pdf}
-%% \caption[]{.}\label{fig:2}
-%% \end{figure}
 \begin{figure}[ht]
@@
 
+\begin{figure}[ht]
+  \includegraphics[width=0.45\textwidth]{figures/Y.pdf}
+  \caption[]{...}\label{fig:6}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics[width=0.45\textwidth]{figures/More-cand.pdf}
+  \caption[]{...}\label{fig:7}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics[width=0.45\textwidth]{figures/mir-17-cluster.pdf}
+  \caption[]{...}\label{fig:8}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics[width=0.45\textwidth]{figures/let-7-cluster.pdf}
+  \caption[]{...}\label{fig:9}
+\end{figure}
+
 \end{bmcformat}

### 2013-09-23 paper/manuscript.tex
updated manuscript.

@@
 
-\title{Non-coding RNAs of bird genomes}
+%\title{Non-coding RNAs of bird genomes}
+%\title{Wins and losses: Non-coding RNAs of bird genomes}
+\title{Fly away Peter, come back Paul: the gains and losses of avian Non-coding RNAs}
 
@@
 is generally biased towards model organisms
-\cite{Gardner:2010,Hoeppner:2012}. Consequently the bulk of the
-families discussed below are either conserved across the vertebrates
-or have been apparently lost or diversified in the avian
-lineage(s).
+\cite{Gardner:2010,Hoeppner:2012}. We focus our analysis on the
+unusual results within the avian lineages. These are either
+unexpectedly well-conserved RNAs or unexpectedly poorly-conserved
+RNAs. The former are RNAs we would not have expected to be conserved
+between the birds and the organisms these genes were initially
+identified in; Usually, this hypothesis is based upon the function of
+the RNA which is not conserved in avian species. The latter are
+apparent losses of RNA genes that were expected to be conserved;
+Usually, this hypothesis is based upon the conservation of these RNAs
+in other vertebrate species. We have three models of ``loss'' that we
+use to explain the data: Firstly, these could be genuine cases of an
+ancestral gene-loss along the avian lineage. Secondly, this could be a
+case of ``divergence'' where a RNA gene has undergone significant
+sequence and structural alterations, so much so, that our homology
+search tools can no longer detect a relationship between vertebrate
+exemplars and the avian varieties. Thirdly, we consider the
+possibility that the available genome assemblies have independently
+failed to capture these genes.
 
-\subsection*{Classic RNAs: LUCA and LECA}
+\subsection*{Unusually, well conserved RNAs}
 
-Many RNA families constitute the most evolutionarily conserved genes
-across all life on this planet \cite{Jeffares:1998}. Examples of RNAs
-derived from the Last Universal Common Ancestor (LUCA) include, the
-transfer RNAs (tRNA), ribosomal RNAs (rRNA), RNA components of RNase P
-(RNase P RNA), RNase MRP (RNase MRP RNA) and the signal recognition
-particle (SRP RNA). Other classes of RNA are likely to have been
-components of the Last Eukaryotic Common Ancestor (LECA) include the
-telomerase RNA, major spliceosomal RNAs (U1, U2, U4, U5, and U6) and
-the minor spliceosomal RNAs (U11, U12, U4atac, and U6atac)
-\cite{Hoeppner:2012}.
-
-%\subsubsection*{Losses}
-
-Unsurprisingly, the bulk of these classes of RNAs are well represented
-across the bird genomes (See Figure~\ref{fig:1}). However, there
-appear to have been ``losses'' of a few of these RNAs in certain bird
-species. Some of these may be due to sequence divergence, of which
-there are several notable examples e.g.
-\cite{Leonardi:2008,Webb:2008,Mao:,Lai:2010,Chan:2011}. Other losses
-may be due to incomplete genome coverage.
-
-A number of the classic RNAs are incorporated into RNA-protein
-complexes (RNPs) involved in core cellular processes. An example of
-this are the spliceosomal RNAs. Based upon the presence/absence
-patterns of the major spliceosomal RNAs they are all well represented
-in these genome sequences. Except for the U4 RNA in cormorant and the
-U5 RNA in the bee eater. Again these two genomes are low coverage
-suggesting these genes weren't captured in the current assembly. The
-minor spliceosomal RNAs are more interesting, the U4atac and U11
-snRNAs show widespread patterns of loss, even in some of the high
-coverage genomes. These RNAs are frequently missed in bioinformatic
-screens. Indicating either frequent loss \cite{Davila_Lopez:2008} or
-sequences that have diverged beyond the ability of detection by
-covariance models \cite{Marz:2008}.
-
-The telomerase RNA is also largely missing from the avian
-annotations. This RNA acts as a template for the telomerase enzyme
-that extends the telomeres found on chromosome ends. It is only found
-in the chicken, bald eagle, kea, budgerigar, crow and
-zebrafinch. Homology searches searches with the telomerase reverse
-transcriptase (TERT) protein show that the protein component of the
-telomerase RNP is conserved across all the bird genomes (data not
-shown). This pattern of presumably divergent telomerase RNA and
-conserved telomerase protein has been noted previously, most notably
-in the fungi \cite{Leonardi:2008,Webb:2008}.
-
-The the RNA components of RNase P and RNase MRP also appear to have
-undergone dramatic losses within the bird lineage. RNase P is required
-for the maturation of tRNA, the paralogous enzyme, RNase MRP is
-required for the maturation of rRNA. Each RNP cleaves smaller RNAs
-from larger transcripts \cite{Lopez:2009}. It is unlikely that the
-these genes have been lost in any of the birds. Homology searches with
-the RNase associated protein coding genes (POP1, POP4, POP5, POP7,
-RPP1, RPP14, RPP25, RPP38, RPP40 and RPR2), identified viable homologs
-of each in all of the bird genomes \cite{Rosenblad:2006} (data not
-shown). This suggests that the bird RNase P and MRP RNAs may have
-diverged slightly from the canonical models.
-
-The 5.8 S component of the ribosome in the turtle, turkey bustard,
-hoatzin, flamingo, tropicbird, seriema, owl, cuckoo roller, trogon,
-bee eater and falcon appears to have been lost (See
-Figure~\ref{fig:1}). The rRNA repeats are frequently not assembled,
-consequently it is not surprising to see ``losses'' in these \cite{Floutsakou:2013}.
-Furthermore, the genomes for these species are also
-low-coverage.
-
-\subsection*{Small nucleolar RNAs}
-
-Small nucleolar RNAs (snoRNAs) are important ncRNAs that participate
-in the maturation of other functional RNAs \cite{Gardner:2010}. The
-bulk of the characterised snoRNAs guide either methylation or
-pseudouridylation modifications, primarily of rRNAs but also
-spliceosomal RNAs. The two types of modifications are guided by two
-different types of RNA, the box C/D and the H/ACA snoRNAs
-respectively, each with a characteristic cohort of motifs and
-secondary structures \cite{Marz:2011}.
-
-There are 66 ribosomal modification sites, guided by 59 snoRNA
-families, that are preserved between \emph{H. sapiens} and
-\emph{S. cerevisiae} \cite{Lestrade:2006}. Of these, 45 snoRNA
-families are conserved in the bird dataset.
-Over a third of the apparent losses of the yeast-human conserved
-snoRNA families appear to cluster on 2 locii of the ancestral
-vertebrate genome. We investigated these losses further.
-
-The first cluster is found at chr11:62620797-62622484 on the human
-genome (hg19) and contains SNORD27, SNORD29 and SNORD31 of the
-human-yeast conserved snoRNAs. These snoRNAs are located in the
-inside-out gene SNHG1 which hosts a total of eight C/D box snoRNAs:
-SNORD25, SNORD26, SNORD27, SNORD28, SNORD29, SNORD22, SNORD30 and
-SNORD31 \cite{Tycowski:1996}. Each of which are also found in the
-alligator and turtle genomes within a 3-4 KB locus, yet these have
-largely been lost in the birds. Notable exceptions are five of the
-eight are located in the tinamou genome, these are located on the same
-scaffold and are within 2 KB of each other. Implying that SNHG1 is
-conserved in the tinamou. Loci with four of the eight snoRNAs can be
-found in zebrafinch, ground-finch, and bald eagle. Still, three of the
-eight are located in the ostrich, crow, and cockoo genomes, again
-within 2 KB of each other on the same scaffolds. This complex pattern
-of loss could be attributed to many different models, e.g. multiple
-losses in birds, poor homology modelling or incomplete genome
-sequences.
-
-%Furthermore our results contradict the text in the paper. We find
-%SNORD26 stated to be lost in sauropsids in the 5 species:
-%C.vociferus, M.unicolor, H.leucocephalus, L.discolor and T.guttata.
-
-%%GAS5?
-%SNORD81, SNORD47, SNORD80, SNORD79, SNORD78, SNORD44, SNORD77, SNORD76, SNORD75, SNORD74
-%egrep 'SNORD81;|SNORD47;|SNORD80;|SNORD79;|SNORD78;|SNORD44;|SNORD77;|SNORD76;|SNORD75;|SNORD74' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
-
-The second cluster is located at chr19:49993222-49994231 on the human
-genome (hg19) and contains two copies of SNORD33 and one SNORD34 all
-within a 1 KB genomic region. The turtle and alligator genomes retain
-the two copies of SNORD33 yet don't have an obvious SNORD34 gene on
-the same scaffold. Within the bird genomes, the crow and rifleman each
-retain a single SNORD33 and SNORD34 gene on the same scaffold. While
-the ground-finch and bald eagle retain a single SNORD33 and the
-zebrafinch and seriema retain a single SNORD34 (See
-Figure~\ref{fig:3}).  In human these snoRNAs are intronic to the host
-gene, ribosomal protein L13a (RPL13A). Based on BLASTP (version
-2.2.18) homology searches for the RPL13A gene, the protein is
-conserved in the human and turtle genomes and in the bald eagle, crow,
-rifleman and zebrafinch avian genomes (data not shown). Therefore the
-conservation patterns of the RPL13A host gene and corresponding
-intronic snoRNAs largely mimic each other, each independently supports
-a pattern of loss of the RPL13A gene and the intronic snoRNAs that it
-hosts in the bird genomes.
-
-\subsection*{MicroRNAs}
-MicroRNAs are an important class of non-coding RNA. They have been
-found in the genomes of Chromalveolata \cite{Cock:2010,Huang:2011},
-Metazoa \cite{Lee:1993,Lau:2001,Hertel:2006}, Mycetozoa
-\cite{Hinas:2007,Avesson:2012}, Viridiplantae
-\cite{Reinhart:2002,Fattash:2007,Axtell:2007,Molnar:2007} and Viruses
-\cite{Pfeffer:2004,Ouellet:2008,Pfeffer:2005,Landgraf:2007}. The
-miRNAs have been shown to regulate the expression of large numbers of
-messenger RNAs \cite{Lim:2005}. The mature miRNA product is generally
-22 nucleotides long which is usually processed from a larger RNA that
-is characterised by a stable hairpin-shaped secondary structure.
-
-Chicken and zebrafinch are the only birds with previously annotated
-microRNAs. We searched for homologs of these and other vertebrate
-microRNAs in the genomes of the 48 birds, American alligator and green
-turtle. Overall, we annotate a total of 16617 putative microRNA loci,
-homologous to 543 known microRNA genes, of which 487 are annotated in
-chicken and/or zebra finch, while 56 have been so far known only in
-non-avian vertebrates. The numbers of annotated loci in the individual
-species are approximately equal - 300-400 per species, except for the
-turkey (Meleagris gallopavo) where we identified 543 sequences
-homologous to known microRNAs.
-
-Nearly all the microRNAs that are broadly conserved in fish,
-amphibians and mammals are also conserved in the birds. Nevertheless,
-there are obvious instances of microRNAs lost in all birds. For
-example, mammalian, amphibian and fish genomes contain three loci of
-clustered microRNAs from the mir-17 and mir-92 families. One of these
-clusters (mir-93~25) was not found in turtles, crocodiles and
-birds. In addition, we can confidently identify a further 3 microRNA
-families that are present in mammals, and turtle and/or crocodile, but
-not in any avian genome (mir-150, mir-208, mir-590). This suggests
-that these sequences were lost in the last common ancestor of
-archosaurs or birds. There are also a number of microRNAs that are
-predicted to be present in turtles and/or crocodiles, and only a small
-number of bird genomes. Indeed, there are many missing annotations,
-species-specific and otherwise, that are not consistent with the
-consensus phylogeny, and could be due to either incomplete genomes or
-widespread microRNA loss.
-
-The turkey genome contains a high number (190) of microRNAs so far
-found only in chicken, which account for the higher number of
-annotated sequences in this genome compared with other birds. This is
-consistent with its phylogenetic position as the closest chicken
-relative among the examined birds. However, 101 chicken microRNAs
-have no homolog in the turkey or other bird genomes, suggesting that
-these genes are chicken-specific. This is consistent with previous
-reports of large number of speciesspecific microRNAs in all animals,
-and supports the view of fast microRNA turnover during animal
-evolution.
-
-\subsection*{Long non-coding RNAs}
-
-%%{\bf MAY BE MOVED INTO THE lncRNA PAPER!}
-
-As mentioned in the introduction, lncRNAs are a diverse group of RNAs
-that have been implicated in a multitude of functional processes
-\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have mainly been
-characterised in mammalian species, particularly human and mouse.
+The bulk of the ``unusually well conserved RNAs'' belong to the long
+non-coding RNA (lncRNA) group.  The lncRNAs are a diverse group of
+RNAs that have been implicated in a multitude of functional processes
+\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have largely been
+characterised in mammalian species, particularly human and
+mouse. Consequently, generally do not expect these to be conserved
+outside of mammalia. Notable examples include Xist \cite{Duret:2006}
+and H19 \cite{Smits:2008}.
 
@@
 short, well-conserved regions with evolutionarily conserved secondary
-structures are included in Rfam, we will refer to these as RNA-domains
-\cite{Burge:2013}.
+structures are included in Rfam. By analogy to protein-domains, we
+refer to these as RNA-domains \cite{Burge:2013}.
 
 When analysing the RNA-domain annotations it is striking that many of
-the lncRNAs with multiple RNA-domains appear to be consistently
-preserved in the birds. The annotations of these domains are all in
-the same genomic region supporting a high degree of evolutionary
-conservation for the entire lncRNA. In particular the HOXA11-AS1,
-PCA3, RMST, Six3os1, SOX2OT and ST7-OT3 lncRNAs have multiple well
-conserved RNA-domains (See Figure~\ref{fig:5}).
+the lncRNAs with multiple RNA-domains are consistently preserved in
+the birds. The annotations of these domains lie in the same genomic
+region, in the same order as in the mammalian homologs. Thus
+supporting a high degree of evolutionary conservation for the entire
+lncRNA. In particular the HOXA11-AS1, PCA3, RMST, Six3os1, SOX2OT and
+ST7-OT3 lncRNAs have multiple, well conserved RNA-domains (See
+Figure~\ref{fig:5}).
 
@@
 RNA-domains 2 and 4 respectively, again the ordering of the domains
-was consistent with human. This suggests that the RMAST lncRNA is
+was consistent with human. This suggests that the RMST lncRNA is
 highly conserved. However, little is known about the function of this
@@
 both of which are also well conserved across the vertebrates (See
-Figure~\ref{fig:3}). DLEU2 is thought to be a tumor-suppressor
-gene as it is frequently deleted in malignant tumours \cite{Lerner:2009,Klein:2010}.
+Supplemental Figure~2). DLEU2 is thought to be a tumor-suppressor gene
+as it is frequently deleted in malignant tumours
+\cite{Lerner:2009,Klein:2010}.
 
@@
 The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
-\cite{Xu:1997}. Both are expressed in a broad ranch of tissues. Extensive
-research on BRCA1 has shown that it is involved in DNA repair
-\cite{}. The function of NBR2 remains unknown, yet its conservation
-across the vertebrates certainly implies a function (See
-Figure~\ref{fig:5}).
+\cite{Xu:1997}. Both are expressed in a broad ranch of
+tissues. Extensive research on BRCA1 has shown that it is involved in
+DNA repair \cite{Moynahan:1999}. The function of NBR2 remains unknown,
+yet its conservation across the vertebrates certainly implies a
+function (See Figure~\ref{fig:5}).
+
+{\bf more detail on the prostate lncRNAs?}
+
+Of the other classes of RNAs, none showed an unexpected degree of
+conservation or expansion within the avian lineage. The only exception
+being the snoRNA SNORD93 which we show has 92 copies in the tinamou
+genome, whereas it only has 1-2 copies in all the other vertebrate
+genomes.
 
@@
 
-\subsection*{Cis-regulatory elements}
-%% Histone3, IRE, IRES, K chan RES, SECIS, Antizyme FSE, Vimentin3, GABA3,
-%% CAESAR,
+%% Also, it would be good to check regarding the HOTAIRM/HOTTIP/HOXA11AS
+%% etc lncRNA elements that their syntenic location relativ to the HOXA genes
+%% (similar for SOXAS...) is correct. Mario checked for HOTAIRM and HOXA1/HOXA2
+%% and this is fine, it would be good for my piece of mind though it is
+%% at least OK in chicken, turkey, duck, zebrafinch. Also a word of caution
+%% of the Rfam models: if you look at HOTAIRM1_2, there's basically no
+%% covariation, looks pretty much like an ultraconserved element, so
+%% technically we don't really know whether this is an RNA element or,
+%% say, a TF binding site cluster that is functional at DNA level. Hence any
+%% evidence that we have indeed a transcript would be nice to have. We
+%% checked for HOTAIRM1 and it looks good in chicken. For the rest, it would be
+%% nice to have seen transcription it at least one bird :)
 
-The cis-regulatory RNAs are a group of RNA structures encoded on
-mRNAs. Generally they are involved in regulating the expression of the
-mRNA they are encoded within. Others may recode the translated protein
-product into an alternate sequence.
+\subsection*{RNA Losses, divergence or missing data?}
 
-This group includes the iron response element (IRE) \cite{Stevens:}
-and the histone $3^\prime$ UTR (histone3)
-\cite{Davila_Lopez:2008a}. These are structured motifs bound by
-regulatory proteins. The selenocysteine insertion sequence (SECIS) is
-a structured motif that recodes UGA stop codons to selenocysteines
-\cite{Lambert:2002} and the GABRA3 stem-loop is a structure recognised
-by the ADAR enzyme family. This enzyme edits adenine nucleotides to
-inosine, in this case recoding an isoleucine codon to methionine in
-exon 9 of the GABRA3 gene \cite{Ohlson:2007}.
+Much of the number of apparent losses and reduction in genomic
+sequence has been extensively discussed elsewhere
+\cite{Organ:2007}. Unsurprisingly, this reduction is reflected in the
+copy-number of RNA genes. Some of the most dramatic examples are the
+transfer RNAs and pseudogenes which average $\approx 900$ and $\approx
+580$ copies in the human, turtle and alligator genomes, the average
+copies numbers of these drop to $\approx 280$ and $\approx 100$ copies
+in the avian genomes.
 
-These regulatory elements and others, including an internal ribosome
-entry site (IRES), potassium channel RNA editing signal (K chan RES),
-Antizyme RNA frameshifting stimulation element (Antizyme FSE),
-vimentin $3^\prime$ UTR protein-binding region (Vimentin3) and a
-connective tissue growth factor (CTGF) $3^\prime$ UTR element (CAESAR)
-are conserved across a diverse group of vertebrates, including the
-bird lineages explored here (See Figure~\ref{fig:1}).
+The absence of several well-conserved ncRNA families from many or even most
+bird genomes is unlikely to represent true gene losses. This concerns in
+particular the telomerase RNA, the RNA components of RNase P and MRP, the
+minor spliceosomal snRNAs U4atac and U11, the selenocystein tRNA (tRNA-Sec)
+as well as the vault RNAs. In order to get an idea to what extent the
+absence of these RNAs from the \texttt{infernal-based} annotation is caused 
+by sequence divergence beyond the thresholds of the Rfam CMs and/or 
+missing or incomplete data, we complemented our analysis by dedicated
+searches for a few of these RNA groups.
 
-\subsection*{Exceptional RNAs}
+The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned for
+specificity and thus misses several occurrances that are easily found by
+\texttt{blastn} with $E\le 10^{-30}$. In some cases the sequences appear
+degraded at the ends, which may be explained e.g.\ by low sequence quality
+at the very ends of contigs or scaffolds. A \texttt{blastn} search also
+readily retrieves additional RNAse P and RNAse MRP RNAs, capturing only the
+best conserved regions. In many cases these additional candidates are
+incomplete or contain undetermined sequence, explaining why they are missed
+by the CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs
+in the majority of the genomes. An additional candidate could also be
+retrieved for telomerase RNA. Telomerase is well known to exhibit very poor
+sequence conservation and rapid variations in size that make it notoriously
+hard to identify by homology search \cite{Xie:08a}. The poor return thus
+does not come as a surprise. Since \texttt{blastn} searches remained
+unsuccessfull we constructed a sauropsid-specific CM for the vault RNA. In
+addition to the hits identified by the Rfam model we obtained three
+additional homologs. Vault RNAs, with a size of about 100 nt, exhibit
+conserved sequence patterns only at their ends, with essentially
+unconstrained sequence in the central part. Their identification is also a
+well-know difficult problem for homology search
+\cite{Stadler:09b}.
 
-%Background to this group...
+Our ability to find additional homologs for several RNA families that fill
+gaps in the abundance matrices (Supplemental Figure~5)
+strongly suggests that that conspicuous absences, in particular of LUCA and
+LECA RNAs, are caused by incomplete data in the current assemblies and
+sequence divergence rather then true losses.
 
-A number of other ncRNAs can also be found in Eukaryotic
-genomes. These do not fit into the main classes of RNA but still
-perform vital roles in the function and evolution of Eukaryotes.
-Their functions are diverse and many have not yet been characterised.
+\subsection*{Y RNA clusters}
 
-An example of an uncharacterised RNA is the ultraconserved element,
-uc.338 (also known as TUC338)
-\cite{Bejerano:2004,Bejerano:2006,Braconi:2011}. The uc.338 element
-is derived from a short interspersed element (SINE) called the
-lobe-finned fishes SINE (LF-SINE) as it is conserved between the
-coelacanth and mammals \cite{Bejerano:2006}. Analysis of the
-expression of uc.338 implies that it plays a role in the progression
-of hepatocellular carcinoma, possibly by influencing cell growth
-\cite{Braconi:2011}. This RNA is conserved in the birds and appears to
-have been duplicated in several lineages.
+Vertrebrate Y RNAs typically form a cluster comprising four
+well-defined paralog groups Y1, Y3, Y4, and Y5. In line with
+\emph{Mosig:07a} we find that the Y5 paralog family is absent from all
+bird genomes, while it is still present in both alligator and turtle,
+see Supplemental Figure~4. Within bird, we find an the conserved
+Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most cases
+consistent with breaks in the available sequence assemblies. In
+several genomes we observe one or a few additional Y RNA homologs
+unlinked to the canonical Y RNA cluster. These sequences can be
+identified unambiguously as derived members of one of the three
+ancestral paralog groups, they almost always fit less well to the
+consensus (as measured by the CM bit score of paralog group specific
+covariance models) than the paralog linked to cluster, and there is no
+indication that any of these additional copies is evolutionarily
+conserved over longer time scales. We therefore suggest that most or
+all of these interspersed copies are in fact pseudogenes.
 
-The Y RNA is an enigmatic ncRNA where we know very little about the
-function. It was discovered in the 1980s in ribonucleoprotein
-complexes \cite{Lerner:1981}. The function of the Y RNAs remain
-unknown, but evidence is emerging that they may be associated with DNA
-replication \cite{Christov:2006}. There are 4 functional Y RNAs
-encoded in the human genome, Y1, Y3, Y4 and Y5. However, there are
-hundreds of pseudogenised copies of the Y RNA scattered throughout the
-human genome \cite{Mosig:2007}. In the birds and other lizards, we
-identify between two and seven Y RNA paralogs (See
-Figure~\ref{fig:1}).
+\subsection*{MicroRNA Clusters}
 
-The Vault RNA forms a major component of the vault ribonucleoprotein
-complex, this is one of the largest particles found in the vertebrate
-cell; In fact, it is larger than the ribosome \cite{Kong:1999}. As yet
-not much is known about the function of Vault. The Vault RNA has been
-shown to be broadly conserved in metazoans
-\cite{Stadler:2009}. However, in the bird lineages it appears to have
-either been lost or diversified.
+Nearly all microRNAs that are broadly conserved in fish, amphibians and
+mammals are also conserved in the birds. Nevertheless, there are obvious
+instances of microRNAs lost in all birds. For example, mammalian and
+ambhibian genomes contain three loci of clustered microRNAs from the mir-17
+and mir-92 families \cite{Tanzer:04}. One of these clusters (cluster II,
+with families mir-106b, mir-93 and mir-25) was not found in turtles,
+crocodiles and birds, see Supplemental Figure~6.
 
-% Vault, 7SK, HAR1A
-
-%% \subsection*{Pseudogenes}
-
-%% The human genome is littered with pseudogenised copies of ncRNAs. The
-%% Dfam database of human repeats estimates that approximately $54\%$ of
-%% the human genome is composed of repeat elements
-%% \cite{Wheeler:2013}. Of these annotations XXX\% are derived from
-%% ncRNAs, including the large Alu family of repeats, which is derived
-%% from a retransposed SRP RNA pseudogene \cite{}.
-
-%% tRNA, RSV RNA,
-%%  \cite{International_Chicken_Genome_Sequencing_Consortium:2004}
-
-%% \emph{Endogenous retroviruses:} RSV RNA, Alfamo CPB
-
-\subsection*{Contamination}
-
-Bacterial families can be used to identify problematic sequences that
-are likely to be the result of contamination from non-avian sources.
+The microRNA family let-7 is the most abundant microRNA family with 14
+paralogs in human. These genes also localize in 7 genomic clusters,
+together with mir-100 and mir-125 miRNA families (see previous study on the
+evolution of the let-7 miRNA cluster in \cite{Hertel:2012}). In Sauropsids
+we observed that cluster A - that is strongly conserved in vertebrates has
+been completely lost in the avian lineage. Another obious lost in birds is
+cluster F, containing two let-7 microRNA paralogs. Cluster H, on the other
+hand has been retained in all oviparous animals and completely lost later,
+after the split of Theria. See Supplemental Figure~7 for details.
 
@@
 1.1 and the covariance models from the Rfam database
-v11.0\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}. All
-matches above the curated GA threshold were included. Subsequently,
-all hits with an E value greater than 0.0005 were discarded, so only
-matches which passed the model-specific GA threshold and had an E-value smaller than 0.0005 were retained.
+v11.0 \cite{Gardner:2011a,Burge:2013}. All matches above the curated GA
+threshold were included. Subsequently, all hits with an E-value
+greater than 0.0005 were discarded, so only matches which passed the
+model-specific GA threshold and had an E-value smaller than 0.0005
+were retained.
 
-tRNA-scan version 1.3.1, ... \cite{Lowe:1997,Chan:2009}
+In order to obtain good annotations of tRNA genes we also ran the
+specialist tRNA-scan version 1.3.1 annotation tool. This method also
+uses covariance models to identify tRNAs. However it also uses some
+heuristics to increase the search-speed, annotates the Isoacceptor
+Type of each prediction and uses sequence analysis to infer if
+predictions are likely to be functional or tRNA-derived pseudogenes
+\cite{Lowe:1997,Chan:2009}.
 
 Rfam matches and the tRNA-scan results for families belonging to the
-same clan were then competed so that only the best match was retained.
-Matches were further sorted into taxonomic groups {\bf ONLY MATCHES IN 5 OR MORE ...}
+same clan were then ``competed'' so that only the best match was
+retained for any genomic region \cite{Gardner:2011a}.  To further
+increase the specificity of our annotations we filtered out families
+that were indentified in four or fewer of the 51 vertebrate species we
+have analysed in this work. These filtered families largely
+corresponded to bacterial contamination within the genomic sequences.
 
@@
 \begin{figure}[ht]
-  \includegraphics[width=0.4\textwidth]{figures/RNA.pdf}
+  \includegraphics[width=0.45\textwidth]{figures/lncRNA.pdf}
   \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
-  \includegraphics[width=0.4\textwidth]{figures/snoRNA-human-yeast-correspondences.pdf}
-  \caption[]{This figure illustrates the number of copies of Rfam and
-    tRNAscan annotations for each RNA family in each avian species and
-    outgroups. The families are ordered alphabetically vertically and
-    the species are in phylogenetic order horizontally. Darker shades
-    indicate high copy numbers, lighter shades indicate fewer, white
-    indicates no predictions were made.}\label{fig:1}
-\end{figure}
-
-%% \begin{figure}[ht]
-%% \includegraphics[width=0.65\textwidth]{figures/snoRNA-human-yeast-correspondences.pdf}
-%% \caption[]{.}\label{fig:2}
-%% \end{figure}
-\begin{figure}[ht]
-  \includegraphics[height=0.95\textheight]{figures/miRNA.pdf}
-  \includegraphics[height=0.95\textheight]{figures/snoRNA.pdf}
   \caption[]{Heatmaps showing the prescence/abscence and approximate
-    copy number of {\bf miRNA} families on the right and {\bf snoRNA} families on
-    the left.}\label{fig:3}
-\end{figure}
-
-\begin{figure}[ht]
-  \includegraphics[width=0.45\textwidth]{figures/lncRNA.pdf}
-  \includegraphics[width=0.45\textwidth]{figures/tRNA.pdf}
-  \caption[]{Heatmaps showing the prescence/abscence and approximate
-    copy number of {\bf lncRNA} families on the right and {\bf tRNA}
-    families on the left.}\label{fig:5}
+    copy number of {\bf lncRNA} families.}\label{fig:5}
 \end{figure}

### 2013-09-23 paper/supplementary-material.tex
updated manuscript.

@@
 
-\begin{abstract}
-...
-\end{abstract}
-
-%% Non-coding RNAs in bird genomes
-
-%% Paul P. Gardner1,2, Sarah W. Burge3, Jana Hertel5, Maria Ninova4, Stephanie Kehr5, Anne Nitsche5, Mario Fasold5, Tammy E. Steeves1, Sam Griffiths-Jones4, Peter Stadler5
-
-%% 1 School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
-2 Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
-3 European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
-4Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
-5 Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, Hartelstrasse 16-18, D-04107 Leipzig, Germany
-
-%% Talking points:
-%% 1.
-Although the majority of classical eukaryotic ncRNAs are conserved in birds, several notable classical ncRNAs are not.
-Homology searches in associated genes indicate these apparent "losses" are most likely due to divergence.
-%% 2.
-Of the 59 snoRNA conserved between humans and yeast, we find 45 are conserved in birds.
-Homology searches in associated genes for the remaining non-conserved snoRNAs indicate these apparent losses are indeed due to multiple losses in several bird lineages.
-%% 3. We find that 3 families of broadly conserved vertebrate microRNA families are lost in the avian lineage. We also annotate a number of bird-specific microRNAs.
-%% 4. Several well-characterised lncRNA families of known function are conserved between mammals and birds.
-%% 5. We discuss the conservation of HOX associated lncRNAs and several lncRNAs that have been implicated in processes associated with cancer.
-
-%% %% Draft Abstract
-%% We present the results of a large-scale bioinformatic annotation of non-coding RNAs in 48 avian genomes.
-We use probabilistic models of hand-curated families from the Rfam database to infer conserved RNA families within each avian genome.
-We supplement these annotations with predictions from the tRNA annotation tool, tRNAscan-SE, microRNAs from miRBase and selected sequences from analyses of the chicken transcriptome.
-Although we show extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds, we also demonstrate apparent ``losses'' in several RNA families.
-These include the divergence of some classical ncRNAs and the genuine loss of several snoRNAs and microRNAs.
-In contrast, we show a signicant number of lncRNAs are surprisingly well conserved between birds and mammals including several intriguing cases where the reported mammalian lncRNA function is not conserved in birds.
-These combined results illustrate the utility of applying homology based methods for annotating vertebrate genomes and illustrate many complex evolutionary patterns within the avian ncRNA cohort.
-
-%% LNCRNA:
+%\begin{abstract}
+%...
+%\end{abstract}
 
@@
 
-\section*{Supplementary introduction}
-...
+%\section*{Supplementary introduction}
+%...
 
 \section*{Supplementary results}
-In the following we explore ...
 
-\section*{Supplementary methods}
-...
+In the following we explore in further detail the results that were
+not discussed in the main manuscript.
+
+\subsection*{Classic RNAs: LUCA and LECA}
+
+Many RNA families constitute the most evolutionarily conserved genes
+across all life on this planet \cite{Jeffares:1998}. Examples of RNAs
+derived from the Last Universal Common Ancestor (LUCA) include, the
+transfer RNAs (tRNA), ribosomal RNAs (rRNA), RNA components of RNase P
+(RNase P RNA), RNase MRP (RNase MRP RNA) and the signal recognition
+particle (SRP RNA). Other classes of RNA are likely to have been
+components of the Last Eukaryotic Common Ancestor (LECA) include the
+telomerase RNA, major spliceosomal RNAs (U1, U2, U4, U5, and U6) and
+the minor spliceosomal RNAs (U11, U12, U4atac, and U6atac)
+\cite{Hoeppner:2012}.
+
+Unsurprisingly, the bulk of these classes of RNAs are well represented
+across the bird genomes (See Figure~\ref{fig:1}). However, there
+appear to have been ``losses'' of a few of these RNAs in certain bird
+species. Some of these may be due to sequence divergence, of which
+there are several notable examples e.g.
+\cite{Leonardi:2008,Webb:2008,Mao:,Lai:2010,Chan:2011}. Other losses
+may be due to incomplete genome coverage.
+
+A number of the classic RNAs are incorporated into RNA-protein
+complexes (RNPs) involved in core cellular processes. An example of
+this are the spliceosomal RNAs. Based upon the presence/absence
+patterns of the major spliceosomal RNAs they are all well represented
+in these genome sequences. Except for the U4 RNA in cormorant and the
+U5 RNA in the bee eater. Again these two genomes are low coverage
+suggesting these genes weren't captured in the current assembly. The
+minor spliceosomal RNAs are more interesting, the U4atac and U11
+snRNAs show widespread patterns of loss, even in some of the high
+coverage genomes. These RNAs are frequently missed in bioinformatic
+screens. Indicating either frequent loss \cite{Davila_Lopez:2008} or
+sequences that have diverged beyond the ability of detection by
+covariance models \cite{Marz:2008}.
+
+The telomerase RNA is also largely missing from the avian
+annotations. This RNA acts as a template for the telomerase enzyme
+that extends the telomeres found on chromosome ends. It is only found
+in the chicken, bald eagle, kea, budgerigar, crow and
+zebrafinch. Homology searches searches with the telomerase reverse
+transcriptase (TERT) protein show that the protein component of the
+telomerase RNP is conserved across all the bird genomes (data not
+shown). This pattern of presumably divergent telomerase RNA and
+conserved telomerase protein has been noted previously, most notably
+in the fungi \cite{Leonardi:2008,Webb:2008}.
+
+The the RNA components of RNase P and RNase MRP also appear to have
+undergone dramatic losses within the bird lineage. RNase P is required
+for the maturation of tRNA, the paralogous enzyme, RNase MRP is
+required for the maturation of rRNA. Each RNP cleaves smaller RNAs
+from larger transcripts \cite{Lopez:2009}. It is unlikely that the
+these genes have been lost in any of the birds. Homology searches with
+the RNase associated protein coding genes (POP1, POP4, POP5, POP7,
+RPP1, RPP14, RPP25, RPP38, RPP40 and RPR2), identified viable homologs
+of each in all of the bird genomes \cite{Rosenblad:2006} (data not
+shown). This suggests that the bird RNase P and MRP RNAs may have
+diverged slightly from the canonical models.
+
+The 5.8 S component of the ribosome in the turtle, turkey bustard,
+hoatzin, flamingo, tropicbird, seriema, owl, cuckoo roller, trogon,
+bee eater and falcon appears to have been lost (See
+Figure~\ref{fig:1}). The rRNA repeats are frequently not assembled,
+consequently it is not surprising to see ``losses'' in these \cite{Floutsakou:2013}.
+Furthermore, the genomes for these species are also
+low-coverage.
+
+\subsection*{Small nucleolar RNAs}
+
+Small nucleolar RNAs (snoRNAs) are important ncRNAs that participate
+in the maturation of other functional RNAs \cite{Gardner:2010}. The
+bulk of the characterised snoRNAs guide either methylation or
+pseudouridylation modifications, primarily of rRNAs but also
+spliceosomal RNAs. The two types of modifications are guided by two
+different types of RNA, the box C/D and the H/ACA snoRNAs
+respectively, each with a characteristic cohort of motifs and
+secondary structures \cite{Marz:2011}.
+
+There are 66 ribosomal modification sites, guided by 59 snoRNA
+families, that are preserved between \emph{H. sapiens} and
+\emph{S. cerevisiae} \cite{Lestrade:2006}. Of these, 45 snoRNA
+families are conserved in the bird dataset.
+Over a third of the apparent losses of the yeast-human conserved
+snoRNA families appear to cluster on 2 locii of the ancestral
+vertebrate genome. We investigated these losses further.
+
+The first cluster is found at chr11:62620797-62622484 on the human
+genome (hg19) and contains SNORD27, SNORD29 and SNORD31 of the
+human-yeast conserved snoRNAs. These snoRNAs are located in the
+inside-out gene SNHG1 which hosts a total of eight C/D box snoRNAs:
+SNORD25, SNORD26, SNORD27, SNORD28, SNORD29, SNORD22, SNORD30 and
+SNORD31 \cite{Tycowski:1996}. Each of which are also found in the
+alligator and turtle genomes within a 3-4 KB locus, yet these have
+largely been lost in the birds. Notable exceptions are five of the
+eight are located in the tinamou genome, these are located on the same
+scaffold and are within 2 KB of each other. Implying that SNHG1 is
+conserved in the tinamou. Loci with four of the eight snoRNAs can be
+found in zebrafinch, ground-finch, and bald eagle. Still, three of the
+eight are located in the ostrich, crow, and cockoo genomes, again
+within 2 KB of each other on the same scaffolds. This complex pattern
+of loss could be attributed to many different models, e.g. multiple
+losses in birds, poor homology modelling or incomplete genome
+sequences.
+
+The second cluster is located at chr19:49993222-49994231 on the human
+genome (hg19) and contains two copies of SNORD33 and one SNORD34 all
+within a 1 KB genomic region. The turtle and alligator genomes retain
+the two copies of SNORD33 yet don't have an obvious SNORD34 gene on
+the same scaffold. Within the bird genomes, the crow and rifleman each
+retain a single SNORD33 and SNORD34 gene on the same scaffold. While
+the ground-finch and bald eagle retain a single SNORD33 and the
+zebrafinch and seriema retain a single SNORD34 (See
+Figure~\ref{fig:3}).  In human these snoRNAs are intronic to the host
+gene, ribosomal protein L13a (RPL13A). Based on BLASTP (version
+2.2.18) homology searches for the RPL13A gene, the protein is
+conserved in the human and turtle genomes and in the bald eagle, crow,
+rifleman and zebrafinch avian genomes (data not shown). Therefore the
+conservation patterns of the RPL13A host gene and corresponding
+intronic snoRNAs largely mimic each other, each independently supports
+a pattern of loss of the RPL13A gene and the intronic snoRNAs that it
+hosts in the bird genomes.
+
+\subsection*{MicroRNAs}
+MicroRNAs are an important class of non-coding RNA. They have been
+found in the genomes of Chromalveolata \cite{Cock:2010,Huang:2011},
+Metazoa \cite{Lee:1993,Lau:2001,Hertel:2006}, Mycetozoa
+\cite{Hinas:2007,Avesson:2012}, Viridiplantae
+\cite{Reinhart:2002,Fattash:2007,Axtell:2007,Molnar:2007} and Viruses
+\cite{Pfeffer:2004,Ouellet:2008,Pfeffer:2005,Landgraf:2007}. The
+miRNAs have been shown to regulate the expression of large numbers of
+messenger RNAs \cite{Lim:2005}. The mature miRNA product is generally
+22 nucleotides long which is usually processed from a larger RNA that
+is characterised by a stable hairpin-shaped secondary structure.
+
+Chicken and zebrafinch are the only birds with previously annotated
+microRNAs. We searched for homologs of these and other vertebrate
+microRNAs in the genomes of the 48 birds, American alligator and green
+turtle. Overall, we annotate a total of 16617 putative microRNA loci,
+homologous to 543 known microRNA genes, of which 487 are annotated in
+chicken and/or zebra finch, while 56 have been so far known only in
+non-avian vertebrates. The numbers of annotated loci in the individual
+species are approximately equal - 300-400 per species, except for the
+turkey (Meleagris gallopavo) where we identified 543 sequences
+homologous to known microRNAs.
+
+Nearly all the microRNAs that are broadly conserved in fish,
+amphibians and mammals are also conserved in the birds. Nevertheless,
+there are obvious instances of microRNAs lost in all birds. For
+example, mammalian, amphibian and fish genomes contain three loci of
+clustered microRNAs from the mir-17 and mir-92 families. One of these
+clusters (mir-93~25) was not found in turtles, crocodiles and
+birds. In addition, we can confidently identify a further 3 microRNA
+families that are present in mammals, and turtle and/or crocodile, but
+not in any avian genome (mir-150, mir-208, mir-590). This suggests
+that these sequences were lost in the last common ancestor of
+archosaurs or birds. There are also a number of microRNAs that are
+predicted to be present in turtles and/or crocodiles, and only a small
+number of bird genomes. Indeed, there are many missing annotations,
+species-specific and otherwise, that are not consistent with the
+consensus phylogeny, and could be due to either incomplete genomes or
+widespread microRNA loss.
+
+The turkey genome contains a high number (190) of microRNAs so far
+found only in chicken, which account for the higher number of
+annotated sequences in this genome compared with other birds. This is
+consistent with its phylogenetic position as the closest chicken
+relative among the examined birds. However, 101 chicken microRNAs
+have no homolog in the turkey or other bird genomes, suggesting that
+these genes are chicken-specific. This is consistent with previous
+reports of large number of speciesspecific microRNAs in all animals,
+and supports the view of fast microRNA turnover during animal
+evolution.
+
+\subsection*{Cis-regulatory elements}
+%% Histone3, IRE, IRES, K chan RES, SECIS, Antizyme FSE, Vimentin3, GABA3,
+%% CAESAR,
+
+The cis-regulatory RNAs are a group of RNA structures encoded on
+mRNAs. Generally they are involved in regulating the expression of the
+mRNA they are encoded within. Others may recode the translated protein
+product into an alternate sequence.
+
+This group includes the iron response element (IRE) \cite{Stevens:}
+and the histone $3^\prime$ UTR (histone3)
+\cite{Davila_Lopez:2008a}. These are structured motifs bound by
+regulatory proteins. The selenocysteine insertion sequence (SECIS) is
+a structured motif that recodes UGA stop codons to selenocysteines
+\cite{Lambert:2002} and the GABRA3 stem-loop is a structure recognised
+by the ADAR enzyme family. This enzyme edits adenine nucleotides to
+inosine, in this case recoding an isoleucine codon to methionine in
+exon 9 of the GABRA3 gene \cite{Ohlson:2007}.
+
+These regulatory elements and others, including an internal ribosome
+entry site (IRES), potassium channel RNA editing signal (K chan RES),
+Antizyme RNA frameshifting stimulation element (Antizyme FSE),
+vimentin $3^\prime$ UTR protein-binding region (Vimentin3) and a
+connective tissue growth factor (CTGF) $3^\prime$ UTR element (CAESAR)
+are conserved across a diverse group of vertebrates, including the
+bird lineages explored here (See Figure~\ref{fig:1}).
+
+\subsection*{Exceptional RNAs}
+
+%Background to this group...
+
+A number of other ncRNAs can also be found in Eukaryotic
+genomes. These do not fit into the main classes of RNA but still
+perform vital roles in the function and evolution of Eukaryotes.
+Their functions are diverse and many have not yet been characterised.
+
+An example of an uncharacterised RNA is the ultraconserved element,
+uc.338 (also known as TUC338)
+\cite{Bejerano:2004,Bejerano:2006,Braconi:2011}. The uc.338 element
+is derived from a short interspersed element (SINE) called the
+lobe-finned fishes SINE (LF-SINE) as it is conserved between the
+coelacanth and mammals \cite{Bejerano:2006}. Analysis of the
+expression of uc.338 implies that it plays a role in the progression
+of hepatocellular carcinoma, possibly by influencing cell growth
+\cite{Braconi:2011}. This RNA is conserved in the birds and appears to
+have been duplicated in several lineages.
+
+The Y RNA is an enigmatic ncRNA where we know very little about the
+function. It was discovered in the 1980s in ribonucleoprotein
+complexes \cite{Lerner:1981}. The function of the Y RNAs remain
+unknown, but evidence is emerging that they may be associated with DNA
+replication \cite{Christov:2006}. There are 4 functional Y RNAs
+encoded in the human genome, Y1, Y3, Y4 and Y5. However, there are
+hundreds of pseudogenised copies of the Y RNA scattered throughout the
+human genome \cite{Mosig:2007}. In the birds and other lizards, we
+identify between two and seven Y RNA paralogs (See
+Figure~\ref{fig:1}).
+
+The Vault RNA forms a major component of the vault ribonucleoprotein
+complex, this is one of the largest particles found in the vertebrate
+cell; In fact, it is larger than the ribosome \cite{Kong:1999}. As yet
+not much is known about the function of Vault. The Vault RNA has been
+shown to be broadly conserved in metazoans
+\cite{Stadler:2009}. However, in the bird lineages it appears to have
+either been lost or diversified.
+
+\subsection*{Contamination}
+
+Bacterial families can be used to identify problematic sequences that
+are likely to be the result of contamination from non-avian sources.
+
+%{\bf WRITE MORE HERE!}
 
@@
 \begin{figure}[ht]
+  \includegraphics[width=0.4\textwidth]{figures/RNA.pdf}
+  \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
+  \includegraphics[width=0.4\textwidth]{figures/snoRNA-human-yeast-correspondences.pdf}
+  \caption[]{This figure illustrates the number of copies of Rfam and
+    tRNAscan annotations for each RNA family in each avian species and
+    outgroups. The families are ordered alphabetically vertically and
+    the species are in phylogenetic order horizontally. Darker shades
+    indicate high copy numbers, lighter shades indicate fewer, white
+    indicates no predictions were made.}\label{fig:1}
+\end{figure}
+
+\begin{figure}[ht]
   \includegraphics[height=0.95\textheight]{figures/miRNA.pdf}
@@
 \begin{figure}[ht]
-  \includegraphics[width=0.45\textwidth]{figures/lncRNA.pdf}
   \includegraphics[width=0.45\textwidth]{figures/tRNA.pdf}
   \caption[]{Heatmaps showing the prescence/abscence and approximate
-    copy number of {\bf lncRNA} families on the right and {\bf tRNA}
-    families on the left.}\label{fig:5}
+    copy number of {\bf tRNA} families.}\label{fig:5}
 \end{figure}

### 2013-09-23 scripts/gffs2heatmaps.pl

@@ -281,6 +281,8 @@ print "System calls: creating [$rDir/snoRNA-human-yeast-correspondences.dat] & r
 #GAS5 snoRNAs: \|SNORD81\$\|SNORD47\$\|SNORD80\$\|SNORD79\$\|SNORD78\$\|SNORD44\$\|SNORD77\$\|SNORD76\$\|SNORD75\$\|SNORD74\$
 #DAMN THIS IS NASTY!
 system("egrep \47^human\|snoR38\$\|SNORA13\$\|SNORA16\$\|SNORA2\$\|SNORA21\$\|SNORA26\$\|SNORA27\$\|SNORA28\$\|SNORA3\$\|SNORA36\$\|SNORA4\$\|SNORA44\$\|SNORA48\$\|SNORA5\$\|SNORA50\$\|SNORA52\$\|SNORA56\$\|SNORA58\$\|SNORA62\$\|SNORA64\$\|SNORA65\$\|SNORA66\$\|SNORA69\$\|SNORA7\$\|SNORA74\$\|SNORA76\$\|SNORA8\$\|SNORA9\$\|SNORD12\$\|SNORD14\$\|SNORD15\$\|SNORD16\$\|SNORD17\$\|SNORD18\$\|SNORD2\$\|SNORD24\$\|SNORD27\$\|SNORD29\$\|SNORD31\$\|SNORD33\$\|SNORD34\$\|SNORD35\$\|SNORD36\$\|SNORD38\$\|SNORD41\$\|SNORD43\$\|SNORD46\$\|SNORD51\$\|SNORD52\$\|SNORD57\$\|SNORD59\$\|SNORD60\$\|SNORD62\$\|SNORD65\$\|SNORD74\$\|SNORD77\$\|SNORD88\$\|SNORND104\$\|snosnR60_Z15\$\47 $rDir/snoRNA.dat > $rDir/snoRNA-human-yeast-correspondences.dat");
+#$rDir/allRNA.dat
+#egrep -i '^human|HOXA11-AS1|PCA3|RMST|Six3os1|SOX2OT|ST7-OT3|HOTAIRM1|HOTTIP|RMST|DLEU2|NBR2|SNORD93|miR-15$|miR-16$' ../data/R/allRNA.dat
 
 system("R CMD BATCH --no-save scripts/heatmaps.R");
 

### 2013-10-21 paper/manuscript.tex
added Marco's RNA-seq validated RNAs.
Updated manuscript, scripts, fixed chicken seqIds, ...

@@
 %\title{Wins and losses: Non-coding RNAs of bird genomes}
-\title{Fly away Peter, come back Paul: the gains and losses of avian Non-coding RNAs}
+\title{Fly away Peter, come back Paul: the conservation and losses of avian non-coding RNAs}
 
@@
 supplement these annotations with predictions from the tRNA annotation
-tool, tRNAscan-SE and microRNAs from miRBase. Although we show
-extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and
-more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds,
-we also demonstrate apparent ``losses'' in several RNA families. These
-include the divergence of some classical ncRNAs and the loss of
-several snoRNAs and microRNAs. In contrast, we show a significant
-number of lncRNAs are surprisingly well conserved between birds and
-mammals including several intriguing cases where the reported
-mammalian lncRNA function is not conserved in birds. These combined
+tool, tRNAscan-SE and microRNAs from miRBase.  Although we show a
+significant number of lncRNAs are surprisingly well conserved between
+birds and mammals including several intriguing cases where the
+reported mammalian lncRNA function is not conserved in birds.  In
+contrast, we show extensive conservation of classical ncRNAs (e.g.,
+tRNAs and rRNAs) and more recently discovered ncRNAs (e.g., snoRNAs
+and miRNAs) in birds, we also demonstrate apparent ``losses'' in
+several RNA families. These include the divergence of some classical
+ncRNAs and the loss of several snoRNAs and microRNAs.  These combined
 results illustrate the utility of applying homology based methods for
@@
 they pose serious research challenges, particularly for the field of
-genomics. They lack the strong statistical signals associated with
-protein coding genes, e.g. open reading frames, G+C content and
-codon-usage biases \cite{Rivas:2000}.
+genomics. For example, they lack the strong statistical signals
+associated with protein coding genes, e.g. open reading frames, G+C
+content and codon-usage biases \cite{Rivas:2000}.
 
@@
 microRNAs (miRNAs) and the long non-coding RNAs (lncRNAs)
-\cite{Hoeppner:2012}. The publication of avian genomes provides
-exciting opportunities to to explore ncRNA conservation in
-unprecedented detail.
+\cite{Hoeppner:2012}. The publication of 48 avian genomes, including
+the previously published chicken
+\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
+finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the newly
+published 45 avian genomes \cite{}, provides exciting opportunities to
+to explore ncRNA conservation in unprecedented detail.
 
 Despite promising techniques such as RNA-seq \cite{Croucher:2010},
-homology based methods namely covariance models (CMs) are state of the
+homology based methods namely covariance models (CMs) remain state of the
 art for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
@@
 
-%Still needs a {\bf what have we done} paragraph!
+In the following we explore the conservation patterns of the major
+classes of avian ncRNA in further detail. The collection of ncRNA
+sequences is generally biased towards model organisms
+\cite{Gardner:2010,Hoeppner:2012}. We focus our report on the unusual
+results within the avian lineages. These are either unexpectedly
+well-conserved RNAs or unexpectedly poorly-conserved RNAs. The former
+are RNAs we would not have expected to be conserved between the birds
+and the organisms these genes were initially identified in; Usually,
+this hypothesis is based upon the function of the RNA which is not
+conserved in avian species. The latter are apparent losses of RNA
+genes that were expected to be conserved; Usually, this hypothesis is
+based upon the conservation of these RNAs in other vertebrate species.
+We use three models of ``loss'' that explain the data: Firstly, these
+could be genuine cases of an ancestral gene-loss along the avian
+lineage. Secondly, this could be a case of ``divergence'' where a RNA
+gene has undergone significant sequence and structural alterations, so
+much so, that our homology search tools can no longer detect a
+relationship between vertebrate exemplars and the avian
+varieties. Thirdly, we consider the possibility that the available
+genome assemblies have independently failed to capture these genes.
 
 \section*{Results}
-In the following we explore the conservation patterns of the major
-classes of ncRNA in further detail. The collection of ncRNA sequences
-is generally biased towards model organisms
-\cite{Gardner:2010,Hoeppner:2012}. We focus our analysis on the
-unusual results within the avian lineages. These are either
-unexpectedly well-conserved RNAs or unexpectedly poorly-conserved
-RNAs. The former are RNAs we would not have expected to be conserved
-between the birds and the organisms these genes were initially
-identified in; Usually, this hypothesis is based upon the function of
-the RNA which is not conserved in avian species. The latter are
-apparent losses of RNA genes that were expected to be conserved;
-Usually, this hypothesis is based upon the conservation of these RNAs
-in other vertebrate species. We have three models of ``loss'' that we
-use to explain the data: Firstly, these could be genuine cases of an
-ancestral gene-loss along the avian lineage. Secondly, this could be a
-case of ``divergence'' where a RNA gene has undergone significant
-sequence and structural alterations, so much so, that our homology
-search tools can no longer detect a relationship between vertebrate
-exemplars and the avian varieties. Thirdly, we consider the
-possibility that the available genome assemblies have independently
-failed to capture these genes.
 
@@
 characterised in mammalian species, particularly human and
-mouse. Consequently, generally do not expect these to be conserved
+mouse. Consequently, we generally do not expect these to be conserved
 outside of mammalia. Notable examples include Xist \cite{Duret:2006}
-and H19 \cite{Smits:2008}.
+and H19 \cite{Smits:2008}. Some of the main results of this group are
+discussed in the comparative analysis of avian genomes article \cite{}.
 
-In general Rfam cannot include the entire length of any large, spliced
+In general, Rfam cannot include the entire length of any large, spliced
 RNAs. This is a limitation of the covariance-models used for the
@@
 RNA-domains are located at chr7:27135743-27245922 (hg19) and are
-modelled by 15 RNA-domains in total, the % (5, 4 and 6 respectively)
+modelled by 15 RNA-domains in total, the 
 lncRNAs are ordered HOTAIRM1 (5 RNA-domains) $\rightarrow$ HOXA11-AS1
@@
 the HOX cluster lncRNA RNA-domains were predicted on the same
-scaffold, this gene order was preserved.
+scaffold, this gene order was also preserved.
 %egrep 'HOXA11-AS1|HOTAIR|HOTTIP' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
@@
 
-%(MIR1251 is in the same locus)
-%PCA3 RNA-domains inverted relative to human...
-
 The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
@@
 
-{\bf more detail on the prostate lncRNAs?}
-
 Of the other classes of RNAs, none showed an unexpected degree of
@@
 genomes.
-
-%(neighbor of BRCA1 gene 2)
-%MHM results!!!!
-%% Another of the functions associated with lncRNAs is chromosomal dosage
-%% compensation. In mammals this is carried out by the lncRNAs Xist and
-%% Tsix \cite{}. In birds it has been proposed that the lncRNA MHM might
-%% play a related role \cite{}.
-
-%DLEU2, frequently deleted in malignancy, functions as a critical host
-%gene of the cell cycle inhibitory microRNAs miR-15a and miR-16-1
-
-%% TUMOR-SUPPRESSOR:
-%% DLEU2, part of a tumor-suppressor cluster (with two microRNAs), most recent refs here:
-%% http://www.ncbi.nlm.nih.gov/pubmed/19591824
-%% http://www.ncbi.nlm.nih.gov/pubmed/20060366
-%% SMAD5, tumor suppressor candidate:
-%% http://www.ncbi.nlm.nih.gov/pubmed/10049768?dopt=Abstract
-
-%% DEVELOPMENT:
-%% HOXA11, most complete reference list here:
-%% http://www.lncrnadb.org/Detail.aspx?TKeyID=171
-%% Six3OS, retinal development:
-%% http://www.ncbi.nlm.nih.gov/pubmed/21936910
-%% ZEB2, brain development:
-%% http://www.ncbi.nlm.nih.gov/pubmed/18347095?dopt=Abstract
-
-%% MAT2A:
-%% http://www.ncbi.nlm.nih.gov/pubmed/1426236%2C9703951?dopt=Abstract
-
-%% RMST, doesn't look we know it does (also shows up in review paper at top):
-%% http://www.ncbi.nlm.nih.gov/pubmed/20062813
-
-%% ST7-OT, another candidate:
-%% http://www.ncbi.nlm.nih.gov/pubmed/12213198?dopt=Abstract
-
-%% PART1, meant to be human prostate-specific, birds don't even have a prostate:
-%% http://www.ncbi.nlm.nih.gov/pubmed/10706094?dopt=Abstract
-%% PCA3, prostate/androgen receptor signalling (curious there's heaps of it in male and female birds? same for PART1?):
-%% http://www.ncbi.nlm.nih.gov/pubmed/?term=ferreira+and+pca3
-
-%% Also, it would be good to check regarding the HOTAIRM/HOTTIP/HOXA11AS
-%% etc lncRNA elements that their syntenic location relativ to the HOXA genes
-%% (similar for SOXAS...) is correct. Mario checked for HOTAIRM and HOXA1/HOXA2
-%% and this is fine, it would be good for my piece of mind though it is
-%% at least OK in chicken, turkey, duck, zebrafinch. Also a word of caution
-%% of the Rfam models: if you look at HOTAIRM1_2, there's basically no
-%% covariation, looks pretty much like an ultraconserved element, so
-%% technically we don't really know whether this is an RNA element or,
-%% say, a TF binding site cluster that is functional at DNA level. Hence any
-%% evidence that we have indeed a transcript would be nice to have. We
-%% checked for HOTAIRM1 and it looks good in chicken. For the rest, it would be
-%% nice to have seen transcription it at least one bird :)
 
@@
 
-Vertrebrate Y RNAs typically form a cluster comprising four
+Vertebrate Y RNAs typically form a cluster comprising four
 well-defined paralog groups Y1, Y3, Y4, and Y5. In line with
-\emph{Mosig:07a} we find that the Y5 paralog family is absent from all
+\cite{Mosig:07a} we find that the Y5 paralog family is absent from all
 bird genomes, while it is still present in both alligator and turtle,
@@
 
-\subsection*{MicroRNA Clusters}
+\subsection*{MicroRNA clusters}
 
@@
 we observed that cluster A - that is strongly conserved in vertebrates has
-been completely lost in the avian lineage. Another obious lost in birds is
+been completely lost in the avian lineage. Another obvious loss in birds is
 cluster F, containing two let-7 microRNA paralogs. Cluster H, on the other
@@
 %avian-specific alterations in gene regulation.
+
+In this work we have summarised the main results after carefully
+annotating non-coding RNAs in genome sequences. We have identified
+previously unrecognised conservation of ncRNAs in avian genomes as
+well as some surprising ``losses'' of otherwise well conserved
+ncRNAs. We suspect many of these losses are due to a combination of
+limitations in the homology search tools that we use for annotation
+and the ability of ncRNAs to tolerate large amounts of sequence
+variation while remaining functional, rather than \emph{bona fide}
+gene loss. In some cases these losses could be due to missing data
+from the genome assemblies, but this unlikely to be the case for
+multiple independant assemblies.
+
+These results indicate we are still in the very early phases of
+determining the functions of many RNA families. This is illustrated by
+the fact that the reported functions of some ncRNAs are
+mammal-specific, yet some are also found in bird genomes.
 

### 2013-10-21 scripts/gffs2heatmaps.pl

@@ -50,7 +50,15 @@ if (-s "$rDir/allRNA.dat"){
     }
     close(W);
 }
-$whitelist{'SNORD34'}=1;
+my @whiteList=qw(SNORD34
+DLEU2_1 DLEU2_2 DLEU2_3 DLEU2_4 DLEU2_5 DLEU2_6 HOXA11-AS1_1 HOXA11-AS1_2 HOXA11-AS1_3 HOXA11-AS1_4 HOXA11-AS1_5 HOXA11-AS1_6 mir-15_1 mir-15_2 mir-15_3 mir-16_1 mir-16_2 mir-16_3 NBR2 PCA3_1 PCA3_2 RMST_1 RMST_10 RMST_2 RMST_3 RMST_4 RMST_5 RMST_6 RMST_7 RMST_8 RMST_9 Six3os1_1 Six3os1_2 Six3os1_3 Six3os1_4 Six3os1_5 Six3os1_6 Six3os1_7 SNORD93 SOX2OT_exon1 SOX2OT_exon2 SOX2OT_exon3 SOX2OT_exon4 ST7-OT3_1 ST7-OT3_2 ST7-OT3_3 ST7-OT3_4
+Metazoa_SRP RNase_MRP RNaseP_nuc Telomerase-vert U1 U11 U12 U2 U4 U4atac U5 U6 U6atac Vault
+); 
+
+foreach my $wl (@whiteList){
+    $whitelist{$wl}=1;
+}
+
 #replace with a family2type hash:
 my %lncRNAlist;
 if (-s "data/rfam11_lncRNAs.txt"){
@@ -123,7 +131,7 @@ foreach my $f (@gffs){
 	$familiesTotalCounts{$family}=0 if (not defined($familiesTotalCounts{$family}));
 	$familiesTotalCounts{$family}++;
 	$HoHspeciesRfamCount{$species}{$family}++;
-	print OUT "$g\n" if defined($whitelist{$family});
+	print OUT "$g\n" if (defined($whitelist{$family}));
     }
     close(GFF);
     close(OUT);
@@ -224,7 +232,7 @@ foreach my $family ( @family ) {
     
     #Family must be found in >10% of all species
[…]

### 2013-10-21 scripts/heatmaps.R

@@ -100,5 +100,30 @@ heatmap.2(as.matrix(allrna),trace="none",col=gray(lb:0/lb),
           )
 dev.off()
 
+allrna<-read.table("data/R/diverged.dat",header = T, sep = "\t",row.names=52)
+pdf(file="paper/figures/diverged.pdf",width=8, height=15)
+heatmap.2(as.matrix(allrna),trace="none",col=gray(lb:0/lb),
+          dendrogram = "none", scale = "none",
+          breaks=breaks,
+          sepcolor="black", Rowv=F, Colv=F, key = F,
+          lmat=lmat, lwid=lwid, lhei=lhei,
+          cexCol=1.1, cexRow=1.1
+          )
+dev.off()
+
+allrna<-read.table("data/R/unusual-conserved.dat",header = T, sep = "\t",row.names=52)
+pdf(file="paper/figures/unusual-conserved.pdf",width=8, height=15)
+heatmap.2(as.matrix(allrna),trace="none",col=gray(lb:0/lb),
+          dendrogram = "none", scale = "none",
+          breaks=breaks,
+          sepcolor="black", Rowv=F, Colv=F, key = F,
+          lmat=lmat, lwid=lwid, lhei=lhei,
+          cexCol=1.1, cexRow=1.1
+          )
+dev.off()
+
+
+
+
 
 

### 2013-10-22 paper/manuscript.tex
Making it look like an article, updated phylogenetic orderings.

@@
 outside of mammalia. Notable examples include Xist \cite{Duret:2006}
-and H19 \cite{Smits:2008}. Some of the main results of this group are
-discussed in the comparative analysis of avian genomes article \cite{}.
+and H19 \cite{Smits:2008}.  There is emerging evidence for the
+conservation of “mammalian” lncRNAs in other vertebrates
+\cite{Chodroff:2010,Ulitsky:2011}), however, like most lncRNAs, the
+function of these lncRNAs remains largely unknown. Here, we show the
+conservation of several well-characterised lncRNAs of known function
+in humans.
 
@@
 supporting a high degree of evolutionary conservation for the entire
-lncRNA. In particular the HOXA11-AS1, PCA3, RMST, Six3os1, SOX2OT and
+lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1, SOX2OT and
 ST7-OT3 lncRNAs have multiple, well conserved RNA-domains (See
 Figure~\ref{fig:5}).
+
+ The conservation of these ``human'' lncRNAs among birds
+suggests they may also be functional in birds but what these
+functions is not immediately obvious. For example, PART1 and
+PCA3 are both described as prostate-specific lncRNAs that play a role
+in the human androgen-receptor pathway
+\cite{Bussemakers:1999,Lin:2000,Ferreira:2012}. Birds lack a prostate
+but both males and females express the androgen receptor (AR or NR3C4) in gonadal
+and non- gonadal tissue
+\cite{Yoshimura:1993,Veney:2004,Fuxjager:2012,Leska:2012}. Thus, we
+postulate that PART1 and PCA3 also play a role in the
+androgen-receptor pathway in birds but whether the expression of these
+lncRNAs are tissue specific is unknown at present.
 
@@
 Our ability to find additional homologs for several RNA families that fill
-gaps in the abundance matrices (Supplemental Figure~5)
+gaps in the abundance matrices (Figure~\ref{fig:7})
 strongly suggests that that conspicuous absences, in particular of LUCA and
@@
 
-\subsection*{Y RNA clusters}
+%\subsection*{Y RNA clusters}
 
@@
 
-\subsection*{MicroRNA clusters}
+%\subsection*{MicroRNA clusters}
 
@@
 In this work we have summarised the main results after carefully
-annotating non-coding RNAs in genome sequences. We have identified
-previously unrecognised conservation of ncRNAs in avian genomes as
-well as some surprising ``losses'' of otherwise well conserved
-ncRNAs. We suspect many of these losses are due to a combination of
-limitations in the homology search tools that we use for annotation
-and the ability of ncRNAs to tolerate large amounts of sequence
-variation while remaining functional, rather than \emph{bona fide}
-gene loss. In some cases these losses could be due to missing data
-from the genome assemblies, but this unlikely to be the case for
-multiple independant assemblies.
+annotating non-coding RNAs in genome sequences using homology-based
+methods. The homology-based tools have distinct advantages over
+experimental-based approaches as not all RNAs are expressed in any
+particular tissue-type or developmental-stage, in fact some RNAs have
+extremely specific expression profiles \cite{Johnston:2003}.
+
+ We have identified previously unrecognised conservation of
+ncRNAs in avian genomes as well as some surprising ``losses'' of
+otherwise well conserved ncRNAs. We suspect many of these losses are
+due to a combination of limitations in the homology search tools that
+we use for annotation and the ability of ncRNAs to tolerate large
+amounts of sequence variation while remaining functional, rather than
+\emph{bona fide} gene loss. In some cases these losses could be due to
+missing data from the genome assemblies, but this unlikely to be the
+case for multiple independant assemblies.
 
@@
 the fact that the reported functions of some ncRNAs are
-mammal-specific, yet some are also found in bird genomes.
+mammal-specific, yet some are also found in bird genomes. CITE DAVE
+BURT ET AL AND THE COMPARATIVE ARTICLE?
 
@@
 
+The snoRNA and miRNA annotations were validated using RNA-seq data
+(\ppg{SOURCED FROM?}). Since not all RNAs are expressed at any given
+time we expect to verify a fraction of RNAs annotated using expression
+data. Never-the-less 242 of 691 (35\%) miRNAs validated as well as 328
+of 376 (87\%) snoRNAs of our homology-based annotations were validated
+using this data.
+
 \section*{Acknowledgements}
@@
 \begin{figure}[ht]
-  \includegraphics[width=0.45\textwidth]{figures/lncRNA.pdf}
+  \includegraphics[width=0.45\textwidth]{figures/unusual-conserved.pdf}
+  \includegraphics[width=0.45\textwidth]{figures/diverged.pdf}
   \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
   \caption[]{Heatmaps showing the prescence/abscence and approximate
-    copy number of {\bf lncRNA} families.}\label{fig:5}
+    genomic copy number of ``unusually, well conserved RNAs''
+    (particularly the lncRNAs) on the left and families that have been
+    identified as surprising RNA Losses, divergence or missing
+    data. In several cases functionally related families have also
+    been included, e.g. the RNA components of the major and minor
+    spliceosomes: U1, U2, U4, U5 and U6; and U11, U12, U4atac, U5 and
+    U6atac, respectively.}\label{fig:5}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics[width=0.75\textwidth]{figures/More-cand.pdf}
+  \caption[]{...}\label{fig:7}
 \end{figure}

### 2013-10-22 scripts/gffs2heatmaps.pl

@@ -53,6 +53,7 @@ if (-s "$rDir/allRNA.dat"){
 my @whiteList=qw(SNORD34
 DLEU2_1 DLEU2_2 DLEU2_3 DLEU2_4 DLEU2_5 DLEU2_6 HOXA11-AS1_1 HOXA11-AS1_2 HOXA11-AS1_3 HOXA11-AS1_4 HOXA11-AS1_5 HOXA11-AS1_6 mir-15_1 mir-15_2 mir-15_3 mir-16_1 mir-16_2 mir-16_3 NBR2 PCA3_1 PCA3_2 RMST_1 RMST_10 RMST_2 RMST_3 RMST_4 RMST_5 RMST_6 RMST_7 RMST_8 RMST_9 Six3os1_1 Six3os1_2 Six3os1_3 Six3os1_4 Six3os1_5 Six3os1_6 Six3os1_7 SNORD93 SOX2OT_exon1 SOX2OT_exon2 SOX2OT_exon3 SOX2OT_exon4 ST7-OT3_1 ST7-OT3_2 ST7-OT3_3 ST7-OT3_4
 Metazoa_SRP RNase_MRP RNaseP_nuc Telomerase-vert U1 U11 U12 U2 U4 U4atac U5 U6 U6atac Vault
+PART1_1 PART1_2 PART1_3
 ); 
 
 foreach my $wl (@whiteList){
@@ -290,14 +291,21 @@ print "System calls: creating [$rDir/snoRNA-human-yeast-correspondences.dat] & r
 #DAMN THIS IS NASTY!
[…]

### 2013-10-22 scripts/heatmaps.R

@@ -101,24 +101,25 @@ heatmap.2(as.matrix(allrna),trace="none",col=gray(lb:0/lb),
 dev.off()
 
 allrna<-read.table("data/R/diverged.dat",header = T, sep = "\t",row.names=52)
-pdf(file="paper/figures/diverged.pdf",width=8, height=15)
+pdf(file="paper/figures/diverged.pdf",width=25, height=30)
 heatmap.2(as.matrix(allrna),trace="none",col=gray(lb:0/lb),
           dendrogram = "none", scale = "none",
           breaks=breaks,
           sepcolor="black", Rowv=F, Colv=F, key = F,
           lmat=lmat, lwid=lwid, lhei=lhei,
-          cexCol=1.1, cexRow=1.1
+          cexCol=3.0, cexRow=3.0, margins=c(15, 5)
           )
 dev.off()
 
 allrna<-read.table("data/R/unusual-conserved.dat",header = T, sep = "\t",row.names=52)
-pdf(file="paper/figures/unusual-conserved.pdf",width=8, height=15)
+pdf(file="paper/figures/unusual-conserved.pdf",width=25, height=30)
+
 heatmap.2(as.matrix(allrna),trace="none",col=gray(lb:0/lb),
           dendrogram = "none", scale = "none",
           breaks=breaks,
           sepcolor="black", Rowv=F, Colv=F, key = F,
           lmat=lmat, lwid=lwid, lhei=lhei,
-          cexCol=1.1, cexRow=1.1
+          cexCol=3.0, cexRow=3.0, margins=c(15, 5)
           )
 dev.off()
 

### 2013-11-05 paper/manuscript.tex
Merged comments from Peter, Sarah, Sam & Mario.

@@
 \author{
-Paul P. Gardner\correspondingauthor$^{1,2}$
-\email{Paul P. Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
+Paul P.\ Gardner\correspondingauthor$^{1,2}$
+\email{Paul P.\ Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
 Jana Hertel$^5$
 \email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
-Sarah W. Burge$^3$
-\email{sb30@sanger.ac.uk},
+Sarah W.\ Burge$^3$
+\email{swb@ebi.ac.uk},
 Maria Ninova$^4$
@@
 \email{mario@bierdepot.bioinf.uni-leipzig.de},
-Tammy E. Steeves$^1$
+Tammy E.\ Steeves$^1$
 \email{tammy.steeves@canterbury.ac.nz},
@@
 and
-Peter Stadler\correspondingauthor$^5$
+Peter F.\ Stadler\correspondingauthor$^5$
 \email{Peter Stadler\correspondingauthor - studla@bioinf.uni-leipzig.de}
@@
 contrast, we show extensive conservation of classical ncRNAs (e.g.,
-tRNAs and rRNAs) and more recently discovered ncRNAs (e.g., snoRNAs
+tRNAs) and more recently discovered ncRNAs (e.g., snoRNAs
 and miRNAs) in birds, we also demonstrate apparent ``losses'' in
@@
 
-Non-coding RNAs (ncRNAs) are an important class of genes. However,
-they pose serious research challenges, particularly for the field of
-genomics. For example, they lack the strong statistical signals
-associated with protein coding genes, e.g. open reading frames, G+C
-content and codon-usage biases \cite{Rivas:2000}.
+Non-coding RNAs (ncRNAs) are an important class of genes, responsible
+for the regulation of many key cellular functions. The major RNA
+families include the classical, highly conserved RNAs, sometimes
+called ``molecular fossils'', such as the transfer RNAs, ribosomal
+RNAs, RNA components of RNase P and the signal recognition particle
+\cite{Jeffares:1998}. Other classes appear to have have evolved more
+recently, e.g. the small nucleolar RNAs (snoRNAs), microRNAs (miRNAs)
+and the long non-coding RNAs (lncRNAs) \cite{Hoeppner:2012}.
 
-Non-coding RNAs are classified into hierarchical groupings by the Rfam
-database. The basic units are ``families'' which are groups of
-homologous, alignable sequences; ``clans'' which are groups of
-un-alignable (or functionally distinct), homologous families; and
-``classes'' which are groups of clans and families with related
-biological functions e.g. spliceosomal RNAs, miRNAs and snoRNAs
-\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
-The major RNA families include the classical, highly conserved RNAs,
-sometimes called ``molecular fossils'', such as the transfer RNAs,
-ribosomal RNAs, RNA components of RNase P and the signal recognition
-particle \cite{Jeffares:1998}. Other classes appear to have have
-evolved more recently, e.g. the small nucleolar RNAs (snoRNAs),
-microRNAs (miRNAs) and the long non-coding RNAs (lncRNAs)
-\cite{Hoeppner:2012}. The publication of 48 avian genomes, including
-the previously published chicken
-\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
-finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the newly
-published 45 avian genomes \cite{}, provides exciting opportunities to
-to explore ncRNA conservation in unprecedented detail.
-
-Despite promising techniques such as RNA-seq \cite{Croucher:2010},
-homology based methods namely covariance models (CMs) remain state of the
-art for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
+The ncRNAs pose serious research challenges, particularly for the
+field of genomics. For example, they lack the strong statistical
+signals associated with protein coding genes, e.g. open reading
+frames, G+C content and codon-usage biases \cite{Rivas:2000}. Despite
+promising techniques such as RNA-seq \cite{Croucher:2010}, homology
+based methods, namely covariance models (CMs) remain state of the art
+for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
 The CM based approach for annotating ncRNAs in genomes requires
@@
 generate sequences with similar properties, score the likelihood that
-a sequence is was generated by the same evolutionary processes as the
+a sequence is generated by the same evolutionary processes as the
 training sequences and to build alignments based upon sequence and
@@
 that the CM approaches dramatically out-perform alternative methods
-\cite{Freyhult:2007}.
+\cite{Freyhult:2007}, although its sensitivity is limited for the most
+rapidly evolving families such as vault RNAs or telomerase RNA
+\cite{Menzel:09a}.
 
@@
 the long non-coding RNAs (lncRNAs) are a particular challenge
-\cite{Guttman:2009}. Recent technological advances have led to dramatic
-speed and memory-usage enhancements for CM analyses
+\cite{Guttman:2009}. Recent technological advances have led to
+dramatic speed and memory-usage enhancements for CM analyses
 \cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
-cannot model the exon-intron structures of spliced lncRNAs, nor can they
-deal simply with the repeats that many lncRNAs host. Consequently in
-the latest release of Rfam the lncRNA families that were added were
-composed of local conserved and structured elements within lncRNAs,
-analogous to the ``domains'' housed within protein sequences
-\cite{Burge:2013}. The functions determined to date for lncRNAs
-range from regulating chromatin status to chromosomal inactivation
+cannot model the exon-intron structures of spliced lncRNAs, nor can
+they deal simply with the repeats that many lncRNAs host. Consequently
+in the latest release of Rfam the lncRNA families that were added were
+composed of local conserved (and possibly structured elements) within
+lncRNAs, analogous to the ``domains'' housed within protein sequences
+\cite{Burge:2013}. The functions determined to date for lncRNAs range
+from regulating chromatin status to chromosomal inactivation
 \cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
 genes is a lengthy and expensive process \cite{Guttman:2009}.
+
+The publication of 48 avian genomes, including the previously
+published chicken
+\cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
+finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the newly
+published 45 avian genomes \cite{birds:14,birds:14a}, provides
+exciting opportunities to to explore ncRNA conservation in
+unprecedented detail.
 
@@
 
-\subsection*{Unusually, well conserved RNAs}
+\subsection*{Unusually well conserved RNAs}
 
@@
 and H19 \cite{Smits:2008}.  There is emerging evidence for the
-conservation of “mammalian” lncRNAs in other vertebrates
+conservation of ``mammalian'' lncRNAs in other vertebrates
 \cite{Chodroff:2010,Ulitsky:2011}), however, like most lncRNAs, the
@@
 the birds. The annotations of these domains lie in the same genomic
-region, in the same order as in the mammalian homologs. Thus
-supporting a high degree of evolutionary conservation for the entire
+region, in the same order as in the mammalian homologs. Thus they
+support a high degree of evolutionary conservation for the entire
 lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1, SOX2OT and
@@
 
- The conservation of these ``human'' lncRNAs among birds
-suggests they may also be functional in birds but what these
+The conservation of these ``human'' lncRNAs among birds suggests they may 
+also be functional in birds but what these
 functions is not immediately obvious. For example, PART1 and
@@
 
-The HOX cluster lncRNAs, HOTAIRM1, HOTTIP and HOXA11-AS1, are
-remarkably well conserved. In the human genome the corresponding
-RNA-domains are located at chr7:27135743-27245922 (hg19) and are
-modelled by 15 RNA-domains in total, the 
-lncRNAs are ordered HOTAIRM1 (5 RNA-domains) $\rightarrow$ HOXA11-AS1
-(6 RNA-domains) $\rightarrow$ HOTTIP (4 RNA-domains). In the alligator
-and turtle genomes five and six of the HOX RNA-domains are conserved,
-each supporting an ancestral gene order of HOTAIRM1 $\rightarrow$
-HOXA11-AS1 $\rightarrow$ HOTTIP. In the birds, where two or more of
-the HOX cluster lncRNA RNA-domains were predicted on the same
-scaffold, this gene order was also preserved.
+The HOX cluster lncRNAs HOTAIRM1 (5 RNA-domains), HOXA11-AS1 (6
+RNA-domains), and HOTTIP (4 RNA domains) are remarkably well conserved. In
+the human genome they are located in the HOXA cluster (hg coordinates
+chr7:27135743-27245922), one of the most highly conserved regions in
+vertebrate genomes \cite{PascualAnaya:13}, in antisense orientation between
+HoxA1 and HoxA2, between HoxA11 and HoxA13, and upstream of HoxA13,
+respectively. Conservation and expression of HOTAIRM1 and HOXA11-AS1 within
+the HOXA cluster has been studied in some detail in marsupials
+\cite{Yu:12}.  Of the 15 RNA-domains five and six representing all three
+lncRNAs were recovered in the alligator and turtle genomes. All of them
+appear in the correct order at the expected, syntenically conserved
+positions within the HOXA cluster.  In the birds, where two or more of the
+HOX cluster lncRNA RNA-domains were predicted on the same scaffold, this
+gene order and location within HOX was also preserved.
+
 %egrep 'HOXA11-AS1|HOTAIR|HOTTIP' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
@@
 
-The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned for
-specificity and thus misses several occurrances that are easily found by
-\texttt{blastn} with $E\le 10^{-30}$. In some cases the sequences appear
-degraded at the ends, which may be explained e.g.\ by low sequence quality
-at the very ends of contigs or scaffolds. A \texttt{blastn} search also
-readily retrieves additional RNAse P and RNAse MRP RNAs, capturing only the
-best conserved regions. In many cases these additional candidates are
-incomplete or contain undetermined sequence, explaining why they are missed
-by the CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs
-in the majority of the genomes. An additional candidate could also be
-retrieved for telomerase RNA. Telomerase is well known to exhibit very poor
-sequence conservation and rapid variations in size that make it notoriously
-hard to identify by homology search \cite{Xie:08a}. The poor return thus
-does not come as a surprise. Since \texttt{blastn} searches remained
-unsuccessfull we constructed a sauropsid-specific CM for the vault RNA. In
-addition to the hits identified by the Rfam model we obtained three
-additional homologs. Vault RNAs, with a size of about 100 nt, exhibit
-conserved sequence patterns only at their ends, with essentially
-unconstrained sequence in the central part. Their identification is also a
-well-know difficult problem for homology search
-\cite{Stadler:09b}.
+The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned
+for specificity and thus misses several occurrances that are easily
+found by \texttt{blastn} with $E\le 10^{-30}$. In some cases the
+sequences appear degraded at the ends, which may be explained e.g.\ by
+low sequence quality at the very ends of contigs or scaffolds. A
+\texttt{blastn} search also readily retrieves additional RNAse P and
+RNAse MRP RNAs, capturing only the best conserved regions. In many
+cases these additional candidates are incomplete or contain
+undetermined sequence, explaining why they are missed by the
+CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
+the majority of the genomes. An additional candidate could also be
+retrieved for telomerase RNA. Telomerase is well known to exhibit very
+poor sequence conservation and rapid variations in size that make it
+notoriously hard to identify by homology search \cite{Xie:08a}. The
+poor return thus does not come as a surprise. Since \texttt{blastn}
+searches remained unsuccessfull we constructed a sauropsid-specific CM
+for the vault RNA. In addition to the hits identified by the Rfam
+model we obtained three additional homologs. Vault RNAs, with a size
+of about 100 nt, exhibit conserved sequence patterns only at their
+ends, with essentially unconstrained sequence in the central
+part. Their identification is one of the well-know difficult problems
+for homology search \cite{Stadler:09b}.
 
@@
 gaps in the abundance matrices (Figure~\ref{fig:7})
-strongly suggests that that conspicuous absences, in particular of LUCA and
+strongly suggests that conspicuous absences, in particular of LUCA and
 LECA RNAs, are caused by incomplete data in the current assemblies and
@@
 
-The microRNA family let-7 is the most abundant microRNA family with 14
+The microRNA family let-7 is the most diverse microRNA family with 14
 paralogs in human. These genes also localize in 7 genomic clusters,
@@
 evolution of the let-7 miRNA cluster in \cite{Hertel:2012}). In Sauropsids
-we observed that cluster A - that is strongly conserved in vertebrates has
+we observed that cluster A - which is strongly conserved in vertebrates has
 been completely lost in the avian lineage. Another obvious loss in birds is
@@
 
-In this work we have summarised the main results after carefully
-annotating non-coding RNAs in genome sequences using homology-based
-methods. The homology-based tools have distinct advantages over
-experimental-based approaches as not all RNAs are expressed in any
-particular tissue-type or developmental-stage, in fact some RNAs have
-extremely specific expression profiles \cite{Johnston:2003}.
-
- We have identified previously unrecognised conservation of
-ncRNAs in avian genomes as well as some surprising ``losses'' of
-otherwise well conserved ncRNAs. We suspect many of these losses are
-due to a combination of limitations in the homology search tools that
-we use for annotation and the ability of ncRNAs to tolerate large
-amounts of sequence variation while remaining functional, rather than
-\emph{bona fide} gene loss. In some cases these losses could be due to
-missing data from the genome assemblies, but this unlikely to be the
-case for multiple independant assemblies.
+In this work we have provided a comprehensive annotation of non-coding
+RNAs in genome sequences using homology-based methods. The
+homology-based tools have distinct advantages over experimental-based
+approaches as not all RNAs are expressed in any particular tissue-type
+or developmental-stage, in fact some RNAs have extremely specific
+expression profiles \cite{Johnston:2003}.  We have identified
+previously unrecognised conservation of ncRNAs in avian genomes as
+well as some surprising ``losses'' of otherwise well conserved
+ncRNAs. We suspect many of these losses are due to a combination of
+limitations in the homology search tools that we use for annotation
+and the ability of ncRNAs to tolerate large amounts of sequence
+variation while remaining functional, rather than \emph{bona fide}
+gene loss. In some cases these losses could be due to missing data
+from the genome assemblies, but this unlikely to be the case for
+multiple independant assemblies.
 
@@
 model-specific GA threshold and had an E-value smaller than 0.0005
-were retained.
+were retained. The Rfam database classifies non-coding RNAs into
+hierarchial groupings. The basic units are ``families'' which are
+groups of homologous, alignable sequences; ``clans'' which are groups
+of un-alignable (or functionally distinct), homologous families; and
+``classes'' which are groups of clans and families with related
+biological functions e.g. spliceosomal RNAs, miRNAs and snoRNAs
+\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013};
+these categories have been used to classify our results.
 
@@
 
-SnoRNA annotations overlapping miRNAs annotations where manually
-inspected \ppg{and?...}.
+%SnoRNA annotations overlapping miRNAs annotations where manually
+%inspected \ppg{and?...}.
 
-The snoRNA and miRNA annotations were validated using RNA-seq data
-(\ppg{SOURCED FROM?}). Since not all RNAs are expressed at any given
-time we expect to verify a fraction of RNAs annotated using expression
-data. Never-the-less 242 of 691 (35\%) miRNAs validated as well as 328
-of 376 (87\%) snoRNAs of our homology-based annotations were validated
-using this data.
+Chicken snoRNA and miRNA annotations were validated using small
+RNA-seq data comprising 27 samples from 14 diferent tissues. Since not
+all RNAs are expressed at any given time we expect to verify only a
+fraction of annotated RNAs using the expression data. Nevertheless 242
+of 691 (35\%) miRNAs as well as 328 of 376 (87\%) snoRNAs of our
+homology-based annotations were found to be supported by the RNA-seq
+data.
 
@@
 
-Erich Jarvis, Guojie Zhang and Tom Gilbert. Magnus Alm Rosenblad. 
+Erich Jarvis (Duke University), Guojie Zhang (BGI-Shenzhen \&
+University of Copenhagen) and Tom Gilbert (University of Copenhagen)
+for access to data and for invaluable feedback on the manuscript.
+
+Magnus Alm Rosenblad (Univ. of Gothenburg) for useful discussions.
+
+We thank Fiona McCarthy (Mississippi State University) and Carl
+Schmidt (University of Delaware) for providing the RNA-seq data.
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
   \includegraphics[width=0.75\textwidth]{figures/More-cand.pdf}
-  \caption[]{...}\label{fig:7}
+  \caption[]{Additional homologs of some sparsely represented RNA
+    families were discovered using dedicated search strategies
+    combined with highly sensitive settings, synteny information,
+    lineage-specific CMs and subsequent manual
+    inspection.}\label{fig:7}
 \end{figure}

### 2013-11-05 paper/supplementary-material.tex
Updated manuscript, scripts, fixed chicken seqIds, ...
Making it look like an article, updated phylogenetic orderings.
Merged comments from Peter, Sarah, Sam & Mario.

@@
 
-Nearly all the microRNAs that are broadly conserved in fish,
-amphibians and mammals are also conserved in the birds. Nevertheless,
-there are obvious instances of microRNAs lost in all birds. For
-example, mammalian, amphibian and fish genomes contain three loci of
-clustered microRNAs from the mir-17 and mir-92 families. One of these
-clusters (mir-93~25) was not found in turtles, crocodiles and
-birds. In addition, we can confidently identify a further 3 microRNA
+In addition, we can confidently identify a further 3 microRNA
 families that are present in mammals, and turtle and/or crocodile, but
@@
 Bacterial families can be used to identify problematic sequences that
-are likely to be the result of contamination from non-avian sources.
+are likely to be the result of contamination from non-avian
+sources. We identified a number of RNA families of bacterial origin in
+the avian genomes. These have been reported and will be dealt with in
+later updates to the avian genome sequences.
 
@@
 \begin{figure}[ht]
-  \includegraphics[width=0.45\textwidth]{figures/tRNA.pdf}
+  \includegraphics[width=0.85\textwidth]{figures/tRNA.pdf}
   \caption[]{Heatmaps showing the prescence/abscence and approximate
@@
 \begin{figure}[ht]
-  \includegraphics[width=0.45\textwidth]{figures/Y.pdf}
+  \includegraphics[width=0.85\textwidth]{figures/Y.pdf}
   \caption[]{...}\label{fig:6}
@@
 \begin{figure}[ht]
-  \includegraphics[width=0.45\textwidth]{figures/More-cand.pdf}
-  \caption[]{...}\label{fig:7}
-\end{figure}
-
-\begin{figure}[ht]
-  \includegraphics[width=0.45\textwidth]{figures/mir-17-cluster.pdf}
+  \includegraphics[width=0.85\textwidth]{figures/mir-17-cluster.pdf}
   \caption[]{...}\label{fig:8}
@@
 \begin{figure}[ht]
-  \includegraphics[width=0.45\textwidth]{figures/let-7-cluster.pdf}
+  \includegraphics[width=0.85\textwidth]{figures/let-7-cluster.pdf}
   \caption[]{...}\label{fig:9}

### 2013-11-05 scripts/compete_clans.pl

@@ -45,6 +45,12 @@ else{
     print "WARNING: no Clan information! (which annotations should be merged?)\n"
 }
 
+#rethreshold some problematic families
+my %rethreshold=(
+    'RF00254'=>70, #mir-16
+    'RF00606'=>50 #SNORD93
+    );
+
 my $fh = IO::File->new();
 # sort features so they can be merged in coordinate
 # order
@@ -65,6 +71,10 @@ while($cur = <$fh>) {
 	$rfid="undef";	
 	if($cur[8]=~/rfam-acc=(RF\d+)/){
 	    $rfid=$1;
+	    if( defined($rethreshold{$rfid})  && ($cur[5] < $rethreshold{$rfid})  ){
+		printf STDERR "skipping [$gffs[0]] {[$cur[5]] > [$rethreshold{$rfid}]} [$cur]\n";
+		next;
+	    }
 	}
 	
 	if ($cur[0] ne $prev[0]){#not the same sequence

### 2013-11-05 scripts/gffs2heatmaps.pl

@@ -75,6 +75,7 @@ my %blacklist=(
     mraW=>1,    
     );
 
+
 print "Reading GFFs\n" if (defined($verbose));
 my (%HoHspeciesRfamCount, %familiesSpeciesCounts, %familiesTotalCounts, %statistics, @bitscores, %tRNA, %tRNAfamilies); 
 foreach my $f (@gffs){
@@ -97,7 +98,8 @@ foreach my $f (@gffs){
 	my $family; 
 	if($g[8]=~/rfam-id=(\S+);evalue/ or $g[8]=~/Alias=(\S+);Note/){#Rfam
 	    $family=$1;
-	    #push(@bitscores, $g[5]) if (defined($whitelist{$family}) && $g[5]=~/^\d+\.\d+$/); 
+	    
+#push(@bitscores, $g[5]) if (defined($whitelist{$family}) && $g[5]=~/^\d+\.\d+$/); 
 	}
 	elsif($g[8]=~/ID=(\S+)\_\d+/){#miRBase
 	    $family=$1;

### 2013-11-11 paper/manuscript.tex
Updated figures and supp. mat.

@@
 supplement these annotations with predictions from the tRNA annotation
-tool, tRNAscan-SE and microRNAs from miRBase.  Although we show a
-significant number of lncRNAs are surprisingly well conserved between
-birds and mammals including several intriguing cases where the
-reported mammalian lncRNA function is not conserved in birds.  In
-contrast, we show extensive conservation of classical ncRNAs (e.g.,
-tRNAs) and more recently discovered ncRNAs (e.g., snoRNAs
-and miRNAs) in birds, we also demonstrate apparent ``losses'' in
-several RNA families. These include the divergence of some classical
-ncRNAs and the loss of several snoRNAs and microRNAs.  These combined
-results illustrate the utility of applying homology based methods for
+tool, tRNAscan-SE and microRNAs from miRBase.  We show a significant
+number of lncRNAs are surprisingly well conserved between birds and
+mammals including several intriguing cases where the reported
+mammalian lncRNA function is not conserved in birds.  We also
+demonstrate extensive conservation of classical ncRNAs (e.g., tRNAs)
+and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in
+birds. Furthermore, we have discovered apparent ``losses'' in several
+RNA families, these include the divergence of some classical ncRNAs
+and the loss of several snoRNAs and microRNAs.  These combined results
+illustrate the utility of applying homology based methods for
 annotating vertebrate genomes and illustrate many complex evolutionary
@@
 conserved in avian species. The latter are apparent losses of RNA
-genes that were expected to be conserved; Usually, this hypothesis is
+genes that were expected to be conserved; usually, this hypothesis is
 based upon the conservation of these RNAs in other vertebrate species.
@@
 
-In addition, the lncRNA DLEU2, is well conserved conserved across the
+In addition, the lncRNA DLEU2 is well conserved across the
 vertebrates, it is a host gene for two miRNA genes, miR-15 and miR-16,

### 2013-11-11 paper/supplementary-material.tex
Updated figures and supp. mat.

@@
 \begin{figure}[ht]
-  \includegraphics[width=0.85\textwidth]{figures/Y.pdf}
+  \includegraphics[width=0.95\textwidth]{figures/Y.pdf}
   \caption[]{...}\label{fig:6}
@@
 \begin{figure}[ht]
-  \includegraphics[width=0.85\textwidth]{figures/mir-17-cluster.pdf}
-  \caption[]{...}\label{fig:8}
+  \includegraphics[width=0.95\textwidth]{figures/mir-17-cluster.pdf}
+  \caption[Presence/absence table of mir-17 miRNA clusters]{This
+    figure illustrates the distribution of all members of the mir-17
+    clusters across birds, compared to fish and mammals. Columns
+    correspond to the single miRNA sequences, ordered by cluster and
+    position within its cluster. Rows correspond to the species and
+    the cells contain the number of copies of the miRNA in the
+    respective species. Colors correspond to orthologous miRNA
+    families: mir-17 (orange), mir-19 (blue) and mir-92/25
+    (green).There are 3 clusters, I-1 and I-X, and cluster II.  While
+    the two copies of cluster I are completely conserved, cluster II
+    has been lost as a whole in birds.}\label{fig:8}
 \end{figure}
@@
 \begin{figure}[ht]
-  \includegraphics[width=0.85\textwidth]{figures/let-7-cluster.pdf}
-  \caption[]{...}\label{fig:9}
+  \includegraphics[width=0.95\textwidth]{figures/let-7-cluster.pdf}
+  \caption[Presence/absence table of let-7 miRNA clusters]{This figure
+    illustrates the distribution of the let-7 paralogs and clustered
+    miRNAs of the families mir-100 and mir-125. Columns correspond to
+    the single miRNA sequences, ordered by cluster and position within
+    its cluster. Rows correspond to the species and the cells contain
+    the number of copies of the miRNA in the respective
+    species. Cluster A, which is strongly conserved in vertebrates has
+    been completely lost in the avian lineage. Another obvious loss in
+    birds is cluster F. However, in tinamou we detected an unclustered
+    miRNA that best fits to F-let-7-1 in sequence and structure, while
+    we could not find an I-let-7 miRNA sequence in this
+    species.}\label{fig:9}
 \end{figure}

### 2013-11-12 paper/manuscript.tex
Minor tweaks to the manuscript. Added a link to repository.

@@
 \usepackage{xspace}
+\usepackage{hyperref}
 \usepackage[utf8]{inputenc} %unicode support
@@
 \email{Paul P.\ Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
-Jana Hertel$^5$
-\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
 Sarah W.\ Burge$^3$
@@
 \email{Maria.Ninova@postgrad.manchester.ac.uk},
+Jana Hertel$^5$
+\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
 Stephanie Kehr$^5$
@@
 the fact that the reported functions of some ncRNAs are
-mammal-specific, yet some are also found in bird genomes. CITE DAVE
-BURT ET AL AND THE COMPARATIVE ARTICLE?
+mammal-specific, yet are also found in bird genomes.
 
@@
 %inspected \ppg{and?...}.
+
+Our diverse sets of genome annotations were combined. We collapsed the
+overlapping annotations into a single annotation. We also generated
+heatmaps for different groups of ncRNA genes (see Figure~\ref{fig:5}
+and Sup. Figs. 1-3). All the scripts and annotations presented here
+are available from Github \cite{gitrepo}.
 

### 2014-03-11 scripts/makeTable.pl
Amending manuscript, explaining losses, adding more expression data, fixing bugs

@@ -1,0 +1,131 @@
+#!/usr/bin/perl 
+
+use warnings;
+use strict;
+
+my %rfam2type;
+if (-s "data/rfam2type.txt"){
+    open(W, "< data/rfam2type.txt");    
+    while(my $w=<W>){
+	chomp($w);
+	my @w = split(/\t/,$w);
+	$rfam2type{$w[1]}=$w[2];
+	if($w=~/Cis-reg/){
+	    $rfam2type{$w[1]}='Cis-regulatory element';
+	}
+	elsif($w=~/HACA-box/){
+	    $rfam2type{$w[1]}='H/ACA box snoRNA';
+	}
+	elsif($w=~/CD-box/){
+	    $rfam2type{$w[1]}='C/D box snoRNA';
+	}
+	elsif($w=~/scaRNA/){
+	    $rfam2type{$w[1]}='Small cajal body RNA';
+	}
+	elsif($w=~/rRNA/){
+	    $rfam2type{$w[1]}='Ribosomal RNA';
+	}
+	elsif($w=~/miRNA/){
+	    $rfam2type{$w[1]}='microRNA';
+	}
+	elsif($w=~/lncRNA/){
+	    $rfam2type{$w[1]}='Long non-coding RNA';
+	}
+	elsif($w=~/tRNA/){
+	    $rfam2type{$w[1]}='Transfer RNA';
+	}
+	elsif($w[1]=~/U1/ or $w[1]=~/U2/ or $w[1]=~/U4$/ or $w[1]=~/U5/ or $w[1]=~/U6$/){
+	    $rfam2type{$w[1]}='Major spliceosomal RNA';
+	}
+	elsif($w[1]=~/U11/ or $w[1]=~/U12/ or $w[1]=~/U4atac$/ or $w[1]=~/U6atac$/){
+	    $rfam2type{$w[1]}='Minor spliceosomal RNA';
+	}
+	elsif($w=~/RNase/){
+	    $rfam2type{$w[1]}='RNase P/MRP RNA';
+	}
+	elsif($w=~/SRP/){
+	    $rfam2type{$w[1]}='SRP RNA';
+	}
+	elsif($w=~/7SK/){
+	    $rfam2type{$w[1]}='7SK RNA';
+	}
+	elsif($w=~/Y_RNA/){
+	    $rfam2type{$w[1]}='Y RNA';
+	}
+	elsif($w=~/Telomerase/){
[…]

### 2014-03-11 paper/manuscript.tex
Amending manuscript, explaining losses, adding more expression data, fixing bugs

@@
 signals associated with protein coding genes, e.g. open reading
-frames, G+C content and codon-usage biases \cite{Rivas:2000}. Despite
+frames, G+C content and codon-usage biases \cite{Rivas:2000}.
+
+{\bf Say something about the promise of RNA-seq... but requires many
+  tissues, developmental stages, fractions(e.g. poly-A, small, ...)
+  and the FPR is unknown.}
+
+Despite
 promising techniques such as RNA-seq \cite{Croucher:2010}, homology
@@
 
+{\bf REDUCE?}
 The CM based approach works well for almost all classes of ncRNA, but
@@
 published 45 avian genomes \cite{birds:14,birds:14a}, provides
-exciting opportunities to to explore ncRNA conservation in
-unprecedented detail.
+exciting opportunities to explore ncRNA conservation in
+unprecedented detail. {\bf MORE ON BIRD GENOMES. ESP. MACRO \& MICRO
+  CHROMOSOMES.}
 
@@
 sequences is generally biased towards model organisms
-\cite{Gardner:2010,Hoeppner:2012}. We focus our report on the unusual
-results within the avian lineages. These are either unexpectedly
-well-conserved RNAs or unexpectedly poorly-conserved RNAs. The former
-are RNAs we would not have expected to be conserved between the birds
-and the organisms these genes were initially identified in; Usually,
-this hypothesis is based upon the function of the RNA which is not
-conserved in avian species. The latter are apparent losses of RNA
-genes that were expected to be conserved; usually, this hypothesis is
-based upon the conservation of these RNAs in other vertebrate species.
-We use three models of ``loss'' that explain the data: Firstly, these
-could be genuine cases of an ancestral gene-loss along the avian
-lineage. Secondly, this could be a case of ``divergence'' where a RNA
-gene has undergone significant sequence and structural alterations, so
-much so, that our homology search tools can no longer detect a
-relationship between vertebrate exemplars and the avian
-varieties. Thirdly, we consider the possibility that the available
-genome assemblies have independently failed to capture these genes.
+\cite{Gardner:2010,Hoeppner:2012}.  {\bf Using accurate homology
+  search tools and evolutionary constraints, we produce an accurate
+  set of genome annotations for 48 bird species for ncRNAs that are
+  conserved across the avian species. This conservative set of
+  annotations is expected to contain the core avian ncRNAs.}
+We focus our
+report on the unusual results within the avian lineages. These are
+either unexpectedly well-conserved RNAs or unexpectedly
+poorly-conserved RNAs. The former are RNAs we would not have expected
+to be conserved between the birds and the organisms these genes were
+initially identified in; Usually, this hypothesis is based upon the
+function of the RNA which is not conserved in avian species. The
+latter are apparent losses of RNA genes that were expected to be
+conserved; usually, this hypothesis is based upon the conservation of
+these RNAs in other vertebrate species.  {\bf REWRITE: We use three
+  models of ``loss'' that explain the data: Firstly, these could be
+  genuine cases of an ancestral gene-loss along the avian
+  lineage. Secondly, this could be a case of ``divergence'' where a
+  RNA gene has undergone significant sequence and structural
+  alterations, so much so, that our homology search tools can no
+  longer detect a relationship between vertebrate exemplars and the
+  avian varieties. Thirdly, we consider the possibility that the
+  available genome assemblies have independently failed to capture
+  these genes.}
 
@@
 
-In general, Rfam cannot include the entire length of any large, spliced
-RNAs. This is a limitation of the covariance-models used for the
-homology-searches Rfam runs \cite{Nawrocki:2009}. Consequently, only
-short, well-conserved regions with evolutionarily conserved secondary
-structures are included in Rfam. By analogy to protein-domains, we
-refer to these as RNA-domains \cite{Burge:2013}.
+In general, Rfam cannot include the entire length of any large,
+spliced RNAs. This is a limitation of the covariance-models used for
+the homology-searches Rfam runs \cite{Nawrocki:2009}. Consequently,
+only short, well-conserved regions with evolutionarily conserved
+secondary structures are included in Rfam. By analogy to
+protein-domains, we refer to these as RNA-domains
+\cite{Burge:2013}. {\bf SOME OF THESE MAY BE DNA ELEMENTS RATHER THAN
+  LNCRNA ELEMENTS!!!!}
 
@@
 ST7-OT3 lncRNAs have multiple, well conserved RNA-domains (See
-Figure~\ref{fig:5}).
+Figure~\ref{fig:1}).
 
@@
 yet its conservation across the vertebrates certainly implies a
-function (See Figure~\ref{fig:5}).
+function (See Figure~\ref{fig:1}).
 
@@
 conservation or expansion within the avian lineage. The only exception
-being the snoRNA SNORD93 which we show has 92 copies in the tinamou
+being the snoRNA SNORD93 which has 92 copies in the tinamou
 genome, whereas it only has 1-2 copies in all the other vertebrate
@@
 
-\subsection*{RNA Losses, divergence or missing data?}
-
+\subsection*{RNA losses, divergence or missing data?}
 Much of the number of apparent losses and reduction in genomic
@@
 copy-number of RNA genes. Some of the most dramatic examples are the
-transfer RNAs and pseudogenes which average $\approx 900$ and $\approx
-580$ copies in the human, turtle and alligator genomes, the average
-copies numbers of these drop to $\approx 280$ and $\approx 100$ copies
-in the avian genomes.
+transfer RNAs and pseudogenes which average $\sim900$ and $\sim580$
+copies in the human, turtle and alligator genomes, the average copies
+numbers of these drop to $\sim280$ and $\sim100$ copies in the avian
+genomes.
 
-The absence of several well-conserved ncRNA families from many or even most
-bird genomes is unlikely to represent true gene losses. This concerns in
-particular the telomerase RNA, the RNA components of RNase P and MRP, the
-minor spliceosomal snRNAs U4atac and U11, the selenocystein tRNA (tRNA-Sec)
-as well as the vault RNAs. In order to get an idea to what extent the
-absence of these RNAs from the \texttt{infernal-based} annotation is caused 
-by sequence divergence beyond the thresholds of the Rfam CMs and/or 
-missing or incomplete data, we complemented our analysis by dedicated
-searches for a few of these RNA groups.
+The absence of seven well-conserved ncRNA families from many or even
+most bird genomes is unlikely to represent true gene losses. This
+concerns in particular the telomerase RNA, the RNA components of RNase
+P and MRP, the minor spliceosomal snRNAs U4atac and U11, the
+selenocystein tRNA (tRNA-Sec) as well as the vault RNAs (See
+Figure~\ref{fig:1}).
+
+{\bf Microchromosome background...}
+
+We propose three possible models of ``loss'' to explain the data:
+Firstly, these could be genuine cases of an ancestral gene-loss along
+the avian lineage. Secondly, this could be a case of ``divergence''
+where a RNA gene has undergone significant sequence and structural
+alterations, so much so, that homology search tools no longer
+detect a relationship between vertebrate exemplars and the avian
+varieties. Thirdly, we consider the possibility that the available
+genome assemblies have independently failed to capture these genes.
+
+The third model of loss is the most likely explanation. These families
+range in conservation level, from being ubiquitous to cellular-life
+(RNase P RNA and tRNA-Sec), to present in the majority of eukaryotes
+(RNase MRP, U4atac and U11) to present in most Bilateria
+(Vault). Therefore, the loss or even diversification of these families
+in the avian lineage is unlikely.
+
+One defining feature of the avian genomes is the presence of
+``microchromosomes''. As mentioned above, these are often difficult to
+assemble. The most complete available avian genomes are the chicken
+\cite{International_Chicken_Genome_Sequencing_Consortium:2004} and the
+zebrafinch \cite{Warren:2010}. The ``lost ncRNAs'' (loncRNAs) were
+found to more prevalent in the more complete genomes 6/7 in chicken
+and 3/7 in the zebrafinch. We examined the distribution of these
+loncRNAs and discovered that only the chicken RNase MRP gene is found
+on a macrochromosome. A Fisher's exact test showed that the loncRNAs
+are significantly enriched within microchromosomes ($P<10^{16}$) in
+both the chicken and the zebrafinch genomes (combined $P<2\times
+10^{31}$).
+
+{\bf CUT?}
+
+In order to get an idea to what extent the
+absence of these RNAs from the \texttt{infernal-based} annotation is
+caused by sequence divergence beyond the thresholds of the Rfam CMs
+and/or missing or incomplete data, we complemented our analysis by
+dedicated searches for a few of these RNA groups.
 
@@
 low sequence quality at the very ends of contigs or scaffolds. A
-\texttt{blastn} search also readily retrieves additional RNAse P and
+\texttt{blastn} search also readily retrieves additional RNase P and
 RNAse MRP RNAs, capturing only the best conserved regions. In many
@@
 Our ability to find additional homologs for several RNA families that fill
-gaps in the abundance matrices (Figure~\ref{fig:7})
+gaps in the abundance matrices (Figure~\ref{fig:2})
 strongly suggests that conspicuous absences, in particular of LUCA and
@@
 after the split of Theria. See Supplemental Figure~7 for details.
+
+\section*{Pseudogenes}
+
+{\bf Briefly mention the reduction in number of pseudogenes. Pick a
+  few key human ones and compare with the birds. }
 
@@
 overlapping annotations into a single annotation. We also generated
-heatmaps for different groups of ncRNA genes (see Figure~\ref{fig:5}
+heatmaps for different groups of ncRNA genes (see Figure~\ref{fig:1}
 and Sup. Figs. 1-3). All the scripts and annotations presented here
@@
 
+Thanks to @ewanbirney for the following tweet: ``@ctitusbrown
+@BioMickWatson So ... missing orthologs to chicken often mean 'gene
+might be on the microchromosome'''.
+
+We thank the anonymous referees for providing invaluable suggestions
+that improved this work.
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-\begin{figure}[ht]
-  \includegraphics[width=0.45\textwidth]{figures/unusual-conserved.pdf}
-  \includegraphics[width=0.45\textwidth]{figures/diverged.pdf}
-  \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
-  \caption[]{Heatmaps showing the prescence/abscence and approximate
-    genomic copy number of ``unusually, well conserved RNAs''
-    (particularly the lncRNAs) on the left and families that have been
-    identified as surprising RNA Losses, divergence or missing
-    data. In several cases functionally related families have also
-    been included, e.g. the RNA components of the major and minor
-    spliceosomes: U1, U2, U4, U5 and U6; and U11, U12, U4atac, U5 and
-    U6atac, respectively.}\label{fig:5}
-\end{figure}
+\section*{Figures}
+  \subsection*{Figure 1 - Heatmaps}
 
-\begin{figure}[ht]
-  \includegraphics[width=0.75\textwidth]{figures/More-cand.pdf}
-  \caption[]{Additional homologs of some sparsely represented RNA
-    families were discovered using dedicated search strategies
-    combined with highly sensitive settings, synteny information,
-    lineage-specific CMs and subsequent manual
-    inspection.}\label{fig:7}
-\end{figure}
+Heatmaps showing the prescence/abscence and approximate genomic copy
+number of ``unusually, well conserved RNAs'' (particularly the
+lncRNAs) on the left and families that have been identified as
+surprising RNA Losses, divergence or missing data. In several cases
+functionally related families have also been included, e.g. the RNA
+components of the major and minor spliceosomes: U1, U2, U4, U5 and U6;
+and U11, U12, U4atac, U5 and U6atac, respectively.\label{fig:1}
+
+%% \begin{figure}[ht]
+%%   \includegraphics[width=0.45\textwidth]{figures/unusual-conserved.pdf}
+%%   \includegraphics[width=0.45\textwidth]{figures/diverged.pdf}
+%%   \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
+%%   \caption[]{Heatmaps showing the prescence/abscence and approximate
+%%     genomic copy number of ``unusually, well conserved RNAs''
+%%     (particularly the lncRNAs) on the left and families that have been
+%%     identified as surprising RNA Losses, divergence or missing
+%%     data. In several cases functionally related families have also
+%%     been included, e.g. the RNA components of the major and minor
+%%     spliceosomes: U1, U2, U4, U5 and U6; and U11, U12, U4atac, U5 and
+%%     U6atac, respectively.}\label{fig:1}
+%% \end{figure}
+
+  \subsection*{Figure 2 - Identifying missing RNAs}
+
+Additional homologs of some sparsely represented RNA families were
+discovered using dedicated search strategies combined with highly
+sensitive settings, synteny information, lineage-specific CMs and
+subsequent manual inspection.\label{fig:2}
+
+{\bf TABLE ``AVERAGE'' NUMBER OF ncRNAs IN A BIRD GENOME?}
+
+%% \begin{figure}[ht]
+%%   \includegraphics[width=0.75\textwidth]{figures/More-cand.pdf}
+%%   \caption[]{Additional homologs of some sparsely represented RNA
+%%     families were discovered using dedicated search strategies
+%%     combined with highly sensitive settings, synteny information,
+%%     lineage-specific CMs and subsequent manual
+%%     inspection.}\label{fig:2}
+%% \end{figure}
 

### 2014-03-11 paper/supplementary-material.tex
Amending manuscript, explaining losses, adding more expression data, fixing bugs

@@
   \includegraphics[width=0.95\textwidth]{figures/mir-17-cluster.pdf}
-  \caption[Presence/absence table of mir-17 miRNA clusters]{This
-    figure illustrates the distribution of all members of the mir-17
-    clusters across birds, compared to fish and mammals. Columns
-    correspond to the single miRNA sequences, ordered by cluster and
-    position within its cluster. Rows correspond to the species and
-    the cells contain the number of copies of the miRNA in the
-    respective species. Colors correspond to orthologous miRNA
-    families: mir-17 (orange), mir-19 (blue) and mir-92/25
-    (green).There are 3 clusters, I-1 and I-X, and cluster II.  While
-    the two copies of cluster I are completely conserved, cluster II
-    has been lost as a whole in birds.}\label{fig:8}
+  \caption[Presence/absence table of mir-17 miRNA clusters]{
+This figure illustrates the distribution of all members of the mir-17 clusters across birds, compared to fish and mammals.
+Columns correspond to single miRNA sequences, grouped by clusters and ordered by their position within the corresponding cluster.
+Rows correspond to the species and the cells contain the number of copies of the miRNA in the respective species.
+Colors correspond to orthologous miRNA families: mir-17 (orange), mir-19 (blue) and mir-92/25 (green).There are 3 clusters, I-1 and I-X, and cluster II.
+While the two copies of cluster I are completely conserved, cluster II has been lost as a whole in birds.}\label{fig:8}
 \end{figure}
@@
   \includegraphics[width=0.95\textwidth]{figures/let-7-cluster.pdf}
-  \caption[Presence/absence table of let-7 miRNA clusters]{This figure
-    illustrates the distribution of the let-7 paralogs and clustered
-    miRNAs of the families mir-100 and mir-125. Columns correspond to
-    the single miRNA sequences, ordered by cluster and position within
-    its cluster. Rows correspond to the species and the cells contain
-    the number of copies of the miRNA in the respective
-    species. Cluster A, which is strongly conserved in vertebrates has
-    been completely lost in the avian lineage. Another obvious loss in
-    birds is cluster F. However, in tinamou we detected an unclustered
-    miRNA that best fits to F-let-7-1 in sequence and structure, while
-    we could not find an I-let-7 miRNA sequence in this
-    species.}\label{fig:9}
+  \caption[Presence/absence table of let-7 miRNA clusters]{This figure illustrates the distribution of the let-7 paralogs and clustered miRNAs of the families mir-100 and mir-125.
+Columns correspond to single miRNA sequences, grouped by clusters and ordered by their position within the corresponding cluster.
+Rows correspond to the species and the cells contain the number of copies of the miRNA in the respective species.
+Cluster A, which is strongly conserved in vertebrates has been completely lost in the avian lineage.
+Another obvious loss in birds is cluster F.
+However, in tinamou we detected an unclustered miRNA that best fits to F-let-7-1 in sequence and structure, while we could not find an I-let-7 miRNA sequence in this species.}\label{fig:9}
 \end{figure}

### 2014-03-11 scripts/compete_clans.pl

@@ -69,7 +69,7 @@ while($cur = <$fh>) {
 	my @prev = split(/\t/, $prev);
 	next if (not defined $cur[3] || not defined $cur[4] || not defined $cur[6]); #not a GFF
 	$rfid="undef";	
-	if($cur[8]=~/rfam-acc=(RF\d+)/){
+	if($cur[8]=~/rfam-acc=(RF\d+)/ or $cur[8]=~/Name=(RF\d+);Alias/){
 	    $rfid=$1;
 	    if( defined($rethreshold{$rfid})  && ($cur[5] < $rethreshold{$rfid})  ){
 		printf STDERR "skipping [$gffs[0]] {[$cur[5]] > [$rethreshold{$rfid}]} [$cur]\n";
@@ -111,6 +111,16 @@ while($cur = <$fh>) {
 		$prev = $cur; 
 		print "#COMPETE miRBase >> RFAM (2) [$cnt] [$cur[8]]\n" if(defined $verbose);
 	    }
+	    elsif($sameStrand  && ($cur[1] eq "MFASTA") && ($prev[1] eq "Rfam")){#MERGE: Stadler snoRNAs & Rfam (1)
+		$cur = mergeGff(\@cur,\@prev); 
+		$cur = join("\t", @{$cur}); 
+		print "#COMPETE MFASTA >> RFAM (1) [$cnt] [$cur[8]]\n" if(defined $verbose);
+	    }
+	    elsif($sameStrand  && ($cur[1] eq "Rfam") && ($prev[1] eq "MFASTA")){#MERGE: Stadler snoRNAs & Rfam (2)
+		$cur = mergeGff(\@cur,\@prev); 
+		$cur = join("\t", @{$cur}); 
+		print "#COMPETE MFASTA >> RFAM (2) [$cnt] [$cur[8]]\n" if(defined $verbose);
+	    }
 	    elsif($sameStrand  && defined($prfid) && defined($clans{$rfid}) && defined($clans{$prfid}) && ($clans{$prfid} eq $clans{$rfid})){#COMPETE IF IN THE SAME CLAN AND ON THE SAME STRAND
 		my $curp  = compete(\@cur, \@prev);
 		$cur = join("\t", @{$curp}); 
@@ -124,7 +134,7 @@ while($cur = <$fh>) {
 	    else{#An unclassified overlap, print lowest scoring one to an overlaps file
 		
 		my $curp  = anticompete(\@cur, \@prev);
-		my $overlow = join("\t", @{$curp}); 
[…]

### 2014-03-11 scripts/gffs2heatmaps.pl

@@ -71,6 +71,17 @@ if (-s "data/rfam11_lncRNAs.txt"){
     close(W);
 }
 
+my %rfam2type;
+if (-s "data/rfam2type.txt"){
+    open(W, "< data/rfam2type.txt");    
+    while(my $w=<W>){
+	chomp($w);
+	$rfam2type{$w}=1;
+    }
+    close(W);
+}
+
+
 my %blacklist=(
     mraW=>1,    
     );
@@ -103,6 +114,9 @@ foreach my $f (@gffs){
 	}
 	elsif($g[8]=~/ID=(\S+)\_\d+/){#miRBase
 	    $family=$1;
+	}
+	elsif($g[1]=~/MFASTA/){#Stadler snoRNAs
+	    $family="$g[8]";
 	}
 	elsif($g[8]=~/type=Pseudo;/){#tRNAscan
 	    $family="tRNA-pseudogene";

### 2014-03-11 scripts/heatmaps.R

@@ -1,4 +1,4 @@
-#R CMD BATCH --no-save heatmaps.R
+#R CMD BATCH --no-save scripts/heatmaps.R
 
 library(gplots)
 breaks<-c(-1000,0,1,5,10,50,100,500,1000)
@@ -109,6 +109,15 @@ heatmap.2(as.matrix(allrna),trace="none",col=gray(lb:0/lb),
           lmat=lmat, lwid=lwid, lhei=lhei,
           cexCol=3.0, cexRow=3.0, margins=c(15, 5)
           )
+par(mar=c(21, 0, 0, 18) + 0.1, oma=c(0,0,0,0))
+#xvals<-c(8.9,10.6,  58.5,60.2,  65.4,67.1)
+xvals<-c(8.9,10.6,  65.4,67.1)
+for (x in xvals){
+    lines(c(x,x)/100, c(0,2), lwd=5)
+}
+box("plot", col="black",lwd=3)  
+
+
 dev.off()
 
 allrna<-read.table("data/R/unusual-conserved.dat",header = T, sep = "\t",row.names=52)

### 2014-03-14 paper/manuscript.tex
Added a summary table to the manuscript.

@@
 
-{\bf TABLE ``AVERAGE'' NUMBER OF ncRNAs IN A BIRD GENOME?}
+  \subsection*{Table 1 - A summary of ncRNA genes in human, chicken and all bird genomes}
+
+This table contains the total number of annotated ncRNAs from
+different RNA types in human, chicken and the median number for each
+of the 48 birds. The number of chicken ncRNA that show evidence for
+expression is also indicated (the percentage is given in parentheses).
+
+\clearpage
+\newpage
+
+\begin{tabular}{|r|r|r|r|l|}
+\hline 
+\multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\
+\hline 
+Number in human & median(48 birds) & Number in chicken & Number in chicken  & RNA type\\
+                &            &                   & confirmed with RNA-seq & \\
+\hline
+62&25.0&34&18 (52.9\%)&Long non-coding RNA\\ 
+356&499.5&427&323 (75.6\%)&microRNA\\ 
+281&120.0&106&91 (85.8\%)&C/D box snoRNA\\ 
+336&85.5&68&49 (72.1\%)&H/ACA box snoRNA\\ 
+34&13.0&12&12 (100.0\%)&Small cajal body RNA\\ 
+1754&48.5&71&37 (52.1\%)&Major spliceosomal RNA\\ 
+58&3.0&6&4 (66.7\%)&Minor spliceosomal RNA\\ 
+525&82.0&122&107 (87.7\%)&Cis-regulatory element\\ 
+316&6.5&9&5 (55.6\%)&7SK RNA\\ 
+1&0.0&2&0 (0.0\%)&Telomerase RNA\\ 
+9&0.0&2&1 (50.0\%)&Vault RNA\\ 
+892&3.0&3&2 (66.7\%)&Y RNA\\ 
+1084&173.5&300&291 (97.0\%)&Transfer RNA\\ 
+80&9.5&4&4 (100.0\%)&Transfer RNA pseudogene\\ 
+941&3.0&4&4 (100.0\%)&SRP RNA\\ 
+607&7.0&22&12 (54.5\%)&Ribosomal RNA\\ 
+4&1.0&2&2 (100.0\%)&RNase P/MRP RNA\\ 
+\hline
+\end{tabular}
 

### 2014-03-14 scripts/heatmaps.R

@@ -133,7 +133,11 @@ heatmap.2(as.matrix(allrna),trace="none",col=gray(lb:0/lb),
 dev.off()
 
 
-
+######################################################################
+rnaExp<-read.table("data/RNA-seq/summed-vals.dat",header = F, sep = "\t")
+pdf(file="paper/figures/expression-distribution.pdf",width=25, height=25)
+hist(log10(rnaExp$V3),breaks=1000)
+dev.off()
 
 
 

### 2014-03-14 scripts/makeTable.pl

@@ -25,7 +25,7 @@ if (-s "data/rfam2type.txt"){
 	elsif($w=~/rRNA/){
 	    $rfam2type{$w[1]}='Ribosomal RNA';
 	}
-	elsif($w=~/miRNA/){
+	elsif($w=~/miRNA/ or $w=~/mir-\d+/){
 	    $rfam2type{$w[1]}='microRNA';
 	}
 	elsif($w=~/lncRNA/){
@@ -58,13 +58,18 @@ if (-s "data/rfam2type.txt"){
 	elsif($w=~/Vault/){
 	    $rfam2type{$w[1]}='Vault RNA';
 	}
+#	else {
+#	    print "UNCLASSIFIED: [$w]\n";
+#	}
 	
     }
     close(W);
-    $rfam2type{'tRNA−pseudogene'}='Transfer RNA pseudogene';
+    $rfam2type{'tRNA-pseudogene'}='Transfer RNA pseudogene';
+    $rfam2type{'CD-snoRNA'}      ='C/D box snoRNA';
+    $rfam2type{'HACA-snoRNA'}    ='H/ACA box snoRNA';
 }
 
-my (@species, @families, %countsTot, %countsFams); 
+my (@species, @families, %countsTot, %countsSpecies, %countsArray, %countsFams, %expressedChicken); 
 if (-s "data/R/allRNA.dat"){
     open(W, "< data/R/allRNA.dat");
     while(my $w=<W>){
@@ -72,10 +77,14 @@ if (-s "data/R/allRNA.dat"){
 	my @w=split(/\t/, $w);
 	my $w=pop(@w);
 	
-	if($w=~/human/){
+	if($w=~/family/){
 	    @species=@w;
 	}
 	else{
+	    
+	    if(not defined($rfam2type{$w}) and $w=~/^mir-\d+$/){
+		$rfam2type{$w}='microRNA';
+	    }
 	    
 	    if(defined($rfam2type{$w})){
 		
@@ -85,17 +94,22 @@ if (-s "data/R/allRNA.dat"){
 		$countsTot{"chicken"}{$rfam2type{$w}}=0 if(not defined($countsTot{"chicken"}{$rfam2type{$w}}));
 		$countsTot{"chicken"}{$rfam2type{$w}}+=$w[7];
 		
+		$expressedChicken{$rfam2type{$w}}+=expressedInChicken($w); 
+		
+		#print "$w:\t$rfam2type{$w}\n";
[…]

### 2014-03-25 paper/manuscript.tex
Added figs to the manuscript.
Updated RNA-seq table for manuscript.

--- 2014-03-11 paper/manuscript.tex
@@
 \email{Paul P.\ Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
+Mario Fasold$^5$
+\email{mario@bierdepot.bioinf.uni-leipzig.de},
 Sarah W.\ Burge$^3$
@@
 \email{steffi@bierdepot.bioinf.uni-leipzig.de},
-Mario Fasold$^5$
-\email{mario@bierdepot.bioinf.uni-leipzig.de},
 Tammy E.\ Steeves$^1$
@@
 birds. Furthermore, we have discovered apparent ``losses'' in several
-RNA families, these include the divergence of some classical ncRNAs
-and the loss of several snoRNAs and microRNAs.  These combined results
-illustrate the utility of applying homology based methods for
-annotating vertebrate genomes and illustrate many complex evolutionary
-patterns within the avian ncRNA cohort.
+RNA families, these illustrate the complexity of bird genomes. In
+particular, issues with assembling microchromosomes using
+next-generation sequencing methods. These combined results illustrate
+the utility of applying homology based methods for annotating
+vertebrate genomes and illustrate many complex evolutionary patterns
+within the avian ncRNA cohort.
 \end{abstract}
@@
 
-{\bf Say something about the promise of RNA-seq... but requires many
-  tissues, developmental stages, fractions(e.g. poly-A, small, ...)
-  and the FPR is unknown.}
-
-Despite
-promising techniques such as RNA-seq \cite{Croucher:2010}, homology
-based methods, namely covariance models (CMs) remain state of the art
-for ncRNA analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
+{\bf needs work:} New sequencing technologies have dramatically
+expanded the rate at which ncRNAs are discovered and their functions
+are determined \cite{}. However, in order to determine the full range
+of ncRNAs across multiple species we would require multiple RNA
+fractions (e.g. long and short), in multiple species, in multiple
+developmental stages and tissues types. The costs of this approach are
+still prohibitive. Consequently, for this work we have concentrated on
+bioinformatic approaches. Primarily, homology based methods, namely
+covariance models (CMs). These remain state of the art for ncRNA
+analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
+established sensitivity and specificity rates \cite{Freyhult:2007}.
 The CM based approach for annotating ncRNAs in genomes requires
@@
 
-{\bf REDUCE?}
+{\bf REDUCE? too much talking about what it can't to compared to what it can do...}
 The CM based approach works well for almost all classes of ncRNA, but
@@
 \cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
-finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the newly
-published 45 avian genomes \cite{birds:14,birds:14a}, provides
+finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the recently
+published 45 avian genomes \cite{birds:14,birds:14a,Huang:2013,Zhan:2013,Shapiro:2013,Howard:2013,Li:2014}, provides
 exciting opportunities to explore ncRNA conservation in
-unprecedented detail. {\bf MORE ON BIRD GENOMES. ESP. MACRO \& MICRO
-  CHROMOSOMES.}
+unprecedented detail. 
+%{\bf MORE ON BIRD GENOMES. ESP. MACRO \& MICRO
+%  CHROMOSOMES.}
 
@@
 conserved; usually, this hypothesis is based upon the conservation of
-these RNAs in other vertebrate species.  {\bf REWRITE: We use three
-  models of ``loss'' that explain the data: Firstly, these could be
-  genuine cases of an ancestral gene-loss along the avian
-  lineage. Secondly, this could be a case of ``divergence'' where a
-  RNA gene has undergone significant sequence and structural
-  alterations, so much so, that our homology search tools can no
-  longer detect a relationship between vertebrate exemplars and the
-  avian varieties. Thirdly, we consider the possibility that the
-  available genome assemblies have independently failed to capture
-  these genes.}
+these RNAs in other vertebrate species.
+
+Here, we consider three categories of apparent loss: First, genuine
+gene losses in the avian lineage. Second, ``divergence'' where RNA genes
+have undergone such significant sequence and structural alternations
+that homology search tools can no longer detect a relationship between
+other vertebrate exemplars and avian varieties. Third, ``missing'' genes
+that failed to be captured in the available, largely
+fragmented, avian genomes. We postulate that the latter
+category is likely to be prevalent in comparative avian genome studies
+given the distinctive organisation of the avian genome. Namely, the
+avian karyotype is characterised by a large number of chromosomes
+(average $2n \approx 80$) generally consisting of a few larger
+``macrochromosomes'' and many smaller ``microchromosomes'' \cite{Griffin:2007}. This 'so many, so small' pattern presents significant
+assembly challenges \cite{Ellegren:2005}. Indeed, of the 48 published avian
+genomes, 20 of which are high-coverage ($>50X$), only two are
+chromosomally assembled (chicken and zebra finch; \cite{birds:14}).
+
+%% {\bf REWRITE: We use three
+%%   models of ``loss'' that explain the data: Firstly, these could be
+%%   genuine cases of an ancestral gene-loss along the avian
+%%   lineage. Secondly, this could be a case of ``divergence'' where a
+%%   RNA gene has undergone significant sequence and structural
+%%   alterations, so much so, that our homology search tools can no
+%%   longer detect a relationship between vertebrate exemplars and the
+%%   avian varieties. Thirdly, we consider the possibility that the
+%%   available genome assemblies have independently failed to capture
+%%   these genes.}
 
@@
 
+We thank Matt Schwatz (Harvard) and Igor Ulitsky (Weizmann Institute
+of Science) for access to the RNA-seq data.
+
 We thank the anonymous referees for providing invaluable suggestions
@@
 
-%% \begin{figure}[ht]
-%%   \includegraphics[width=0.45\textwidth]{figures/unusual-conserved.pdf}
-%%   \includegraphics[width=0.45\textwidth]{figures/diverged.pdf}
-%%   \includegraphics[width=0.1\textwidth]{figures/key2.pdf}
-%%   \caption[]{Heatmaps showing the prescence/abscence and approximate
-%%     genomic copy number of ``unusually, well conserved RNAs''
-%%     (particularly the lncRNAs) on the left and families that have been
-%%     identified as surprising RNA Losses, divergence or missing
-%%     data. In several cases functionally related families have also
-%%     been included, e.g. the RNA components of the major and minor
-%%     spliceosomes: U1, U2, U4, U5 and U6; and U11, U12, U4atac, U5 and
-%%     U6atac, respectively.}\label{fig:1}
-%% \end{figure}
+ \begin{figure}[ht]
+   \includegraphics[width=0.5\textwidth]{figures/figure1.pdf}
+ \end{figure}
 
@@
 
-{\bf TABLE ``AVERAGE'' NUMBER OF ncRNAs IN A BIRD GENOME?}
+ \begin{figure}[ht]
+   \includegraphics[width=0.5\textwidth]{figures/figure2.pdf}
+ \end{figure}
 
-%% \begin{figure}[ht]
-%%   \includegraphics[width=0.75\textwidth]{figures/More-cand.pdf}
-%%   \caption[]{Additional homologs of some sparsely represented RNA
-%%     families were discovered using dedicated search strategies
-%%     combined with highly sensitive settings, synteny information,
-%%     lineage-specific CMs and subsequent manual
-%%     inspection.}\label{fig:2}
-%% \end{figure}
+  \subsection*{Table 1 - A summary of ncRNA genes in human, chicken and all bird genomes}
+
+This table contains the total number of annotated ncRNAs from
+different RNA types in human, chicken and the median number for each
+of the 48 birds. The number of chicken ncRNA that show evidence for
+expression is also indicated (the percentage is given in parentheses).
+
+\clearpage
+\newpage
+
+False positive rate less the 10\%:
+
+\begin{tabular}{|r|r|r|r|l|}
+\hline 
+\multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\
+\hline 
+                &                  &                   & \multicolumn{1}{|l|}{Chicken ncRNAs } & \\
+                &                  &                   & \multicolumn{1}{|l|}{confirmed with RNA-seq} & \\
+\hline
+Number in human & median(48 birds) & Number in chicken & max(RNA$_i$)$>13.0$ & RNA type\\
+\hline
+62&25.0&34&12 (35.3\%) &Long non-coding RNA\\ 
+356&499.5&427&272 (63.7\%) &microRNA\\ 
+281&120.0&106&88 (83.0\%) &C/D box snoRNA\\ 
+336&85.5&68&46 (67.6\%) &H/ACA box snoRNA\\ 
+34&13.0&12&11 (91.7\%) &Small cajal body RNA\\ 
+1754&48.5&71&29 (40.8\%) &Major spliceosomal RNA\\ 
+58&3.0&6&2 (33.3\%) &Minor spliceosomal RNA\\ 
+525&82.0&122&79 (64.8\%) &Cis-regulatory element\\ 
+316&6.5&9&3 (33.3\%) &7SK RNA\\ 
+1&0.0&2&0 (0.0\%) &Telomerase RNA\\ 
+9&0.0&2&1 (50.0\%) &Vault RNA\\ 
+892&3.0&3&2 (66.7\%) &Y RNA\\ 
+1084&173.5&300&256 (85.3\%) &Transfer RNA\\ 
+80&9.5&4&2 (50.0\%) &Transfer RNA pseudogene\\ 
+941&3.0&4&2 (50.0\%) &SRP RNA\\ 
+607&7.0&22&10 (45.5\%) &Ribosomal RNA\\ 
+4&1.0&2&2 (100.0\%) &RNase P/MRP RNA\\ 
+\hline
+7340&1080.0&1194&817 (68.4\%) &Total\\ 
+\hline
+\end{tabular}
 

### 2014-03-25 scripts/compete_clans.pl

@@ -58,7 +58,7 @@ $fh->open( "sortGffs.pl $gffs | uniq | " );
 print "#running [sortGffs.pl $gffs |]\n" if(defined $verbose);
 #$gffs = join('_', @gffs);
 open(OV, "> $overlapDir/$$\_unclassified_overlaps_file.overlaps");
-print OV "#GFFS: [$gffs]\n"; 
+print OV "#GFFS: [$gffs]\n";
 my ($cur, $prev, $ext, $rfid, $prfid, %uniq);
 my $cnt=0;
 while($cur = <$fh>) {

### 2014-03-25 scripts/heatmaps.R

@@ -134,10 +134,38 @@ dev.off()
 
 
 ######################################################################
-rnaExp<-read.table("data/RNA-seq/summed-vals.dat",header = F, sep = "\t")
-pdf(file="paper/figures/expression-distribution.pdf",width=25, height=25)
-hist(log10(rnaExp$V3),breaks=1000)
+rnaExp   <-read.table("data/RNA-seq/summed-vals.dat",header = F, sep = "\t")
+rnaExpIndMc<-read.csv2("data/RNA-seq/mccarthy_expression_tissue.dat")
+rnaExpIndUl<-read.csv2("data/RNA-seq/ulitsky_expression_tissue.dat")
+
+len<-0;
+c<-0; 
+tf<-(rnaExpIndMc[,2]>2)
+thresh<-0.25
+for (i in 2:16) {
+    c<-c(c,as.numeric(rnaExpIndMc[,i]))
+    len<-len+length(rnaExpIndMc[rnaExpIndMc[,i]>thresh,i])
+    tf<-(tf | rnaExpIndMc[,i]>thresh)
+}
+for (i in 2:7 ) {
+    c<-c(c,as.numeric(rnaExpIndUl[,i]))
+    tf<-(tf | rnaExpIndMc[,i]>thresh)
+}
+length(rnaExpIndMc[tf,1])/length(rnaExpIndMc[,1])
+length(rnaExp$V3[rnaExp$V3>10])/length(rnaExp$V3)
+
+breaks<- seq(0, max(log10(rnaExp$V3+1)), length.out=200)
+
+
+pdf(file="paper/figures/expression-distribution.pdf",width=5, height=5)
+par(mfcol=c(2,1))
+hist(log10(rnaExp$V3+1),breaks=breaks,xaxt = "n",xlab="log10(sum(RNA)+1)", main="Summed RNA over all tissues")
+axis(1, 0:6, 10^(0:6))
+hist(log10(c+1),        breaks=breaks,xaxt = "n",xlab="log10(RNA+1)",      main="RNA over all tissues")
+axis(1, 0:6, 10^(0:6))
 dev.off()
+
+length(rnaExp$V3[rnaExp$V3>10])
 
 
 

### 2014-03-25 scripts/makeTable.pl

@@ -69,7 +69,7 @@ if (-s "data/rfam2type.txt"){
     $rfam2type{'HACA-snoRNA'}    ='H/ACA box snoRNA';
 }
 
-my (@species, @families, %countsTot, %countsSpecies, %countsArray, %countsFams, %expressedChicken); 
+my (@species, @families, %countsTot, %countsSpecies, %countsConditions, %countsFams, %expressedChickenPaul, %expressedChickenMario, %expressedChickenPaulRand, %expressedChickenMarioRand); 
 if (-s "data/R/allRNA.dat"){
     open(W, "< data/R/allRNA.dat");
     while(my $w=<W>){
@@ -94,7 +94,18 @@ if (-s "data/R/allRNA.dat"){
 		$countsTot{"chicken"}{$rfam2type{$w}}=0 if(not defined($countsTot{"chicken"}{$rfam2type{$w}}));
 		$countsTot{"chicken"}{$rfam2type{$w}}+=$w[7];
 		
-		$expressedChicken{$rfam2type{$w}}+=expressedInChicken($w); 
+		if(not defined($expressedChickenPaul{$rfam2type{$w}})){
+		    $expressedChickenPaul{$rfam2type{$w}}=0;
+		    $expressedChickenMario{$rfam2type{$w}}=0;
+		    $expressedChickenPaulRand{$rfam2type{$w}}=0;
+		    $expressedChickenMarioRand{$rfam2type{$w}}=0;
+		}
+		my ($paul, $mario, $paulRand, $marioRand)=expressedInChicken($w); 
+		$expressedChickenPaul{$rfam2type{$w}}     +=$paul;
+		$expressedChickenMario{$rfam2type{$w}}+=$mario if (isNumeric($mario));
+		$expressedChickenPaulRand{$rfam2type{$w}}     +=$paulRand;
+		$expressedChickenMarioRand{$rfam2type{$w}}+=$marioRand if (isNumeric($marioRand));
+		#print "mario:[$mario] paul:[$paul]\n"; 
 		
 		#print "$w:\t$rfam2type{$w}\n";
 		$countsTot{"all-birds"}{$rfam2type{$w}}=0 if(not defined($countsTot{"all-birds"}{$rfam2type{$w}}));
@@ -138,48 +149,120 @@ print '
 \hline 
 \multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\\\
 \hline 
[…]

### 2014-04-07 paper/manuscript.tex
Updated manuscript and figures.

@@
 and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in
-birds. Furthermore, we have discovered apparent ``losses'' in several
-RNA families, these illustrate the complexity of bird genomes. In
-particular, issues with assembling microchromosomes using
+birds. Furthermore, we describe numerous ``losses'' of several RNA
+families, and attribute these to genuine loss, divergence or missing
+data.  In particular, issues with assembling microchromosomes using
 next-generation sequencing methods. These combined results illustrate
-the utility of applying homology based methods for annotating
+the utility of applying homology based methods for annotating novel
 vertebrate genomes and illustrate many complex evolutionary patterns
@@
   search tools and evolutionary constraints, we produce an accurate
-  set of genome annotations for 48 bird species for ncRNAs that are
-  conserved across the avian species. This conservative set of
-  annotations is expected to contain the core avian ncRNAs.}
-We focus our
-report on the unusual results within the avian lineages. These are
-either unexpectedly well-conserved RNAs or unexpectedly
-poorly-conserved RNAs. The former are RNAs we would not have expected
-to be conserved between the birds and the organisms these genes were
-initially identified in; Usually, this hypothesis is based upon the
-function of the RNA which is not conserved in avian species. The
-latter are apparent losses of RNA genes that were expected to be
-conserved; usually, this hypothesis is based upon the conservation of
-these RNAs in other vertebrate species.
-
-Here, we consider three categories of apparent loss: First, genuine
-gene losses in the avian lineage. Second, ``divergence'' where RNA genes
-have undergone such significant sequence and structural alternations
-that homology search tools can no longer detect a relationship between
-other vertebrate exemplars and avian varieties. Third, ``missing'' genes
-that failed to be captured in the available, largely
-fragmented, avian genomes. We postulate that the latter
-category is likely to be prevalent in comparative avian genome studies
-given the distinctive organisation of the avian genome. Namely, the
-avian karyotype is characterised by a large number of chromosomes
-(average $2n \approx 80$) generally consisting of a few larger
-``macrochromosomes'' and many smaller ``microchromosomes'' \cite{Griffin:2007}. This 'so many, so small' pattern presents significant
-assembly challenges \cite{Ellegren:2005}. Indeed, of the 48 published avian
-genomes, 20 of which are high-coverage ($>50X$), only two are
-chromosomally assembled (chicken and zebra finch; \cite{birds:14}).
-
-%% {\bf REWRITE: We use three
-%%   models of ``loss'' that explain the data: Firstly, these could be
-%%   genuine cases of an ancestral gene-loss along the avian
-%%   lineage. Secondly, this could be a case of ``divergence'' where a
-%%   RNA gene has undergone significant sequence and structural
-%%   alterations, so much so, that our homology search tools can no
-%%   longer detect a relationship between vertebrate exemplars and the
-%%   avian varieties. Thirdly, we consider the possibility that the
-%%   available genome assemblies have independently failed to capture
-%%   these genes.}
+  set of genome annotations for 48 predominantly non-model bird
+  species for ncRNAs that are conserved across the avian species. This
+  conservative set of annotations is expected to contain the core
+  avian ncRNAs.}  We focus our report on the unusual results within
+the avian lineages. These are either unexpectedly well-conserved
+ncRNAs or unexpectedly poorly-conserved ncRNAs. The former are ncRNAs
+that are not expected to be conserved between the birds and the other
+vertebrates, particularly those ncRNAs whose function is not conserved
+in birds.  The latter are apparent losses of ncRNAs expected to be
+conserved; Here, we consider three categories of such ``loss'': First,
+genuine gene losses in the avian lineage where ncRNAs well conserved
+in other vertebrates are completely absent in birds.  Second,
+``divergence'' where ncRNAs have undergone such significant sequence
+and structural alternations that homology search tools can no longer
+detect a relationship between other vertebrate exemplars and avian
+varieties.  Third, ``missing'' ncRNAs that failed to be captured in
+the available, largely fragmented, avian genomes.  We postulate that
+the latter category is likely to be prevalent in comparative avian
+genome studies given the distinctive organisation of the avian
+genome. Namely, the avian karyotype is characterised by a large number
+of chromosomes (average $2n \approx 80$) generally consisting of a few
+larger ``macrochromosomes'' and many smaller ``microchromosomes''
+\cite{Griffin:2007}. This ’so many, so small’ pattern presents
+significant assembly challenges \cite{Ellegren:2005}. Indeed, of the
+48 published avian genomes, 20 of which are high-coverage ($>50X$ ),
+only two are chromosomally assembled (chicken and zebra finch;
+\cite{birds:14,Warren:2010}).
 
@@
 
-\subsection*{RNA losses, divergence or missing data?}
-Much of the number of apparent losses and reduction in genomic
-sequence has been extensively discussed elsewhere
-\cite{Organ:2007}. Unsurprisingly, this reduction is reflected in the
-copy-number of RNA genes. Some of the most dramatic examples are the
-transfer RNAs and pseudogenes which average $\sim900$ and $\sim580$
-copies in the human, turtle and alligator genomes, the average copies
-numbers of these drop to $\sim280$ and $\sim100$ copies in the avian
-genomes.
+\subsection*{Unexpectedly poorly conserved  ncRNAs: genuine loss, divergence or missing  data?}
 
-The absence of seven well-conserved ncRNA families from many or even
-most bird genomes is unlikely to represent true gene losses. This
-concerns in particular the telomerase RNA, the RNA components of RNase
-P and MRP, the minor spliceosomal snRNAs U4atac and U11, the
-selenocystein tRNA (tRNA-Sec) as well as the vault RNAs (See
-Figure~\ref{fig:1}).
+\subsubsection*{Genuine loss}
 
-{\bf Microchromosome background...}
+The overall reduction in avian genomic size has been extensively
+discussed elsewhere \cite{Organ:2007}. Unsurprisingly, this reduction
+is reflected in the copy-number of ncRNA genes. Some of the most
+dramatic examples are the transfer RNAs and pseudogenes which average
+$\sim900$ and $\sim580$ copies in the human, turtle and alligator
+genomes, the average copies numbers of these drop to $\sim280$ and
+$\sim100$ copies in the avian genomes. In addition to reduction in
+copy-number, the complete absence of several otherwise ubiquitous
+vertebrate ncRNAs in the avian lineage are indicative of genuine gene
+loss.
 
-We propose three possible models of ``loss'' to explain the data:
-Firstly, these could be genuine cases of an ancestral gene-loss along
-the avian lineage. Secondly, this could be a case of ``divergence''
-where a RNA gene has undergone significant sequence and structural
-alterations, so much so, that homology search tools no longer
-detect a relationship between vertebrate exemplars and the avian
-varieties. Thirdly, we consider the possibility that the available
-genome assemblies have independently failed to capture these genes.
+Namely, mammalian and amphibian genomes contain three loci of
+clustered microRNAs from the mir-17 and mir-92 families
+\cite{Tanzer:04}. One of these clusters (cluster II, with families
+mir-106b, mir-93 and mir-25) was not found in turtles, crocodiles and
+birds (see Supplemental Figure~6). In addition, the microRNA family
+let-7 is the most diverse microRNA family with 14 paralogs in
+human. These genes also localize in 7 genomic clusters, together with
+mir-100 and mir-125 miRNA families (see previous study on the
+evolution of the let-7 miRNA cluster in \cite{Hertel:2012}). In Sauropsids we
+observed that cluster A - which is strongly conserved in vertebrates
+has been completely lost in the avian lineage.  Another obvious loss
+in birds is cluster F, containing two let-7 microRNA paralogs. Cluster
+H, on the other hand has been retained in all oviparous animals and
+completely lost later, after the split of Theria (see Supplemental
+Figure~7).
 
-The third model of loss is the most likely explanation. These families
-range in conservation level, from being ubiquitous to cellular-life
-(RNase P RNA and tRNA-Sec), to present in the majority of eukaryotes
-(RNase MRP, U4atac and U11) to present in most Bilateria
-(Vault). Therefore, the loss or even diversification of these families
-in the avian lineage is unlikely.
+\subsubsection*{Divergence}
 
-One defining feature of the avian genomes is the presence of
-``microchromosomes''. As mentioned above, these are often difficult to
-assemble. The most complete available avian genomes are the chicken
-\cite{International_Chicken_Genome_Sequencing_Consortium:2004} and the
-zebrafinch \cite{Warren:2010}. The ``lost ncRNAs'' (loncRNAs) were
-found to more prevalent in the more complete genomes 6/7 in chicken
-and 3/7 in the zebrafinch. We examined the distribution of these
-loncRNAs and discovered that only the chicken RNase MRP gene is found
-on a macrochromosome. A Fisher's exact test showed that the loncRNAs
-are significantly enriched within microchromosomes ($P<10^{16}$) in
-both the chicken and the zebrafinch genomes (combined $P<2\times
-10^{31}$).
-
-{\bf CUT?}
-
-In order to get an idea to what extent the
-absence of these RNAs from the \texttt{infernal-based} annotation is
-caused by sequence divergence beyond the thresholds of the Rfam CMs
-and/or missing or incomplete data, we complemented our analysis by
-dedicated searches for a few of these RNA groups.
-
-The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned
-for specificity and thus misses several occurrances that are easily
-found by \texttt{blastn} with $E\le 10^{-30}$. In some cases the
-sequences appear degraded at the ends, which may be explained e.g.\ by
-low sequence quality at the very ends of contigs or scaffolds. A
-\texttt{blastn} search also readily retrieves additional RNase P and
-RNAse MRP RNAs, capturing only the best conserved regions. In many
-cases these additional candidates are incomplete or contain
-undetermined sequence, explaining why they are missed by the
-CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
-the majority of the genomes. An additional candidate could also be
-retrieved for telomerase RNA. Telomerase is well known to exhibit very
-poor sequence conservation and rapid variations in size that make it
-notoriously hard to identify by homology search \cite{Xie:08a}. The
-poor return thus does not come as a surprise. Since \texttt{blastn}
-searches remained unsuccessfull we constructed a sauropsid-specific CM
-for the vault RNA. In addition to the hits identified by the Rfam
-model we obtained three additional homologs. Vault RNAs, with a size
-of about 100 nt, exhibit conserved sequence patterns only at their
-ends, with essentially unconstrained sequence in the central
-part. Their identification is one of the well-know difficult problems
-for homology search \cite{Stadler:09b}.
-
-Our ability to find additional homologs for several RNA families that fill
-gaps in the abundance matrices (Figure~\ref{fig:2})
-strongly suggests that conspicuous absences, in particular of LUCA and
-LECA RNAs, are caused by incomplete data in the current assemblies and
-sequence divergence rather then true losses.
-
-%\subsection*{Y RNA clusters}
+In order to determine to what extent the absence of these RNAs from
+the infernal-based annotation is caused by sequence divergence beyond
+the thresholds of the Rfam CMs, we complemented our analysis by
+dedicated searches for a few of these RNA groups. Our ability to find
+additional homologs for several RNA families that fill gaps in the
+abundance matrices (Figure ?) strongly suggests that conspicuous
+absences, in particular of LUCA and LECA RNAs, are caused by
+incomplete data in the current assemblies and sequence divergence
+rather then genuine losses.
 
@@
 bird genomes, while it is still present in both alligator and turtle,
-see Supplemental Figure~4. Within bird, we find an the conserved
-Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most cases
-consistent with breaks in the available sequence assemblies. In
+see Supplemental Figure~4. Within the avian lineage, we find a
+conserved Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most
+cases consistent with breaks in the available sequence assemblies.  In
 several genomes we observe one or a few additional Y RNA homologs
@@
 
-%\subsection*{MicroRNA clusters}
+\subsubsection*{Missing data}
 
-Nearly all microRNAs that are broadly conserved in fish, amphibians and
-mammals are also conserved in the birds. Nevertheless, there are obvious
-instances of microRNAs lost in all birds. For example, mammalian and
-ambhibian genomes contain three loci of clustered microRNAs from the mir-17
-and mir-92 families \cite{Tanzer:04}. One of these clusters (cluster II,
-with families mir-106b, mir-93 and mir-25) was not found in turtles,
-crocodiles and birds, see Supplemental Figure~6.
+Seven families of ncRNAs were found in some avian genomes but not
+others (Figure~\ref{fig:1}). These families range in conservation level
+from being ubiquitous to cellular-life (RNase P and tRNA-sec), present
+in most Bilateria (vault), present in the majority of eukaryotes
+(RNase MRP, U4atac and U11) and present in all vertebrates
+(telomerase) \cite{Hoeppner:2012}. Therefore, the genuine loss or even
+diversification of these ncRNA families in the avian lineage is
+unlikely. Rather, this lack of phylogenetic signal, combined with the
+fragmented nature of the vast majority of these genomes described
+above (i.e., of the 48 avian genomes, only the chicken and zebra finch
+are chromosomally assembled \cite{birds:14,Warren:2010}), suggests the most likely
+explanation is that these ncRNA families are indicative of missing
+data. Indeed, of the seven missing ncRNA families, six where found in
+the chicken genome and three were found in the zebra finch
+genome. Furthermore, only one of these (RNase MRP) is found on a
+macrochromosome, and all remaining missing ncRNAs are found on
+microchromosomes (see Supplemental Table ?). A Fisher’s exact test
+showed that there is significantly more missing ncRNAs on
+microchromosomes than macrochromosomes ($P<10^{16}$ for both the
+chicken and zebra finch). Thus, we suggest that many of
+these ncRNAs families are missing because: (1) they are predominantly
+found on microchromosomes [this study] and (2) the vast majority of
+avian microchromosomes remain unassembled \cite{birds:14,Ellegren:2005}.
 
-The microRNA family let-7 is the most diverse microRNA family with 14
-paralogs in human. These genes also localize in 7 genomic clusters,
-together with mir-100 and mir-125 miRNA families (see previous study on the
-evolution of the let-7 miRNA cluster in \cite{Hertel:2012}). In Sauropsids
-we observed that cluster A - which is strongly conserved in vertebrates has
-been completely lost in the avian lineage. Another obvious loss in birds is
-cluster F, containing two let-7 microRNA paralogs. Cluster H, on the other
-hand has been retained in all oviparous animals and completely lost later,
-after the split of Theria. See Supplemental Figure~7 for details.
+To wit, we performed dedicated searches for a selection of these
+missing ncRNA families. Here, tRNAscan is tuned for specificity and
+thus misses several occurrences of tRNA-sec that are easily found in
+the majority of genomes by \texttt{blastn} with $E\le 10^{-30}$. In some cases the
+sequences appear degraded at the ends, which is likely due to low
+sequence quality at the very ends of contigs or scaffolds. A \texttt{blastn}
+search also readily retrieves additional RNase P and RNAse MRP RNAs in
+the majority of genomes, albeit only the best conserved regions are
+captured. In many cases these additional candidates are incomplete or
+contain undetermined sequence, which explains why they are missed by
+the CMs \cite{Stadler:09b,Kolbe:2009}.
 
@@
   few key human ones and compare with the birds. }
+
+\section*{Experimentally confirmed ncRNAs}
+
+The ncRNAs presented here have been identified using homology models
+and are evolutionarily conserved in multiple avian species. In order
+to further validate these predictions we have used RNA-seq from
+multiple chicken tissues. After mapping chicken RNA-seq data to the
+chicken genome (see Methods for details), we identified a threshold
+for calling a gene as expressed by limiting our estimated
+false-positive rate to less than 10\%. This was estimated using
+randomly selected, un-annotated and presumably un-expressed regions of
+the genome. Some of which may be genuinely expressed, meaning the true
+FDR is potentially much lower than 10\%. The $<10\%$ FDR constraint,
+corresponded to identifying at least one experimental condition where
+13 RNA-seq reads or more were identified, that mapped to a gene of
+interest. Overall, the number of ncRNAs we have identified in this
+work that are expressed above background levels is 68.4\% (see
+Table~\ref{table:1}). This shows that 6.9 times more of our ncRNAs are
+expressed than expected by chance (Fisher's exact test:
+$P<10^{16}$). It is likely that this is an underestimate of the
+fraction of our annotations that are genuinely expressed, as only a
+fraction of the developmental stages and tissues essential for forming
+a chicken have been sampled with RNA-seq. Furthermore we now know that
+some ncRNAs are expressed in highly specific conditions \cite{} ({\bf
+  the worm miRNA in 1 cell paper}).
+
+The classes of RNAs where the majority of our annotations were
+confirmed includes microRNAs, snoRNAs, cis-regulatory elements, tRNAs,
+SRP RNA and RNase P/MRP RNA. The RNA-seq data did fail to detect a
+telomerase RNA transcript ({\bf ARE THESE ONLY EXPRESSED DURING
+  MEIOSIS?}). The fractions for the other classes were well above the
+10\% FDR threshold but admittedly were low for 7SK RNA, the minor
+spliceosomal RNAs and the lncRNAs. There are a number of possible
+explanations for this: the multiple copies of the 7SK RNA may be
+functionally redundant and can therefore compensate for one another;
+The minor spliceosome is active during ({\bf SOME OBSCURE CONDITION
+  HOPEFULLY???}); and the lncRNAs are generally expressed at low
+levels under specific conditions ({\bf cite RNACAPTURE and MATTICK'S
+  BRAIN ATLAS PAPER}).
+
+%totalExpressed[817] totalNegControlsExpressed[117] totalGenes[1194]
 
@@
 
+{\bf Rewrite:}
 Chicken snoRNA and miRNA annotations were validated using small
@@
 data.
+
+{\bf WHERE IS ALL THE GENOME SEQUENCE AND RNA-SEQ DATA AVAILABLE
+  FROM?}
 
@@
 
-Thanks to @ewanbirney for the following tweet: ``@ctitusbrown
-@BioMickWatson So ... missing orthologs to chicken often mean 'gene
-might be on the microchromosome'''.
+Thanks to @ewanbirney for the following timely tweet: ``So ... missing
+orthologs to chicken often mean 'gene might be on the
+microchromosome''.
 
-We thank Matt Schwatz (Harvard) and Igor Ulitsky (Weizmann Institute
+We thank Matt Schwartz (Harvard) and Igor Ulitsky (Weizmann Institute
 of Science) for access to the RNA-seq data.
@@
  \begin{figure}[ht]
-   \includegraphics[width=0.5\textwidth]{figures/figure1.pdf}
+   \centering
+   \includegraphics[width=0.95\textwidth]{figures/figure1.pdf}
  \end{figure}
 
-  \subsection*{Figure 2 - Identifying missing RNAs}
-
-Additional homologs of some sparsely represented RNA families were
-discovered using dedicated search strategies combined with highly
-sensitive settings, synteny information, lineage-specific CMs and
-subsequent manual inspection.\label{fig:2}
-
- \begin{figure}[ht]
-   \includegraphics[width=0.5\textwidth]{figures/figure2.pdf}
- \end{figure}
+\clearpage
+\newpage
 
@@
 This table contains the total number of annotated ncRNAs from
-different RNA types in human, chicken and the median number for each
-of the 48 birds. The number of chicken ncRNA that show evidence for
+different RNA types in human, the median number for each of the 48
+birds and chicken. The number of chicken ncRNA that show evidence for
 expression is also indicated (the percentage is given in parentheses).
-
-\clearpage
-\newpage
-
-False positive rate less the 10\%:
+The threshold for determining expression was selected based upon a
+false positive rate of less than 10\%.\label{table:1}
 
@@
 \hline 
-                &                  &                   & \multicolumn{1}{|l|}{Chicken ncRNAs } & \\
-                &                  &                   & \multicolumn{1}{|l|}{confirmed with RNA-seq} & \\
-\hline
-Number in human & median(48 birds) & Number in chicken & max(RNA$_i$)$>13.0$ & RNA type\\
+Number in human & median(48 birds) & Number in chicken & Chicken ncRNAs         & RNA type\\
+                &                  &                   & confirmed with RNA-seq & \\
 \hline

### 2014-04-07 paper/supplementary-material.tex
Updated manuscript and figures.

@@
 \email{Paul P. Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
+Mario Fasold$^5$
+\email{mario@bierdepot.bioinf.uni-leipzig.de},
+Sarah W. Burge$^3$
+\email{sb30@sanger.ac.uk},
 Jana Hertel$^5$
 \email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
-Sarah W. Burge$^3$
-\email{sb30@sanger.ac.uk},
 Maria Ninova$^4$
@@
 \email{steffi@bierdepot.bioinf.uni-leipzig.de},
-Mario Fasold$^5$
-\email{mario@bierdepot.bioinf.uni-leipzig.de},
 Tammy E. Steeves$^1$
@@
 
+\subsection*{Unexpectedly poorly conserved  ncRNAs: testing divergence}
+
+In order to get an idea to what extent the
+absence of these RNAs from the \texttt{infernal-based} annotation is
+caused by sequence divergence beyond the thresholds of the Rfam CMs
+and/or missing or incomplete data, we complemented our analysis by
+dedicated searches for a few of these RNA groups.
+
+The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned
+for specificity and thus misses several occurrances that are easily
+found by \texttt{blastn} with $E\le 10^{-30}$. In some cases the
+sequences appear degraded at the ends, which may be explained e.g.\ by
+low sequence quality at the very ends of contigs or scaffolds. A
+\texttt{blastn} search also readily retrieves additional RNase P and
+RNAse MRP RNAs, capturing only the best conserved regions. In many
+cases these additional candidates are incomplete or contain
+undetermined sequence, explaining why they are missed by the
+CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
+the majority of the genomes. An additional candidate could also be
+retrieved for telomerase RNA. Telomerase is well known to exhibit very
+poor sequence conservation and rapid variations in size that make it
+notoriously hard to identify by homology search \cite{Xie:08a}. The
+poor return thus does not come as a surprise. Since \texttt{blastn}
+searches remained unsuccessfull we constructed a sauropsid-specific CM
+for the vault RNA. In addition to the hits identified by the Rfam
+model we obtained three additional homologs. Vault RNAs, with a size
+of about 100 nt, exhibit conserved sequence patterns only at their
+ends, with essentially unconstrained sequence in the central
+part. Their identification is one of the well-know difficult problems
+for homology search \cite{Stadler:09b,Kolbe:2009}.
+
+Our ability to find additional homologs for several RNA families that fill
+gaps in the abundance matrices (Figure~\ref{fig:2})
+strongly suggests that conspicuous absences, in particular of LUCA and
+LECA RNAs, are caused by incomplete data in the current assemblies and
+sequence divergence rather then true losses.
+
+ \begin{figure}[ht]
+   \includegraphics[width=0.5\textwidth]{figures/figure2.pdf}
+\caption{Additional homologs of some sparsely represented RNA families were
+discovered using dedicated search strategies combined with highly
+sensitive settings, synteny information, lineage-specific CMs and
+subsequent manual inspection.}\label{fig:2}
+ \end{figure}
+
 \subsection*{Exceptional RNAs}

### 2014-04-07 scripts/makeTable.pl

@@ -183,6 +183,8 @@ print '\hline
 \end{tabular}
 ';
 
+print "\n\n\n\n totalExpressed[$totalExpressedChickenMario] totalNegControlsExpressed[$totalExpressedChickenMarioRand] totalGenes[$totalChicken]\n";
+
 # print '
 # \begin{tabular}{|r|r|r|r|r|l|}
 # \hline 

### 2014-04-11 paper/manuscript.tex
Working on the RNA-seq validation.

@@
 
-{\bf needs work:} New sequencing technologies have dramatically
-expanded the rate at which ncRNAs are discovered and their functions
-are determined \cite{}. However, in order to determine the full range
-of ncRNAs across multiple species we would require multiple RNA
+New sequencing technologies have dramatically expanded the rate at
+which ncRNAs are discovered and their functions are determined
+\cite{cech2014noncoding}. However, in order to determine the full
+range of ncRNAs across multiple species we would require multiple RNA
 fractions (e.g. long and short), in multiple species, in multiple
@@
 reliable alignments and consensus secondary structures of
-representative sequences of RNA families. These are used to train
-probabilistic models for each family. These models can be used to
-generate sequences with similar properties, score the likelihood that
-a sequence is generated by the same evolutionary processes as the
-training sequences and to build alignments based upon sequence and
-structural information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
-The tRNAscan-SE software package uses CMs to accurately predict
-transfer RNAs \cite{Lowe:1997,Chan:2009}.  The Rfam database contains
-thousands of curated alignments and consensus structures for diverse
-classes of ncRNAs
-\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
+representative sequences of RNA families, many of which can be found
+at Rfam
+\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}. These
+are used to train probabilistic models that score the likelihood that
+a database sequence is generated by the same evolutionary processes as
+the training sequences based upon both sequence and structural
+information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.  The
+tRNAscan-SE software package uses CMs to accurately predict transfer
+RNAs \cite{Lowe:1997,Chan:2009}.
+
 Independent benchmarks of bioinformatic annotation tools have shown
@@
 
-{\bf REDUCE? too much talking about what it can't to compared to what it can do...}
-The CM based approach works well for almost all classes of ncRNA, but
-the long non-coding RNAs (lncRNAs) are a particular challenge
+%%{\bf REDUCE? too much talking about what it can't to compared to what it can do...}
+The CM based approach works well for most classes of ncRNA, but the
+long non-coding RNAs (lncRNAs) are a particular challenge
 \cite{Guttman:2009}. Recent technological advances have led to
@@
 lncRNAs, analogous to the ``domains'' housed within protein sequences
-\cite{Burge:2013}. The functions determined to date for lncRNAs range
-from regulating chromatin status to chromosomal inactivation
-\cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
-genes is a lengthy and expensive process \cite{Guttman:2009}.
+\cite{Burge:2013}. Whilst some these regions may not reflect
+functional RNA elements instead regulatory regions, enhancers or
+insulators, their syntenic conservation still provides an indication
+of lncRNA conservation \cite{diederichs2014four}.
+
+%% The functions determined to date for lncRNAs range
+%% from regulating chromatin status to chromosomal inactivation
+%% \cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
+%% genes is a lengthy and expensive process \cite{Guttman:2009}.
 
@@
 and are evolutionarily conserved in multiple avian species. In order
-to further validate these predictions we have used RNA-seq from
-multiple chicken tissues. After mapping chicken RNA-seq data to the
-chicken genome (see Methods for details), we identified a threshold
-for calling a gene as expressed by limiting our estimated
-false-positive rate to less than 10\%. This was estimated using
-randomly selected, un-annotated and presumably un-expressed regions of
-the genome. Some of which may be genuinely expressed, meaning the true
-FDR is potentially much lower than 10\%. The $<10\%$ FDR constraint,
-corresponded to identifying at least one experimental condition where
-13 RNA-seq reads or more were identified, that mapped to a gene of
-interest. Overall, the number of ncRNAs we have identified in this
-work that are expressed above background levels is 68.4\% (see
-Table~\ref{table:1}). This shows that 6.9 times more of our ncRNAs are
-expressed than expected by chance (Fisher's exact test:
-$P<10^{16}$). It is likely that this is an underestimate of the
-fraction of our annotations that are genuinely expressed, as only a
-fraction of the developmental stages and tissues essential for forming
-a chicken have been sampled with RNA-seq. Furthermore we now know that
-some ncRNAs are expressed in highly specific conditions \cite{} ({\bf
-  the worm miRNA in 1 cell paper}).
+to further validate these predictions we have used strand-specific
+total RNA-seq and small RNA-seq of multiple chicken tissues. After
+mapping the RNA-seq data to the chicken genome (see Methods for
+details), we identified a threshold for calling a gene as
+expressed by limiting our estimated false-positive rate to less than
+10\%. This FDR was estimated using a negative control of randomly
+selected, un-annotated regions of the genome. Since some regions may
+be genuinely expressed, the true FDR is potentially much lower than
+10\%. Overall, the number of ncRNAs we
+have identified in this work that are expressed above background
+levels is 865 (72.4\%) (see Table~\ref{table:1}). This shows that 7.0
+times more of our ncRNAs are expressed than expected by chance
+(Fisher's exact test: $P<10^{16}$). This number is an underestimate of
+the fraction of our annotations that are genuinely expressed, as only
+a fraction of the developmental stages and tissues of chicken have
+been characterized with RNA-seq. Furthermore, some ncRNAs
+are expressed in highly specific conditions  \cite{mercer2008specific,johnston2003microrna}.
 
 The classes of RNAs where the majority of our annotations were
-confirmed includes microRNAs, snoRNAs, cis-regulatory elements, tRNAs,
-SRP RNA and RNase P/MRP RNA. The RNA-seq data did fail to detect a
-telomerase RNA transcript ({\bf ARE THESE ONLY EXPRESSED DURING
-  MEIOSIS?}). The fractions for the other classes were well above the
-10\% FDR threshold but admittedly were low for 7SK RNA, the minor
-spliceosomal RNAs and the lncRNAs. There are a number of possible
-explanations for this: the multiple copies of the 7SK RNA may be
-functionally redundant and can therefore compensate for one another;
-The minor spliceosome is active during ({\bf SOME OBSCURE CONDITION
-  HOPEFULLY???}); and the lncRNAs are generally expressed at low
-levels under specific conditions ({\bf cite RNACAPTURE and MATTICK'S
-  BRAIN ATLAS PAPER}).
+experimentally confirmed includes microRNAs, snoRNAs, cis-regulatory
+elements, tRNAs, SRP RNA and RNase P/MRP RNA. The RNA-seq data could
+not provide evidence for a telomerase RNA transcript, which are only
+generally only expressed in embryonic, stem or cancerous tissues. Only
+a small fraction of the 7SK RNA, the minor spliceosomal RNAs and the
+lncRNAs could be confirmed with the 10\% FDR threshold. There are a
+number of possible explanations for this: the multiple copies of the
+7SK RNA may be functionally redundant and can therefore compensate for
+one another; The minor spliceosome is, as the name suggests, a rarely
+used alternative spliceosome; and the lncRNAs are generally expressed
+at low levels under specific conditions
+\cite{mercer2008specific,mercer2012targeted}.
 
@@
 
-Our diverse sets of genome annotations were combined. We collapsed the
+The diverse sets of genome annotations were combined. We collapsed the
 overlapping annotations into a single annotation. We also generated
@@
 
-{\bf Rewrite:}
-Chicken snoRNA and miRNA annotations were validated using small
-RNA-seq data comprising 27 samples from 14 diferent tissues. Since not
-all RNAs are expressed at any given time we expect to verify only a
-fraction of annotated RNAs using the expression data. Nevertheless 242
-of 691 (35\%) miRNAs as well as 328 of 376 (87\%) snoRNAs of our
-homology-based annotations were found to be supported by the RNA-seq
-data.
-
-{\bf WHERE IS ALL THE GENOME SEQUENCE AND RNA-SEQ DATA AVAILABLE
-  FROM?}
+Chicken ncRNA predictions were validated using two separate RNA-seq
+data sets. The first data set (Bioproject PRJNA204941) contains 971
+million reads and comprises 27 samples from 14 diferent chicken
+tissues sequenced on Illumina HiSeq2000 using a small RNA-seq
+protocol. The second data set (personal commun. Matt Schwartz and Igor
+Ulitsky) contains 1,46 billion Illumina HiSeq reads sequenced from
+whole chicken embryo RNA from 7 stages using a strand-specific dUTP
+protocol. Raw reads were checked for quality and adapters clipped if
+required by the protocol. Preprocessed reads were mapped to the
+galGal4 reference genome using SEGEMEHL short read aligner
+\cite{segemehl} and then overlapped with the ncRNA annotations.
 
@@
 We thank Fiona McCarthy (Mississippi State University) and Carl
-Schmidt (University of Delaware) for providing the RNA-seq data.
+Schmidt (University of Delaware) as well as Matt Schwartz (Harvard)
+and Igor Ulitsky (Weizmann Institute of Science) for providing the
+RNA-seq data as part of the Avian RNAseq consortium.
 
@@
 microchromosome''.
-
-We thank Matt Schwartz (Harvard) and Igor Ulitsky (Weizmann Institute
-of Science) for access to the RNA-seq data.
 
@@
 
+%% \begin{tabular}{|r|r|r|r|l|}
+%% \hline 
+%% \multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\
+%% \hline 
+%% Number in human & median(48 birds) & Number in chicken & Chicken ncRNAs         & RNA type\\
+%%                 &                  &                   & confirmed with RNA-seq & \\
+%% \hline
+%% 62&25.0&34&12 (35.3\%) &Long non-coding RNA\\ 
+%% 356&499.5&427&272 (63.7\%) &microRNA\\ 
+%% 281&120.0&106&88 (83.0\%) &C/D box snoRNA\\ 
+%% 336&85.5&68&46 (67.6\%) &H/ACA box snoRNA\\ 
+%% 34&13.0&12&11 (91.7\%) &Small cajal body RNA\\ 
+%% 1754&48.5&71&29 (40.8\%) &Major spliceosomal RNA\\ 
+%% 58&3.0&6&2 (33.3\%) &Minor spliceosomal RNA\\ 
+%% 525&82.0&122&79 (64.8\%) &Cis-regulatory element\\ 
+%% 316&6.5&9&3 (33.3\%) &7SK RNA\\ 
+%% 1&0.0&2&0 (0.0\%) &Telomerase RNA\\ 
+%% 9&0.0&2&1 (50.0\%) &Vault RNA\\ 
+%% 892&3.0&3&2 (66.7\%) &Y RNA\\ 
+%% 1084&173.5&300&256 (85.3\%) &Transfer RNA\\ 
+%% 80&9.5&4&2 (50.0\%) &Transfer RNA pseudogene\\ 
+%% 941&3.0&4&2 (50.0\%) &SRP RNA\\ 
+%% 607&7.0&22&10 (45.5\%) &Ribosomal RNA\\ 
+%% 4&1.0&2&2 (100.0\%) &RNase P/MRP RNA\\ 
+%% \hline
+%% 7340&1080.0&1194&817 (68.4\%) &Total\\ 
+%% \hline
+%% \end{tabular}
+
 \begin{tabular}{|r|r|r|r|l|}
@@
 \hline 
-Number in human & median(48 birds) & Number in chicken & Chicken ncRNAs         & RNA type\\
-                &                  &                   & confirmed with RNA-seq & \\
+                &                  &                   & Chicken ncRNAs & \\
+Number in human & median(48 birds) & Number in chicken & confirmed with RNA-seq & RNA type\\
 \hline
 62&25.0&34&12 (35.3\%) &Long non-coding RNA\\ 
-356&499.5&427&272 (63.7\%) &microRNA\\ 
-281&120.0&106&88 (83.0\%) &C/D box snoRNA\\ 
-336&85.5&68&46 (67.6\%) &H/ACA box snoRNA\\ 
-34&13.0&12&11 (91.7\%) &Small cajal body RNA\\ 
-1754&48.5&71&29 (40.8\%) &Major spliceosomal RNA\\ 
-58&3.0&6&2 (33.3\%) &Minor spliceosomal RNA\\ 
-525&82.0&122&79 (64.8\%) &Cis-regulatory element\\ 
+356&499.5&427&280 (65.6\%) &microRNA\\ 
+281&120.0&106&90 (84.9\%) &C/D box snoRNA\\ 
+336&85.5&68&48 (70.6\%) &H/ACA box snoRNA\\ 
+34&13.0&12&12 (100.0\%) &Small cajal body RNA\\ 
+1754&48.5&71&32 (45.1\%) &Major spliceosomal RNA\\ 
+58&3.0&6&3 (50.0\%) &Minor spliceosomal RNA\\ 
+525&82.0&122&88 (72.1\%) &Cis-regulatory element\\ 
 316&6.5&9&3 (33.3\%) &7SK RNA\\ 
@@
 892&3.0&3&2 (66.7\%) &Y RNA\\ 
-1084&173.5&300&256 (85.3\%) &Transfer RNA\\ 
+1084&173.5&300&278 (92.7\%) &Transfer RNA\\ 
 80&9.5&4&2 (50.0\%) &Transfer RNA pseudogene\\ 
@@
 \hline
-7340&1080.0&1194&817 (68.4\%) &Total\\ 
+7340&1080.0&1194&865 (72.4\%) &Total\\ 
 \hline
 \end{tabular}
+
+%totalExpressed[865] totalNegControlsExpressed[123] totalGenes[1194] FPR[10.3]
 

### 2014-04-11 paper/supplementary-material.tex
Working on the RNA-seq validation.

@@
 
+\begin{figure}[ht]
+\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr1-147253243-147253923-miRNA-cluster.pdf}
+  \caption[]{microRNA cluster: mir-19\_3, mir-20\_2, mir-19\_2, mir-18\_2 \& mir-17\_1.}\label{fig:10}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr4-3946041-3946705-miRNA-cluster.pdf}
+  \caption[]{microRNA cluster: mir-92\_2, mir-19\_1, mir-20\_1, mir-18\_1, mir-106\_1.}\label{fig:11}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr23-2507395-2508461-snoRNA-cluster.pdf}
+  \caption[]{H/ACA box snoRNA cluster: SNORA61, SNORA44, SNORA16.}\label{fig:12}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr27-4040511-4042052-tRNA-cluster.pdf}
+  \caption[]{Cysteine tRNA cluster.}\label{fig:13}
+\end{figure}
+
 \end{bmcformat}

### 2014-04-11 scripts/makeTable.pl

@@ -92,7 +92,7 @@ if (-s "data/R/allRNA.dat"){
 		$countsTot{"human"}{$rfam2type{$w}}+=$w[0];
 		
 		$countsTot{"chicken"}{$rfam2type{$w}}=0 if(not defined($countsTot{"chicken"}{$rfam2type{$w}}));
-		$countsTot{"chicken"}{$rfam2type{$w}}+=$w[7];
+		$countsTot{"chicken"}{$rfam2type{$w}}+=$w[5];
 		
 		if(not defined($expressedChickenPaul{$rfam2type{$w}})){
 		    $expressedChickenPaul{$rfam2type{$w}}=0;
@@ -149,8 +149,8 @@ print '
 \hline 
 \multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\\\
 \hline 
-                &                  &                   & \multicolumn{1}{|l|}{Chicken ncRNAs } & \\\\
-                &                  &                   & \multicolumn{1}{|l|}{confirmed with RNA-seq} & \\\\
+                &                  &                   & Chicken ncRNAs & \\\\
+                &                  &                   & confirmed with RNA-seq & \\\\
 \hline
 Number in human & median(48 birds) & Number in chicken & max(RNA$_i$)$>13.0$ & RNA type\\\\
 \hline
@@ -165,8 +165,9 @@ foreach my $k (@printNames){
     }
 #    printf "%d&%0.1f&%d&%d (%0.0f)&$k\\\\ \n", $countsTot{"human"}{$k}, $countsTot{"all-birds"}{$k}/48, $countsTot{"chicken"}{$k}, $expressedChicken{$k}, 100*$expressedChicken{$k}/$countsTot{"chicken"}{$k};
     my $medBirds=median(@allBirds);
+    $countsTot{"chicken"}{$k} = 1 if (not defined($countsTot{"chicken"}{$k}) or $countsTot{"chicken"}{$k} == 0);
     printf "%d&%0.1f&%d&%d (%0.1f\\%%) &$k\\\\ \n", $countsTot{"human"}{$k}, $medBirds, $countsTot{"chicken"}{$k}, 
-    $expressedChickenMario{$k}, 100*$expressedChickenMario{$k}/$countsTot{"chicken"}{$k};    
[…]

### 2014-04-11 paper/manuscript.tex
Working on the RNA-seq validation (2).

@@
 %%{\bf REDUCE? too much talking about what it can't to compared to what it can do...}
-The CM based approach works well for most classes of ncRNA, but the
-long non-coding RNAs (lncRNAs) are a particular challenge
-\cite{Guttman:2009}. Recent technological advances have led to
-dramatic speed and memory-usage enhancements for CM analyses
-\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
-cannot model the exon-intron structures of spliced lncRNAs, nor can
-they deal simply with the repeats that many lncRNAs host. Consequently
-in the latest release of Rfam the lncRNA families that were added were
-composed of local conserved (and possibly structured elements) within
-lncRNAs, analogous to the ``domains'' housed within protein sequences
-\cite{Burge:2013}. Whilst some these regions may not reflect
-functional RNA elements instead regulatory regions, enhancers or
-insulators, their syntenic conservation still provides an indication
-of lncRNA conservation \cite{diederichs2014four}.
+The CM based approach is appropriate for most classes of ncRNA, but
+the long non-coding RNAs (lncRNAs) are a particular challenge
+\cite{Guttman:2009}. 
+%Recent technological advances have led to
+%dramatic speed and memory-usage enhancements for CM analyses
+%\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. 
+CMs cannot model the exon-intron structures of spliced lncRNAs, nor do
+they deal elegantly with the repeats that many lncRNAs
+host. Consequently in the latest release of Rfam the lncRNA families
+that were added were composed of local conserved (and possibly
+structured elements) within lncRNAs, analogous to the ``domains''
+housed within protein sequences \cite{Burge:2013}. Whilst some these
+regions may not reflect functional RNA elements but instead regulatory
+regions, enhancers or insulators, their syntenic conservation still
+provides an indication of lncRNA conservation
+\cite{diederichs2014four}.
 
@@
 \cite{International_Chicken_Genome_Sequencing_Consortium:2004}, zebra
-finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the recently
-published 45 avian genomes \cite{birds:14,birds:14a,Huang:2013,Zhan:2013,Shapiro:2013,Howard:2013,Li:2014}, provides
-exciting opportunities to explore ncRNA conservation in
+finch \cite{Warren:2010} and turkey \cite{Dalloul:2010} with the
+recently published 45 avian genomes
+\cite{birds:14,birds:14a,Huang:2013,Zhan:2013,Shapiro:2013,Howard:2013,Li:2014},
+provides exciting opportunities to explore ncRNA conservation in
 unprecedented detail.
@@
 In the following we explore the conservation patterns of the major
-classes of avian ncRNA in further detail. The collection of ncRNA
-sequences is generally biased towards model organisms
-\cite{Gardner:2010,Hoeppner:2012}.  {\bf Using accurate homology
+classes of avian ncRNA in further detail. 
+{\bf Using accurate homology
   search tools and evolutionary constraints, we produce an accurate
@@
 
+The collection of ncRNA sequences is generally biased towards model
+organisms \cite{Gardner:2010,Hoeppner:2012}. However, we have shown
+that using data from well studied lineages such as mammals can result
+in quality annotations of sister taxa such as aves.
+
 These results indicate we are still in the very early phases of

### 2014-04-15 paper/manuscript.tex
Responding to referee comments.

@@
 %\title{Wins and losses: Non-coding RNAs of bird genomes}
-\title{Fly away Peter, come back Paul: the conservation and losses of avian non-coding RNAs}
+\title{Conservation and losses of avian non-coding RNAs}
 
@@
 
-%% Non-coding RNAs in bird genomes
-
-%% Paul P. Gardner1,2, Sarah W. Burge3, Jana Hertel5, Maria Ninova4, Stephanie Kehr5, Anne Nitsche5, Mario Fasold5, Tammy E. Steeves1, Sam Griffiths-Jones4, Peter Stadler5
-
-%% 1 School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
-2 Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
-3 European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
-4Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
-5 Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, Hartelstrasse 16-18, D-04107 Leipzig, Germany
-
-%% Talking points:
-%% 1.
-Although the majority of classical eukaryotic ncRNAs are conserved in birds, several notable classical ncRNAs are not.
-Homology searches in associated genes indicate these apparent "losses" are most likely due to divergence.
-%% 2.
-Of the 59 snoRNA conserved between humans and yeast, we find 45 are conserved in birds.
-Homology searches in associated genes for the remaining non-conserved snoRNAs indicate these apparent losses are indeed due to multiple losses in several bird lineages.
-%% 3. We find that 3 families of broadly conserved vertebrate microRNA families are lost in the avian lineage. We also annotate a number of bird-specific microRNAs.
-%% 4. Several well-characterised lncRNA families of known function are conserved between mammals and birds.
-%% 5. We discuss the conservation of HOX associated lncRNAs and several lncRNAs that have been implicated in processes associated with cancer.
-
-%% %% Draft Abstract
-%% We present the results of a large-scale bioinformatic annotation of non-coding RNAs in 48 avian genomes.
-We use probabilistic models of hand-curated families from the Rfam database to infer conserved RNA families within each avian genome.
-We supplement these annotations with predictions from the tRNA annotation tool, tRNAscan-SE, microRNAs from miRBase and selected sequences from analyses of the chicken transcriptome.
-Although we show extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds, we also demonstrate apparent ``losses'' in several RNA families.
-These include the divergence of some classical ncRNAs and the genuine loss of several snoRNAs and microRNAs.
-In contrast, we show a signicant number of lncRNAs are surprisingly well conserved between birds and mammals including several intriguing cases where the reported mammalian lncRNA function is not conserved in birds.
-These combined results illustrate the utility of applying homology based methods for annotating vertebrate genomes and illustrate many complex evolutionary patterns within the avian ncRNA cohort.
-
-%% LNCRNA:
-
 \ifthenelse{\boolean{publ}}{\begin{multicols}{2}}{}
@@
 
-%%{\bf REDUCE? too much talking about what it can't to compared to what it can do...}
-The CM based approach is appropriate for most classes of ncRNA, but
-the long non-coding RNAs (lncRNAs) are a particular challenge
-\cite{Guttman:2009}. 
-%Recent technological advances have led to
-%dramatic speed and memory-usage enhancements for CM analyses
-%\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. 
-CMs cannot model the exon-intron structures of spliced lncRNAs, nor do
-they deal elegantly with the repeats that many lncRNAs
-host. Consequently in the latest release of Rfam the lncRNA families
-that were added were composed of local conserved (and possibly
-structured elements) within lncRNAs, analogous to the ``domains''
-housed within protein sequences \cite{Burge:2013}. Whilst some these
-regions may not reflect functional RNA elements but instead regulatory
-regions, enhancers or insulators, their syntenic conservation still
-provides an indication of lncRNA conservation
-\cite{diederichs2014four}.
-
 %% The functions determined to date for lncRNAs range
@@
 In the following we explore the conservation patterns of the major
-classes of avian ncRNA in further detail. 
-{\bf Using accurate homology
-  search tools and evolutionary constraints, we produce an accurate
-  set of genome annotations for 48 predominantly non-model bird
-  species for ncRNAs that are conserved across the avian species. This
-  conservative set of annotations is expected to contain the core
-  avian ncRNAs.}  We focus our report on the unusual results within
-the avian lineages. These are either unexpectedly well-conserved
-ncRNAs or unexpectedly poorly-conserved ncRNAs. The former are ncRNAs
-that are not expected to be conserved between the birds and the other
-vertebrates, particularly those ncRNAs whose function is not conserved
-in birds.  The latter are apparent losses of ncRNAs expected to be
-conserved; Here, we consider three categories of such ``loss'': First,
-genuine gene losses in the avian lineage where ncRNAs well conserved
-in other vertebrates are completely absent in birds.  Second,
-``divergence'' where ncRNAs have undergone such significant sequence
-and structural alternations that homology search tools can no longer
-detect a relationship between other vertebrate exemplars and avian
-varieties.  Third, ``missing'' ncRNAs that failed to be captured in
-the available, largely fragmented, avian genomes.  We postulate that
-the latter category is likely to be prevalent in comparative avian
-genome studies given the distinctive organisation of the avian
-genome. Namely, the avian karyotype is characterised by a large number
-of chromosomes (average $2n \approx 80$) generally consisting of a few
-larger ``macrochromosomes'' and many smaller ``microchromosomes''
+classes of avian ncRNA in further detail.  Using homology search tools
+and evolutionary constraints, we produce an accurate set of genome
+annotations for 48 predominantly non-model bird species for ncRNAs
+that are conserved across the avian species. This conservative set of
+annotations is expected to contain the core avian ncRNAs. We focus our
+report on the unusual results within the avian lineages. These are
+either unexpectedly well-conserved ncRNAs or unexpectedly
+poorly-conserved ncRNAs. The former are ncRNAs that were not expected
+to be conserved between the birds and the other vertebrates,
+particularly those ncRNAs whose function is not conserved in birds.
+The latter are apparent losses of ncRNAs expected to be conserved;
+Here, we consider three categories of such ``loss'': First, genuine
+gene losses in the avian lineage where ncRNAs well conserved in other
+vertebrates are completely absent in birds.  Second, ``divergence''
+where ncRNAs have undergone such significant sequence and structural
+alternations that homology search tools can no longer detect a
+relationship between other vertebrate exemplars and avian varieties.
+Third, ``missing'' ncRNAs that failed to be captured in the available,
+largely fragmented, avian genomes.  We postulate that the latter
+category is likely to be prevalent in comparative avian genome studies
+given the distinctive organisation of the avian genome. Namely, the
+avian karyotype is characterised by a large number of chromosomes
+(average $2n \approx 80$) generally consisting of a few larger
+``macrochromosomes'' and many smaller ``microchromosomes''
 \cite{Griffin:2007}. This ’so many, so small’ pattern presents
@@
 
+There is substantial gain and loss of lncRNA
+\cite{Cabili:2011,Kutter:2012} and more generally ncRNA loci over
+evolutionary time \cite{Hoeppner:2012}. It is difficult to assess how
+much of this is due to bioinformatic sequence alignment tools failing
+to align correctly below 60-50\% sequence identity
+\cite{Gardner:2005:Nucleic-Acids-Res:15860779} or due to genuine gains
+and losses. Nevertheless, sequence conservation, generally speaking,
+provides useful evidence for gene and function conservation.
+
+We have identified 66,879 loci in 48 avian genomes that share sequence
+similarity with previously characterised ncRNAs and are conserved in
+$>10\%$ of these avian genomes. These loci have been classified into
+626 different families, the majority of which are miRNAs and snoRNAs
+(summarised in Table~\ref{table:1}). Out of necessity we have selected
+the most charismatic families for further discussion. These are
+lncRNAs that appear to be conserved between mammals and aves and the
+cases of apparent loss of genes that conserved in most other
+vertebrates.
+
 \subsection*{Unusually well conserved RNAs}
@@
 RNAs that have been implicated in a multitude of functional processes
-\cite{Rinn:2007,Chow:2005,Guttman:2009}. These RNAs have largely been
+\cite{Rinn:2007,Chow:2005,Guttman:2009,Ulitsky:2013}. These RNAs have largely been
 characterised in mammalian species, particularly human and
@@
 
-In general, Rfam cannot include the entire length of any large,
-spliced RNAs. This is a limitation of the covariance-models used for
-the homology-searches Rfam runs \cite{Nawrocki:2009}. Consequently,
-only short, well-conserved regions with evolutionarily conserved
-secondary structures are included in Rfam. By analogy to
-protein-domains, we refer to these as RNA-domains
-\cite{Burge:2013}. {\bf SOME OF THESE MAY BE DNA ELEMENTS RATHER THAN
-  LNCRNA ELEMENTS!!!!}
+The CM based approach is appropriate for most classes of ncRNA, but
+the long non-coding RNAs (lncRNAs) are a particular challenge
+\cite{Guttman:2009}. 
+CMs cannot model the exon-intron structures of spliced lncRNAs, nor do
+they deal elegantly with the repeats that many lncRNAs
+host. Consequently in the latest release of Rfam the lncRNA families
+that were added were composed of local conserved (and possibly
+structured elements) within lncRNAs, analogous to the ``domains''
+housed within protein sequences \cite{Burge:2013}. Whilst some these
+regions may not reflect functional RNA elements but instead regulatory
+regions, enhancers or insulators, their syntenic conservation still
+provides an indication of lncRNA conservation
+\cite{diederichs2014four}.
 
@@
 The NBR2 lncRNA and BRCA1 gene share a bidirectional promotor
-\cite{Xu:1997}. Both are expressed in a broad ranch of
+\cite{Xu:1997}. Both are expressed in a broad range of
 tissues. Extensive research on BRCA1 has shown that it is involved in
@@
 conservation or expansion within the avian lineage. The only exception
-being the snoRNA SNORD93 which has 92 copies in the tinamou
+being the snoRNA, SNORD93 which has 92 copies in the tinamou
 genome, whereas it only has 1-2 copies in all the other vertebrate
@@
 $\sim900$ and $\sim580$ copies in the human, turtle and alligator
-genomes, the average copies numbers of these drop to $\sim280$ and
+genomes, the average copy numbers of these drop to $\sim280$ and
 $\sim100$ copies in the avian genomes. In addition to reduction in
@@
 additional homologs for several RNA families that fill gaps in the
-abundance matrices (Figure ?) strongly suggests that conspicuous
+abundance matrices (Figure~1) strongly suggests that conspicuous
 absences, in particular of LUCA and LECA RNAs, are caused by
@@
 
-{\bf Briefly mention the reduction in number of pseudogenes. Pick a
-  few key human ones and compare with the birds. }
+%{\bf Briefly mention the reduction in number of pseudogenes. Pick a
+%  few key human ones and compare with the birds. }
+
+Non-coding RNA derived pseudogenes are a major problem for many ncRNA
+annotation projects. The human genome, for example, contains $>1$
+million Alu repeats, which are derived from the SRP RNA
+\cite{Wheeler:2013}. The existing Rfam annotation of the human genome,
+in particular, contains a number of problematic families that appear
+to have been excessively pseudogenised. The U6 snRNA, SRP RNA and Y
+RNA families have 1,371, 941 and 892 annotations in the human
+genome. These are a heterogenous mix of pseudogenised, paralogous,
+diverged or functional copies of these families. Unfortunately, a
+generalised model of RNA pseudogenes has not been incorporated into
+the main covariance model package, Infernal. An approach used by tRNAscan
+\cite{Lowe:1997}, is, in theory, generalisable to other RNA families
+but this remains a work in progress.
+
+Is it possible that the avian annotations also contains excessive
+pseudogenes. However, it has previously been noted that avian genomes
+are significantly smaller than other vertebrate species
+\cite{International_Chicken_Genome_Sequencing_Consortium:2004}. We
+have also noted a corresponding reduction in the number of paralogs
+and presumed ncRNA-derived pseudogenes in the avian genomes (See
+Supplemental Figure~12). The problematic human families, U6 snRNA, SRP
+RNA and Y RNA have, for example, just 26, 4 and 3 annotations
+respectively in the chicken genome and 13, 3 and 3 annotations
+respectively, on average, in the 48 avian genomes used here.
+Therefore, we suspect that the majority of our annotations are in fact
+functional orthologs.
 
@@
 mapping the RNA-seq data to the chicken genome (see Methods for
-details), we identified a threshold for calling a gene as
-expressed by limiting our estimated false-positive rate to less than
-10\%. This FDR was estimated using a negative control of randomly
-selected, un-annotated regions of the genome. Since some regions may
-be genuinely expressed, the true FDR is potentially much lower than
-10\%. Overall, the number of ncRNAs we
-have identified in this work that are expressed above background
-levels is 865 (72.4\%) (see Table~\ref{table:1}). This shows that 7.0
-times more of our ncRNAs are expressed than expected by chance
-(Fisher's exact test: $P<10^{16}$). This number is an underestimate of
-the fraction of our annotations that are genuinely expressed, as only
-a fraction of the developmental stages and tissues of chicken have
-been characterized with RNA-seq. Furthermore, some ncRNAs
-are expressed in highly specific conditions  \cite{mercer2008specific,johnston2003microrna}.
+details), we identified a threshold for calling a gene as expressed by
+limiting our estimated false-positive rate to approximately 10\%. This
+FDR was estimated using a negative control of randomly selected,
+un-annotated regions of the genome. Since some regions may be
+genuinely expressed, the true FDR is potentially lower than
+10\%. Overall, the number of ncRNAs we have identified in this work
+that are expressed above background levels is 865 (72.4\%) (see
+Table~\ref{table:1}). This shows that 7.0 times more of our ncRNAs are
+expressed than expected by chance (Fisher's exact test:
+$P<10^{16}$). This number is an underestimate of the fraction of our
+annotations that are genuinely expressed, as only a fraction of the
+developmental stages and tissues of chicken have been characterized
+with RNA-seq. Furthermore, some ncRNAs are expressed in highly
+specific conditions \cite{mercer2008specific,johnston2003microrna}.
 
@@
 or developmental-stage, in fact some RNAs have extremely specific
-expression profiles \cite{Johnston:2003}.  We have identified
-previously unrecognised conservation of ncRNAs in avian genomes as
-well as some surprising ``losses'' of otherwise well conserved
-ncRNAs. We suspect many of these losses are due to a combination of
-limitations in the homology search tools that we use for annotation
-and the ability of ncRNAs to tolerate large amounts of sequence
-variation while remaining functional, rather than \emph{bona fide}
-gene loss. In some cases these losses could be due to missing data
-from the genome assemblies, but this unlikely to be the case for
-multiple independant assemblies.
+expression profiles, e.g. the lsy-6 microRNA \cite{Johnston:2003}.  We
+have identified previously unrecognised conservation of ncRNAs in
+avian genomes and some surprising ``losses'' of otherwise well
+conserved ncRNAs. We have shown that most of these losses are due to
+difficulties assembling avian microchromosomes rather than \emph{bona
+  fide} gene loss. A large fraction of our annotations have been
+confirmed using RNA-seq data, which also showed a 7-fold enriched of
+expression within our annotations relative to unannotated regions.
 
@@
 organisms \cite{Gardner:2010,Hoeppner:2012}. However, we have shown
-that using data from well studied lineages such as mammals can result
-in quality annotations of sister taxa such as aves.
+that using data from well studied lineages such as mammals can also
+result in quality annotations of sister taxa such as aves.
 
-These results indicate we are still in the very early phases of
+In summary, these results indicate we are in the very early phases of
 determining the functions of many RNA families. This is illustrated by
@@
 were retained. The Rfam database classifies non-coding RNAs into
-hierarchial groupings. The basic units are ``families'' which are
+hierarchical groupings. The basic units are ``families'' which are
 groups of homologous, alignable sequences; ``clans'' which are groups
@@
 
-In order to obtain good annotations of tRNA genes we also ran the
+In order to obtain good annotations of tRNA genes we ran the
 specialist tRNA-scan version 1.3.1 annotation tool. This method also
@@
 increase the specificity of our annotations we filtered out families
-that were indentified in four or fewer of the 51 vertebrate species we
+that were identified in four or fewer of the 51 vertebrate species we
 have analysed in this work. These filtered families largely
@@
 INFERNAL (v1.1rc3), and these models were searched against the 48 bird
-genomes, and the genomes of the american alligator and the green
-turtle as outgroups. Hits with e-value <10 realigned with the query
+genomes, and the genomes of the American alligator and the green
+turtle as outgroups. Hits with e-value $<10$ realigned with the query
 sequences and the resultant multiple sequence alignments manually
@@
 
-An additonal snoRNA homology search was performed with snoStrip
+An additional snoRNA homology search was performed with snoStrip
 \cite{Bartschat:2013}. As initial queries we used deutorostomian
@@
 
-The diverse sets of genome annotations were combined. We collapsed the
-overlapping annotations into a single annotation. We also generated
-heatmaps for different groups of ncRNA genes (see Figure~\ref{fig:1}
-and Sup. Figs. 1-3). All the scripts and annotations presented here
-are available from Github \cite{gitrepo}.
+The diverse sets of genome annotations were combined and filtered,
+ensuring conservation in 10\% or more of the avian genomes. We
+collapsed the remaining overlapping annotations into a single
+annotation. We also generated heatmaps for different groups of ncRNA
+genes (see Figure~\ref{fig:1} and Supplemental Figure~1-3). All the
+scripts and annotations presented here are available from Github
+\cite{gitrepo}.
 
@@
 data sets. The first data set (Bioproject PRJNA204941) contains 971
-million reads and comprises 27 samples from 14 diferent chicken
+million reads and comprises 27 samples from 14 different chicken
 tissues sequenced on Illumina HiSeq2000 using a small RNA-seq
@@
 
-Heatmaps showing the prescence/abscence and approximate genomic copy
+Heatmaps showing the presence/absence and approximate genomic copy
 number of ``unusually, well conserved RNAs'' (particularly the

### 2014-04-15 paper/supplementary-material.tex
Responding to referee comments.

@@
 Over a third of the apparent losses of the yeast-human conserved
-snoRNA families appear to cluster on 2 locii of the ancestral
+snoRNA families appear to cluster on 2 loci of the ancestral
 vertebrate genome. We investigated these losses further.
@@
 
-The simplest case are the selenocystein tRNAs. Here, tRNAscan is tuned
-for specificity and thus misses several occurrances that are easily
-found by \texttt{blastn} with $E\le 10^{-30}$. In some cases the
-sequences appear degraded at the ends, which may be explained e.g.\ by
-low sequence quality at the very ends of contigs or scaffolds. A
-\texttt{blastn} search also readily retrieves additional RNase P and
-RNAse MRP RNAs, capturing only the best conserved regions. In many
-cases these additional candidates are incomplete or contain
-undetermined sequence, explaining why they are missed by the
+The simplest case are the selenocysteine tRNAs. Here, tRNAscan is
+tuned for specificity and thus misses several occurrances that are
+easily found by \texttt{blastn} with $E\le 10^{-30}$. In some cases
+the sequences appear degraded at the ends, which may be explained
+e.g.\ by low sequence quality at the very ends of contigs or
+scaffolds. A \texttt{blastn} search also readily retrieves additional
+RNase P and RNAse MRP RNAs, capturing only the best conserved
+regions. In many cases these additional candidates are incomplete or
+contain undetermined sequence, explaining why they are missed by the
 CMs. Overall, we identify tRNA-Sec in most and RNAse P and MRP RNAs in
@@
 notoriously hard to identify by homology search \cite{Xie:08a}. The
-poor return thus does not come as a surprise. Since \texttt{blastn}
-searches remained unsuccessfull we constructed a sauropsid-specific CM
-for the vault RNA. In addition to the hits identified by the Rfam
-model we obtained three additional homologs. Vault RNAs, with a size
-of about 100 nt, exhibit conserved sequence patterns only at their
-ends, with essentially unconstrained sequence in the central
-part. Their identification is one of the well-know difficult problems
-for homology search \cite{Stadler:09b,Kolbe:2009}.
+poor return thus does not come as a surprise. Vault RNA homology
+searches with \texttt{blastn} remained unsuccessful, therefore we
+constructed a sauropsid-specific CM.  In addition to the hits
+identified by the Rfam model we obtained three additional
+homologs. Vault RNAs, with a size of about 100 nt, exhibit conserved
+sequence patterns only at their ends, with essentially unconstrained
+sequence in the central part. Their identification is one of the
+well-known and difficult problems for homology search
+\cite{Stadler:09b,Kolbe:2009}.
 
@@
 
+\begin{figure}[ht]
+\includegraphics[width=0.475\textwidth]{figures/high-copy-numbers.pdf}
+\includegraphics[width=0.475\textwidth]{figures/high-copy-numbers-U6-SRP.pdf}
+  \caption[]{{\bf Left:} A comparison of the genomic copy numbers for
+    26 of the most abundant RNAs found in human, reptile and the bird
+    genomes. {\bf Right:} Bit-score distributions for the Rfam U6,
+    Metazoan SRP and the Y RNA covariance models for human and chicken
+    paralogues/pseudogenes. }\label{fig:14}
+\end{figure}
+
 \end{bmcformat}

### 2014-04-15 scripts/gffs2heatmaps.pl

@@ -313,7 +313,7 @@ Metazoa_SRP RNase_MRP RNaseP_nuc Telomerase-vert U1 U2 U4 U5 U6 U11 U12 U4atac U
 tRNA-pseudogene tRNA);
 system("head -n 1 $rDir/allRNA.dat | perl -lane \47" .  's/_\d+X\t/\t/g; print' . "\47 > $rDir/diverged.dat");
 foreach my $fm7 ( @divFams ){
-    system("egrep \47$fm7\$\47 $rDir/allRNA.dat >> $rDir/diverged.dat")
+    system("egrep \47$fm7\$\47 $rDir/allRNA.dat >> $rDir/diverged.dat");
 }
 system("egrep \47SeC\47 $rDir/tRNA.dat >> $rDir/diverged.dat");
 
@@ -324,8 +324,24 @@ system("egrep \47SeC\47 $rDir/tRNA.dat >> $rDir/diverged.dat");
 my @unConFams = qw(DLEU2_5 DLEU2_6 HOXA11-AS1_1 HOXA11-AS1_2 HOXA11-AS1_3 HOXA11-AS1_4 HOXA11-AS1_5 HOXA11-AS1_6 mir-15 mir-16 NBR2 PART1_1 PART1_2 PART1_3 PCA3_1 PCA3_2 RMST_6 RMST_7 RMST_8 RMST_9 Six3os1_5 Six3os1_6 Six3os1_7 SNORD93 SOX2OT_exon1 SOX2OT_exon2 SOX2OT_exon3 SOX2OT_exon4 ST7-OT3_2);
 system("head -n 1 $rDir/allRNA.dat | perl -lane \47" .  's/_\d+X\t/\t/g; print' . "\47 > $rDir/unusual-conserved.dat");
 foreach my $fm8 (@unConFams){
-    system("egrep \47$fm8\$\47 $rDir/allRNA.dat >> $rDir/unusual-conserved.dat")
+    system("egrep \47$fm8\$\47 $rDir/allRNA.dat >> $rDir/unusual-conserved.dat");
 }
+
+#grep human data/R/allRNA.dat && cat blah* | cut -f 1 | sort -d | uniq -c | sort -nr | nl | head -n 31 | awk '{print $3}' | sort -d | perl -lane 'print "grep $_\$ data/R/allRNA.dat"' | sh
[…]

### 2014-04-15 scripts/heatmaps.R

@@ -132,6 +132,65 @@ heatmap.2(as.matrix(allrna),trace="none",col=gray(lb:0/lb),
           )
 dev.off()
 
+######################################################################
+
+allrna<-read.table("data/R/high-copy-numbers.dat",header = T, sep = "\t",row.names=52)
+pdf(file="paper/figures/high-copy-numbers.pdf",width=8, height=10)
+op<-par(las=2,mar=c(7, 4, 4, 2) + 0.1,cex=1.5); #bottom, left, top, right
+plot(log10(allrna$human), pch=1, yaxt="n", xaxt="n", ylab="Genomic copy number", xlab="", main="Genomic copy numbers for high-copy ncRNAs"); 
+axis(2,0:3, 10^(0:3)); 
+axis(1,1:(length(rownames(allrna))), rownames(allrna)); 
+points(log10(allrna$chicken), pch=2, col="red"); 
+points(log10(allrna$alligator), pch=3); 
+points(log10(allrna$turtle), pch=4);  
+points(log10(allrna$zebrafinch), pch=6, col="red"); 
+points(log10(allrna$budgerigar), pch=7, col="red"); 
+points(log10(rowMeans(allrna[,4:51])),col="red",pch=5); 
+legend("topright", c("Human", "Alligator", "Turtle", "Chicken", "Zebrafinch", "Budgerigar", "mean.48.Birds"), pch=c(1,3,4,2,6,7,5), col=c("black","black","black","red","red","red","red"))
+dev.off()
+
+humU6<-read.table("data/R/U6-human-bitscores.dat",header = F, sep = "\t")
+chickU6<-read.table("data/R/U6-chicken-bitscores.dat",header = F, sep = "\t")
+humSRP<-read.table("data/R/SRP-human-bitscores.dat",header = F, sep = "\t")
+chickSRP<-read.table("data/R/SRP-chicken-bitscores.dat",header = F, sep = "\t")
+humY<-read.table("data/R/Y_RNA-human-bitscores.dat",header = F, sep = "\t")
+chickY<-read.table("data/R/Y_RNA-chicken-bitscores.dat",header = F, sep = "\t")
+
+allU6<-c(humU6$V1,chickU6$V1)
+breaks<-seq(floor(min(allU6)),ceiling(max(allU6)), len=40)
[…]

### 2014-05-01 paper/manuscript.tex
Updated Acknowledgements.

@@
 
-Magnus Alm Rosenblad (Univ. of Gothenburg) for useful discussions.
+Magnus Alm Rosenblad (Univ. of Gothenburg) and Eric Nawrocki (HHMI
+Janelia Farm) for useful discussions.
 
-We thank Fiona McCarthy (Mississippi State University) and Carl
+We thank Fiona McCarthy (University of Arizona) and Carl
 Schmidt (University of Delaware) as well as Matt Schwartz (Harvard)

### 2014-05-19 paper/manuscript.tex
More responses to referees, tidying, etc.

@@
 %\title{Wins and losses: Non-coding RNAs of bird genomes}
-\title{Conservation and losses of avian non-coding RNAs}
+\title{Conservation and losses of avian non-coding RNA loci}
 
@@
 Here we present the results of a large-scale bioinformatic annotation
-of non-coding RNAs in 48 avian genomes. Our approach uses
+of non-coding RNA loci in 48 avian genomes. Our approach uses
 probabilistic models of hand-curated families from the Rfam database
@@
 supplement these annotations with predictions from the tRNA annotation
-tool, tRNAscan-SE and microRNAs from miRBase.  We show a significant
-number of lncRNAs are surprisingly well conserved between birds and
-mammals including several intriguing cases where the reported
-mammalian lncRNA function is not conserved in birds.  We also
-demonstrate extensive conservation of classical ncRNAs (e.g., tRNAs)
-and more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in
-birds. Furthermore, we describe numerous ``losses'' of several RNA
-families, and attribute these to genuine loss, divergence or missing
-data.  In particular, issues with assembling microchromosomes using
-next-generation sequencing methods. These combined results illustrate
-the utility of applying homology based methods for annotating novel
-vertebrate genomes and illustrate many complex evolutionary patterns
-within the avian ncRNA cohort.
+tool, tRNAscan-SE and microRNAs from miRBase.  We show a number of
+lncRNA associated loci are conserved between birds and mammals,
+including several intriguing cases where the reported mammalian lncRNA
+function is not conserved in birds.  We also demonstrate extensive
+conservation of classical ncRNAs (e.g., tRNAs) and more recently
+discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds. Furthermore, we
+describe numerous ``losses'' of several RNA families, and attribute
+these to genuine loss, divergence or missing data.  In particular, we
+show that many of these losses are due to issues with assembling
+microchromosomes using next-generation sequencing methods. These
+combined results illustrate the utility of applying homology-based
+methods for annotating novel vertebrate genomes.
 \end{abstract}
@@
 Independent benchmarks of bioinformatic annotation tools have shown
-that the CM approaches dramatically out-perform alternative methods
+that the CM approaches out-perform alternative methods
 \cite{Freyhult:2007}, although its sensitivity is limited for the most
@@
 \cite{birds:14,birds:14a,Huang:2013,Zhan:2013,Shapiro:2013,Howard:2013,Li:2014},
-provides exciting opportunities to explore ncRNA conservation in
+provides an exciting opportunity to explore ncRNA conservation in
 unprecedented detail.
@@
 In the following we explore the conservation patterns of the major
-classes of avian ncRNA in further detail.  Using homology search tools
-and evolutionary constraints, we produce an accurate set of genome
+classes of avian ncRNA loci in further detail.  Using homology search
+tools and evolutionary constraints, we have produced a set of genome
 annotations for 48 predominantly non-model bird species for ncRNAs
 that are conserved across the avian species. This conservative set of
-annotations is expected to contain the core avian ncRNAs. We focus our
-report on the unusual results within the avian lineages. These are
+annotations is expected to contain the core avian ncRNA loci. We focus
+our report on the unusual results within the avian lineages. These are
 either unexpectedly well-conserved ncRNAs or unexpectedly
-poorly-conserved ncRNAs. The former are ncRNAs that were not expected
-to be conserved between the birds and the other vertebrates,
+poorly-conserved ncRNAs. The former are ncRNA loci that were not
+expected to be conserved between the birds and the other vertebrates,
 particularly those ncRNAs whose function is not conserved in birds.
-The latter are apparent losses of ncRNAs expected to be conserved;
+The latter are apparent losses of ncRNA loci expected to be conserved;
 Here, we consider three categories of such ``loss'': First, genuine
@@
 avian karyotype is characterised by a large number of chromosomes
-(average $2n \approx 80$) generally consisting of a few larger
-``macrochromosomes'' and many smaller ``microchromosomes''
+(average $2n \approx 80$) generally consisting of approximately 5
+larger ``macrochromosomes'' and many smaller ``microchromosomes''
 \cite{Griffin:2007}. This ’so many, so small’ pattern presents
 significant assembly challenges \cite{Ellegren:2005}. Indeed, of the
-48 published avian genomes, 20 of which are high-coverage ($>50X$ ),
+48 published avian genomes, 20 of which are high-coverage ($>50X$),
 only two are chromosomally assembled (chicken and zebra finch;
@@
 
-There is substantial gain and loss of lncRNA
+There is substantial gain and loss of lncRNAs
 \cite{Cabili:2011,Kutter:2012} and more generally ncRNA loci over
 evolutionary time \cite{Hoeppner:2012}. It is difficult to assess how
-much of this is due to bioinformatic sequence alignment tools failing
-to align correctly below 60-50\% sequence identity
-\cite{Gardner:2005:Nucleic-Acids-Res:15860779} or due to genuine gains
-and losses. Nevertheless, sequence conservation, generally speaking,
-provides useful evidence for gene and function conservation.
+many of these observations are due to limited bioinformatic sequence
+alignment tools (these generally fail align correctly below 60-50\%
+sequence identity \cite{Gardner:2005:Nucleic-Acids-Res:15860779}) or
+due to genuine gains and losses. Nevertheless, sequence conservation,
+generally speaking, provides useful evidence for gene and function
+conservation.
 
@@
 $>10\%$ of these avian genomes. These loci have been classified into
-626 different families, the majority of which are miRNAs and snoRNAs
-(summarised in Table~\ref{table:1}). Out of necessity we have selected
-the most charismatic families for further discussion. These are
-lncRNAs that appear to be conserved between mammals and aves and the
-cases of apparent loss of genes that conserved in most other
-vertebrates.
+626 different families, the majority of which correspond to miRNAs and
+snoRNAs (summarised in Table~\ref{table:1}). Out of necessity we have
+selected the most charismatic families for further discussion. These
+include the lncRNAs that appear to be conserved between Mammals and
+Aves and the cases of apparent loss of genes that conserved in most
+other Vertebrates.
 
@@
 and H19 \cite{Smits:2008}.  There is emerging evidence for the
-conservation of ``mammalian'' lncRNAs in other vertebrates
+conservation of ``Mammalian'' lncRNAs in other Vertebrates
 \cite{Chodroff:2010,Ulitsky:2011}), however, like most lncRNAs, the
@@
 The CM based approach is appropriate for most classes of ncRNA, but
-the long non-coding RNAs (lncRNAs) are a particular challenge
-\cite{Guttman:2009}. 
-CMs cannot model the exon-intron structures of spliced lncRNAs, nor do
+the lncRNAs are a particular challenge \cite{Guttman:2009}.  CMs
+cannot model the exon-intron structures of spliced lncRNAs, nor do
 they deal elegantly with the repeats that many lncRNAs
@@
 The HOX cluster lncRNAs HOTAIRM1 (5 RNA-domains), HOXA11-AS1 (6
-RNA-domains), and HOTTIP (4 RNA domains) are remarkably well conserved. In
-the human genome they are located in the HOXA cluster (hg coordinates
-chr7:27135743-27245922), one of the most highly conserved regions in
-vertebrate genomes \cite{PascualAnaya:13}, in antisense orientation between
-HoxA1 and HoxA2, between HoxA11 and HoxA13, and upstream of HoxA13,
-respectively. Conservation and expression of HOTAIRM1 and HOXA11-AS1 within
-the HOXA cluster has been studied in some detail in marsupials
-\cite{Yu:12}.  Of the 15 RNA-domains five and six representing all three
-lncRNAs were recovered in the alligator and turtle genomes. All of them
-appear in the correct order at the expected, syntenically conserved
-positions within the HOXA cluster.  In the birds, where two or more of the
-HOX cluster lncRNA RNA-domains were predicted on the same scaffold, this
-gene order and location within HOX was also preserved.
+RNA-domains), and HOTTIP (4 RNA domains) are conserved across the
+Mammalian and Avian lineages. In the human genome they are located in
+the HOXA cluster (hg coordinates chr7:27135743-27245922), one of the
+most highly conserved regions in vertebrate genomes
+\cite{PascualAnaya:13}, in antisense orientation between HoxA1 and
+HoxA2, between HoxA11 and HoxA13, and upstream of HoxA13,
+respectively. Conservation and expression of HOTAIRM1 and HOXA11-AS1
+within the HOXA cluster has been studied in some detail in marsupials
+\cite{Yu:12}.  Of the 15 RNA-domains five and six representing all
+three lncRNAs were recovered in the alligator and turtle genomes. All
+of them appear in the correct order at the expected, syntenically
+conserved positions within the HOXA cluster.  In the birds, where two
+or more of the HOX cluster lncRNA RNA-domains were predicted on the
+same scaffold, this gene order and location within HOX was also
+preserved.
 
@@
 yet its conservation across the vertebrates certainly implies a
-function (See Figure~\ref{fig:1}).
+function (See Figure~\ref{fig:1}). Possibly at the DNA level, however,
+function at the RNA level cannot be ruled out at this stage.
 
@@
 
-Is it possible that the avian annotations also contains excessive
+It is possible that the avian annotations also contains excessive
 pseudogenes. However, it has previously been noted that avian genomes
@@
 respectively, on average, in the 48 avian genomes used here.
-Therefore, we suspect that the majority of our annotations are in fact
+Therefore, we conclude that the majority of our annotations are in fact
 functional orthologs.
@@
 
+The 48 bird genome sequences used for the following analyses are available from
+the phylogenomics analysis of birds website \cite{birdphylogen}.
+
 Bird genomes were searched using the cmsearch program from INFERNAL
@@
 tissues sequenced on Illumina HiSeq2000 using a small RNA-seq
-protocol. The second data set (personal commun. Matt Schwartz and Igor
-Ulitsky) contains 1,46 billion Illumina HiSeq reads sequenced from
-whole chicken embryo RNA from 7 stages using a strand-specific dUTP
-protocol. Raw reads were checked for quality and adapters clipped if
-required by the protocol. Preprocessed reads were mapped to the
-galGal4 reference genome using SEGEMEHL short read aligner
-\cite{segemehl} and then overlapped with the ncRNA annotations.
+protocol. The second data set (SRA accession SRP041863) contains 1,46
+billion Illumina HiSeq reads sequenced from whole chicken embryo RNA
+from 7 stages using a strand-specific dUTP protocol. Raw reads were
+checked for quality and adapters clipped if required by the
+protocol. Preprocessed reads were mapped to the galGal4 reference
+genome using SEGEMEHL short read aligner \cite{segemehl} and then
+overlapped with the ncRNA annotations.
 
@@
 lncRNAs) on the left and families that have been identified as
-surprising RNA Losses, divergence or missing data. In several cases
+RNA losses, divergence or missing data. In several cases
 functionally related families have also been included, e.g. the RNA

### 2014-05-19 paper/supplementary-material.tex
More responses to referees, tidying, etc.

@@
   \includegraphics[width=0.95\textwidth]{figures/Y.pdf}
-  \caption[]{...}\label{fig:6}
+  \caption[]{Syntenic conservation of the Y RNA cluster in avian and
+    reptile genomes.}\label{fig:6}
 \end{figure}
@@
   \includegraphics[width=0.95\textwidth]{figures/mir-17-cluster.pdf}
-  \caption[Presence/absence table of mir-17 miRNA clusters]{
-This figure illustrates the distribution of all members of the mir-17 clusters across birds, compared to fish and mammals.
-Columns correspond to single miRNA sequences, grouped by clusters and ordered by their position within the corresponding cluster.
-Rows correspond to the species and the cells contain the number of copies of the miRNA in the respective species.
-Colors correspond to orthologous miRNA families: mir-17 (orange), mir-19 (blue) and mir-92/25 (green).There are 3 clusters, I-1 and I-X, and cluster II.
-While the two copies of cluster I are completely conserved, cluster II has been lost as a whole in birds.}\label{fig:8}
+  \caption[Presence/absence table of mir-17 miRNA clusters]{ This
+    figure illustrates the distribution of all members of the mir-17
+    clusters across birds, compared to fish and mammals. Columns
+    correspond to single miRNA sequences, grouped by clusters and
+    ordered by their position within the corresponding cluster. Rows
+    correspond to the species and the cells contain the number of
+    copies of the miRNA in the respective species. Colors correspond
+    to orthologous miRNA families: mir-17 (orange), mir-19 (blue) and
+    mir-92/25 (green).There are 3 clusters, I-1 and I-X, and cluster
+    II.  While the two copies of cluster I are completely conserved,
+    cluster II has been lost as a whole in birds.}\label{fig:8}
 \end{figure}
@@
   \includegraphics[width=0.95\textwidth]{figures/let-7-cluster.pdf}
-  \caption[Presence/absence table of let-7 miRNA clusters]{This figure illustrates the distribution of the let-7 paralogs and clustered miRNAs of the families mir-100 and mir-125.
-Columns correspond to single miRNA sequences, grouped by clusters and ordered by their position within the corresponding cluster.
-Rows correspond to the species and the cells contain the number of copies of the miRNA in the respective species.
-Cluster A, which is strongly conserved in vertebrates has been completely lost in the avian lineage.
-Another obvious loss in birds is cluster F.
-However, in tinamou we detected an unclustered miRNA that best fits to F-let-7-1 in sequence and structure, while we could not find an I-let-7 miRNA sequence in this species.}\label{fig:9}
+  \caption[Presence/absence table of let-7 miRNA clusters]{This figure
+    illustrates the distribution of the let-7 paralogs and clustered
+    miRNAs of the families mir-100 and mir-125. Columns correspond to
+    single miRNA sequences, grouped by clusters and ordered by their
+    position within the corresponding cluster. Rows correspond to the
+    species and the cells contain the number of copies of the miRNA in
+    the respective species. Cluster A, which is strongly conserved in
+    vertebrates has been completely lost in the avian lineage. Another
+    obvious loss in birds is cluster F. However, in tinamou we
+    detected an unclustered miRNA that best fits to F-let-7-1 in
+    sequence and structure, while we could not find an I-let-7 miRNA
+    sequence in this species.}\label{fig:9}
 \end{figure}
@@
 \begin{figure}[ht]
-\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr1-147253243-147253923-miRNA-cluster.pdf}
-  \caption[]{microRNA cluster: mir-19\_3, mir-20\_2, mir-19\_2, mir-18\_2 \& mir-17\_1.}\label{fig:10}
+%\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr1-147253243-147253923-miRNA-cluster.pdf}
+\includegraphics[width=0.99\textwidth]{figures/mir-17-I-1-cluster.pdf}
+  \caption[]{microRNA 17 I-1 cluster: mir-92\_1, mir-19\_3, mir-20\_2,
+    mir-19\_2, mir-18\_2 \& mir-17\_1. The figure indicates the
+    genomic location and the RNA-seq read-depths for these 6
+    microRNAs.}\label{fig:10}
 \end{figure}
@@
 \begin{figure}[ht]
-\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr4-3946041-3946705-miRNA-cluster.pdf}
-  \caption[]{microRNA cluster: mir-92\_2, mir-19\_1, mir-20\_1, mir-18\_1, mir-106\_1.}\label{fig:11}
+\includegraphics[width=0.99\textwidth]{figures/mir-17-I-X-cluster.pdf}
+  \caption[]{microRNA 17 I-X cluster: mir-363\_1, mir-92\_2,
+    mir-19\_1, mir-20\_1, mir-18\_1, mir-106\_1. The figure indicates
+    the genomic location and the RNA-seq read-depths for these 6
+    microRNAs.}\label{fig:11}
 \end{figure}
@@
 \begin{figure}[ht]
-\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr23-2507395-2508461-snoRNA-cluster.pdf}
-  \caption[]{H/ACA box snoRNA cluster: SNORA61, SNORA44, SNORA16.}\label{fig:12}
+\includegraphics[width=0.99\textwidth]{figures/chr23-2507395-2508461-snoRNA-cluster.pdf}
+  \caption[]{H/ACA box snoRNA cluster: SNORA61, SNORA44, SNORA16. The figure indicates
+    the genomic location and the RNA-seq read-depths for these 3
+    snoRNAs.}\label{fig:12}
 \end{figure}
@@
 \begin{figure}[ht]
-\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr27-4040511-4042052-tRNA-cluster.pdf}
-  \caption[]{Cysteine tRNA cluster.}\label{fig:13}
+\includegraphics[width=0.99\textwidth]{figures/chr27-4040511-4042052-tRNA-cluster.pdf}
+  \caption[]{Cysteine tRNA cluster. The figure indicates
+    the genomic location and the RNA-seq read-depths for these 5
+    transfer RNAs.}\label{fig:13}
 \end{figure}
@@
     Metazoan SRP and the Y RNA covariance models for human and chicken
-    paralogues/pseudogenes. }\label{fig:14}
+    paralogues/pseudogenes. Loci with high bit-scores are more likely
+    to be functional than those with low bit-scores (due to excess
+    variation that conflicts with the canonical RNA
+    models).}\label{fig:14}
 \end{figure}

### 2014-05-20 paper/manuscript.tex
Added accessions and a tree to the Sup. Mat.

@@
 
-Many of the lncRNAs have been associated with cancer, sparking a minor
-review industry \cite{Prensner:2011,Spizzo:2012}. Three examples of
-these that are also conserved in the birds are described below.
+The majority ($>80\%$) of genome-wide association studies of cancer
+identify loci outside of protein-coding genes
+\cite{Cheetham:2013}. Many of these are loci are now known to be
+transcribed into lncRNAs. Furthermore, many lncRNAs are differentially
+expressed in tumorous tissues \cite{Chan:2002}, suggesting further
+mechanistic links with the aberrant gene expression associated cancer
+progression. In our work we have identified three examples of these
+that are also conserved in the birds are described below.
 
@@
 Magnus Alm Rosenblad (Univ. of Gothenburg) and Eric Nawrocki (HHMI
-Janelia Farm) for useful discussions.
+Janelia Farm) for useful discussions. Matthew Walters for assistance
+with figures.
 

### 2014-05-20 paper/supplementary-material.tex
Added accessions and a tree to the Sup. Mat.

@@
  \begin{figure}[ht]
-   \includegraphics[width=0.5\textwidth]{figures/figure2.pdf}
+   \includegraphics[width=0.75\textwidth]{figures/figure2.pdf}
 \caption{Additional homologs of some sparsely represented RNA families were
@@
 %{\bf WRITE MORE HERE!}
-
-{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
-\bibliographystyle{bmc_article} % Style BST file
- \bibliography{bird} } % Bibliography file (usually '*.bib' )
 
@@
 
+\clearpage
+\newpage
+
+\begin{figure}[ht]
+\includegraphics[width=0.9\textwidth]{figures/phylo-tree.pdf}
+  \caption[]{The phylogenetic relationships between the Avian species
+    used in this study. This figure has been reproduced, with permission from \cite{birds:14a}.}\label{fig:15}
+\end{figure}
+
+\clearpage
+\newpage
+
+\subsection*{Database accessions}
+
+The NCBI BioProject/SRA and Study IDs for the genomes used in this study
+are are listed below.
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|l|l|l|}
+    \hline
+Species & BioProject ID & SRA ID & Study ID \\
+    \hline
+\emph{Chaetura pelagica} & PRJNA210808  & SRA092327 & SRP026688\\
+\emph{Calypte anna} & PRJNA212866 & SRA096094 & SRP028275\\
+\emph{Charadrius vociferus} & PRJNA212867 & SRA096158 & SRP028286\\
+\emph{Corvus brachyrhynchos} & PRJNA212869 & SRA096200 & SRP028317\\
+\emph{Cuculus canorus} & PRJNA212870 & SRA096365 & SRP028349\\
+\emph{Manacus vitellinus} & PRJNA212872 & SRA096507 & SRP028393\\
+\emph{Ophisthocomus hoazin} & PRJNA212873 & SRA096539 & SRP028409\\
+\emph{Picoides pubescens} & PRJNA212874 & SRA097131 & SRP028625\\
+\emph{Struthio camelus} & PRJNA212875 & SRA097407 & SRP028745\\
+\emph{Tinamus guttatus} & PRJNA212876 & SRA097796 & SRP028753\\
+\emph{Acanthisitta chloris} & PRJNA212877 & SRA097960 & SRP028832\\
+\emph{Apaloderma vittatum} & PRJNA212878 & SRA097967 & SRP028834\\
+\emph{Balearica regulorum} & PRJNA212879 & SRA097970 & SRP028839\\
+\emph{Buceros rhinoceros} & PRJNA212887 & SRA097991 & SRP028845\\
+\emph{Antrostomus carolinensis} & PRJNA212888 & SRA098079 & SRP028883\\
+\emph{Cariama cristata} & PRJNA212889 & SRA098089 & SRP028884\\
+\emph{Cathartes aura} & PRJNA212890 & SRA098145 & SRP028913\\
+\emph{Chlamydotis macqueenii} & PRJNA212891 & SRA098203 & SRP028950\\
+\emph{Colius striatus} & PRJNA212892 & SRA098342 & SRP028965\\
+\emph{Eurypyga helias} & PRJNA212893 & SRA098749 & SRP029147\\
+\emph{Fulmarus glacialis} & PRJNA212894 & SRA098806 & SRP029180\\
+\emph{Gavia stellata} & PRJNA212895 & SRA098829 & SRP029187\\
+\emph{Haliaeetus albicilla} & PRJNA212896 & SRA098868 & SRP029203\\
+\emph{Haliaeetus leucocephalus} & PRJNA237821 & SRX475899 & SRP038924\\
+                         & SRX475900   &           &          \\
+                         & SRX475901   &           &          \\
+                         & SRX475902   &           &          \\
+\emph{Leptosomus discolor} & PRJNA212897 & SRA098894 & SRP029206\\
+\emph{Merops nubicus} & PRJNA212898 & SRA099305 & SRP029278\\
+\emph{Mesitornis unicolor} & PRJNA212899 & SRA099409 & SRP029309\\
+\emph{Nestor notabilis} & PRJNA212900 & SRA099410 & SRP029311\\
+\emph{Pelecanus crispus} & PRJNA212901 & SRA099411 & SRP029331\\
+\emph{Phaethon lepturus} & PRJNA212902 & SRA099412 & SRP029342\\
+\emph{Phalacrocorax carbo} & PRJNA212903 & SRA099413 & SRP029344\\
+\emph{Phoenicopterus ruber} & PRJNA212904 & SRA099414 & SRP029345\\
+\emph{Podiceps cristatus} & PRJNA212905 & SRA099415 & SRP029346\\
+\emph{Pterocles gutturalis} & PRJNA212906 & SRA099416 & SRP029347\\
+\emph{Tauraco erythrolophus} & PRJNA212908 & SRA099418 & SRP029348\\
+\emph{Tyto alba} & PRJNA212909 & SRA099419 & SRP029349\\
+\emph{Nipponia nippon} & PRJNA232572 & SRA122361 & SRP035852\\
+\emph{Egretta garzetta} & PRJNA232959 & SRA123137 & SRP035853\\
+\hline
+    \end{tabular}
+  \end{center}
+  \caption{The NCBI BioProject/SRA and Study IDs used in this study. }
+\end{table}
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|l|l|l|}
+    \hline
+Species & BioProject ID & SRA ID & Study ID \\
+    \hline
+\multicolumn{4}{|l|}{Genomes released before}\\
+\hline
+\emph{Aptenodytes forsteri} & PRJNA235982 & SRA129317 & SRP035855\\
+\emph{Pygoscelis adeliae} & PRJNA235983 & SRA129318 & SRP035856\\
+\emph{Gallus gallus} & PRJNA13342 & SRA030184 & SRP005856\\
+\emph{Taeniopygia guttata} & PRJNA17289 & SRA010067 & SRP001389\\
+\emph{Meleagris gallopavo} & PRJNA42129 & Unknown & Unknown\\
+\emph{Melopsittacus undulatus} & PRJEB1588 & ERA200248 & ERP002324\\
+\emph{Anas platyrhynchos} & PRJNA46621 & SRA010308 & SRP001571\\
+\emph{Columba livia} & PRJNA167554 & SRA054954 & SRP013894\\
+\emph{Falco peregrinus} & PRJNA159791 & SRA055082 & SRP013939\\
+\emph{Geospiza fortis} & PRJNA156703 & SRA051234 & SRP011940\\
+\hline
+\multicolumn{4}{|l|}{Outgroups}\\
+\hline
+\emph{Homo sapiens} & & & \\
+\emph{Alligator mississippiensis} & & & \\
+\emph{Chelonia mydas} & & & \\
+\hline
+\multicolumn{4}{|l|}{RNA-seq data}\\
+\hline
+\emph{Gallus gallus} & PRJNA204941 & NA & NA \\
+\emph{Gallus gallus} & NA & & SRP041863 \\
+\hline
+    \end{tabular}
+  \end{center}
+  \caption{The NCBI BioProject/SRA and Study IDs for the previously
+    published genomes used in this study.}
+\end{table}
+
+{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
+\bibliographystyle{bmc_article} % Style BST file
+ \bibliography{bird} } % Bibliography file (usually '*.bib' )
+
 \end{bmcformat}

### 2014-05-22 paper/supplementary-material.tex
Updated supplementary images.

@@
  \begin{figure}[ht]
-   \includegraphics[width=0.75\textwidth]{figures/figure2.pdf}
+   \includegraphics[width=0.95\textwidth]{figures/figure2.pdf}
 \caption{Additional homologs of some sparsely represented RNA families were
@@
 \begin{figure}[ht]
+\includegraphics[width=0.99\textwidth]{figures/HOTAIR_loci_expression.pdf}
+\includegraphics[width=0.99\textwidth]{figures/HOTAIRM1_2.pdf}
+  \caption[]{Above: Expression of the chicken HOTAIRM1 (RF01976) locus
+    and the surrounding genes: HOXA1, HOXA2, and HOXA3. Below: The
+    predicted secondary structure of the Avian HOTAIRM1 domain.}\label{fig:15}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.99\textwidth]{figures/HOTAIR_loci_synteny.pdf}
+  \caption[]{The preservation of gene order (synteny) surrounding the
+    HOTAIRM1 (RF01976) locus across the Avian and other Vertebrate
+    lineages.}\label{fig:16}
+\end{figure}
+
+\clearpage
+\newpage
+
+\begin{table}[h!]
+  \begin{center}
+    \begin{tabular}{|l|l|r|c|}
+    \hline
+RNA family (Rfam ID) & Chromosome & Coordinates & Strand \\
+    \hline
+\multicolumn{4}{|l|}{Macro-chromosomes:}\\
+RNase MRP (RF00030)     & chr4 & 39276202-39276488 & -\\
+\multicolumn{4}{|l|}{Micro-chromosomes:}\\ 
+U4atac snRNA (RF00618)  & chr7 &	25702714-25702831 & +\\
+RNase P RNA (RF00009)   & chr8 & 6609347-6609668   & -\\
+Telomerase RNA (RF00024)& chr9 & 19429028-19429084 & +\\
+Vault RNA (RF00006)     & chr13& 380747-380840     & +\\
+U11 snRNA (RF00548)     & chr23& 2216932-2217058   & -\\
+\hline
+    \end{tabular}
+  \end{center}
+  \caption{The location of ``missing'' RNA families in the chicken genome (galGal4). }
+\end{table}
+
+\clearpage
+\newpage
+
+\begin{figure}[ht]
 \includegraphics[width=0.9\textwidth]{figures/phylo-tree.pdf}
   \caption[]{The phylogenetic relationships between the Avian species
-    used in this study. This figure has been reproduced, with permission from \cite{birds:14a}.}\label{fig:15}
+    used in this study. This figure has been reproduced, with permission from \cite{birds:14a}.}\label{fig:17}
 \end{figure}
@@
 
-The NCBI BioProject/SRA and Study IDs for the genomes used in this study
+The NCBI BioProject, SRA and Study IDs for the genomes used in this study
 are are listed below.
@@
 \emph{Pygoscelis adeliae} & PRJNA235983 & SRA129318 & SRP035856\\
-\emph{Gallus gallus} & PRJNA13342 & SRA030184 & SRP005856\\
+\emph{Gallus gallus} & PRJNA13342 & SRA030184 & SRP005856 (galGal4)\\
 \emph{Taeniopygia guttata} & PRJNA17289 & SRA010067 & SRP001389\\
@@
 \hline
-\emph{Homo sapiens} & & & \\
-\emph{Alligator mississippiensis} & & & \\
-\emph{Chelonia mydas} & & & \\
+\emph{Homo sapiens} & & & HG19/GRCh37\\
+\emph{Alligator mississippiensis} & & & allMis1\\
+\emph{Chelonia mydas} & PRJNA104937 & & \\
 \hline
@@
 \emph{Gallus gallus} & PRJNA204941 & NA & NA \\
-\emph{Gallus gallus} & NA & & SRP041863 \\
+\emph{Gallus gallus} & NA & NA & SRP041863 \\
 \hline

### 2014-05-23 paper/manuscript.tex
Updated supplementary images.
Updated intro.

@@
 developmental stages and tissues types. The costs of this approach are
-still prohibitive. Consequently, for this work we have concentrated on
-bioinformatic approaches. Primarily, homology based methods, namely
-covariance models (CMs). These remain state of the art for ncRNA
-analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
+still prohibitive in terms of researcher-time and
+finances. Consequently, for this work we have concentrated on
+bioinformatic approaches. However, we do validate the majority of
+these predictions using RNA-seq. Primarily, homology based methods,
+namely covariance models (CMs). These remain state of the art for
+ncRNA bioinformatic analyses
+\cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
 established sensitivity and specificity rates \cite{Freyhult:2007}.
@@
 that the CM approaches out-perform alternative methods
-\cite{Freyhult:2007}, although its sensitivity is limited for the most
+\cite{Freyhult:2007}, although its sensitivity can be limited for
 rapidly evolving families such as vault RNAs or telomerase RNA
@@
 
-There is substantial gain and loss of lncRNAs
-\cite{Cabili:2011,Kutter:2012} and more generally ncRNA loci over
-evolutionary time \cite{Hoeppner:2012}. It is difficult to assess how
-many of these observations are due to limited bioinformatic sequence
-alignment tools (these generally fail align correctly below 60-50\%
-sequence identity \cite{Gardner:2005:Nucleic-Acids-Res:15860779}) or
-due to genuine gains and losses. Nevertheless, sequence conservation,
-generally speaking, provides useful evidence for gene and function
-conservation.
+There is substantial gain and loss of lncRNAs and other ncRNA loci
+over evolutionary time
+\cite{Cabili:2011,Kutter:2012,Hoeppner:2012}. It is difficult to
+assess how many of these ``gains'' and ``losses'' are due to limited
+bioinformatic sequence alignment tools (these generally fail align
+correctly below 60-50\% sequence identity
+\cite{Gardner:2005:Nucleic-Acids-Res:15860779}) or due to genuine
+gains and losses. Nevertheless, sequence conservation, generally
+speaking, provides useful evidence for gene and function conservation.
 
@@
 support a high degree of evolutionary conservation for the entire
-lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1, SOX2OT and
-ST7-OT3 lncRNAs have multiple, well conserved RNA-domains (See
-Figure~\ref{fig:1}).
+lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1,
+SOX2OT and ST7-OT3 lncRNAs have multiple, well conserved RNA-domains
+(See Figure~\ref{fig:1}). The syntenic ordering of these seven lncRNAs
+and the flanking genes are also preserved between the human and
+chicken genomes (data not shown). We illustrate this in detail for the
+HOTAIRM1 lncRNA (See Supplemental Figures~13\&14).
 

### 2014-05-23 paper/manuscript.tex
Updated the well-conserved lncRNA discussion.

@@
 tool, tRNAscan-SE and microRNAs from miRBase.  We show a number of
-lncRNA associated loci are conserved between birds and mammals,
+lncRNA-associated loci are conserved between birds and mammals,
 including several intriguing cases where the reported mammalian lncRNA
@@
 RNAs that have been implicated in a multitude of functional processes
-\cite{Rinn:2007,Chow:2005,Guttman:2009,Ulitsky:2013}. These RNAs have largely been
-characterised in mammalian species, particularly human and
-mouse. Consequently, we generally do not expect these to be conserved
-outside of mammalia. Notable examples include Xist \cite{Duret:2006}
-and H19 \cite{Smits:2008}.  There is emerging evidence for the
-conservation of ``Mammalian'' lncRNAs in other Vertebrates
+\cite{Rinn:2007,Chow:2005,Guttman:2009,Ulitsky:2013}. These RNAs have
+largely been characterised in mammalian species, particularly human
+and mouse. Consequently, we generally do not expect these to be
+conserved outside of Mammals. Notable examples include Xist
+\cite{Duret:2006} and H19 \cite{Smits:2008}.  There is emerging
+evidence for the conservation of ``mammalian'' lncRNAs in Vertebrates
 \cite{Chodroff:2010,Ulitsky:2011}), however, like most lncRNAs, the
 function of these lncRNAs remains largely unknown. Here, we show the
-conservation of several well-characterised lncRNAs of known function
-in humans.
+conservation of several lncRNAs that have been well-characterised in
+humans.
 

### 2014-05-23 paper/manuscript.tex
Addressing Reviewer comments.

@@
 
-The conservation of these ``human'' lncRNAs among birds suggests they may 
-also be functional in birds but what these
-functions is not immediately obvious. For example, PART1 and
-PCA3 are both described as prostate-specific lncRNAs that play a role
-in the human androgen-receptor pathway
+The conservation of these ``human'' lncRNAs among birds suggests they
+may also be functional in birds. But what these functions are, is not
+immediately obvious. For example, PART1 and PCA3 are both described as
+prostate-specific lncRNAs that play a role in the human
+androgen-receptor pathway
 \cite{Bussemakers:1999,Lin:2000,Ferreira:2012}. Birds lack a prostate
-but both males and females express the androgen receptor (AR or NR3C4) in gonadal
-and non- gonadal tissue
+but both males and females express the androgen receptor (AR or NR3C4)
+in gonadal and non- gonadal tissue
 \cite{Yoshimura:1993,Veney:2004,Fuxjager:2012,Leska:2012}. Thus, we
@@
 conservation or expansion within the avian lineage. The only exception
-being the snoRNA, SNORD93 which has 92 copies in the tinamou
+being the snoRNA, SNORD93. SNORD93 has 92 copies in the tinamou
 genome, whereas it only has 1-2 copies in all the other vertebrate
@@
 $\sim900$ and $\sim580$ copies in the human, turtle and alligator
-genomes, the average copy numbers of these drop to $\sim280$ and
+genomes, the average copy-numbers of these drop to $\sim280$ and
 $\sim100$ copies in the avian genomes. In addition to reduction in
-copy-number, the complete absence of several otherwise ubiquitous
-vertebrate ncRNAs in the avian lineage are indicative of genuine gene
-loss.
+copy-number, the absence of several, otherwise ubiquitous vertebrate
+ncRNAs, in the avian lineage are suggestive of genuine gene loss.
 
@@
 conserved Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most
-cases consistent with breaks in the available sequence assemblies.  In
-several genomes we observe one or a few additional Y RNA homologs
-unlinked to the canonical Y RNA cluster. These sequences can be
-identified unambiguously as derived members of one of the three
-ancestral paralog groups, they almost always fit less well to the
-consensus (as measured by the CM bit score of paralog group specific
-covariance models) than the paralog linked to cluster, and there is no
-indication that any of these additional copies is evolutionarily
-conserved over longer time scales. We therefore suggest that most or
-all of these interspersed copies are in fact pseudogenes.
+cases consistent with breaks (e.g. ends of contigs) in the available
+sequence assemblies.  In several genomes we observe one or a few
+additional Y RNA homologs unlinked to the canonical Y RNA
+cluster. These sequences can be identified unambiguously as derived
+members of one of the three ancestral paralog groups, they almost
+always fit less well to the consensus (as measured by the CM bit score
+of paralog group specific covariance models) than the paralog linked
+to cluster, and there is no indication that any of these additional
+copies is evolutionarily conserved over longer time scales. We
+therefore suggest that most or all of these interspersed copies are in
+fact pseudogenes.
 
@@
 
-Heatmaps showing the presence/absence and approximate genomic copy
-number of ``unusually, well conserved RNAs'' (particularly the
-lncRNAs) on the left and families that have been identified as
-RNA losses, divergence or missing data. In several cases
-functionally related families have also been included, e.g. the RNA
-components of the major and minor spliceosomes: U1, U2, U4, U5 and U6;
-and U11, U12, U4atac, U5 and U6atac, respectively.\label{fig:1}
+Heatmaps showing the presence/absence and approximate genomic
+copy-number of ``unusually, well conserved RNAs'' (particularly the
+lncRNAs) on the left and families that have been identified as RNA
+losses, divergence or missing data. In several cases functionally
+related families have also been included, e.g. the RNA components of
+the major and minor spliceosomes: U1, U2, U4, U5 and U6; and U11, U12,
+U4atac, U5 and U6atac, respectively.\label{fig:1}
 

### 2014-05-23 paper/supplementary-material.tex
Addressing Reviewer comments.

@@
 patterns of the major spliceosomal RNAs they are all well represented
-in these genome sequences. Except for the U4 RNA in cormorant and the
-U5 RNA in the bee eater. Again these two genomes are low coverage
-suggesting these genes weren't captured in the current assembly. The
-minor spliceosomal RNAs are more interesting, the U4atac and U11
-snRNAs show widespread patterns of loss, even in some of the high
-coverage genomes. These RNAs are frequently missed in bioinformatic
-screens. Indicating either frequent loss \cite{Davila_Lopez:2008} or
-sequences that have diverged beyond the ability of detection by
-covariance models \cite{Marz:2008}.
+in these genome sequences. The exceptions to this observation are the
+U4 RNA in cormorant and the U5 RNA in the bee eater which are both
+missing. These two genomes are low coverage, suggesting these
+genes weren't captured in the current assembly. The minor spliceosomal
+RNAs are more interesting, the U4atac and U11 snRNAs show widespread
+patterns of loss, even in some of the high coverage genomes. These
+RNAs are frequently missed in bioinformatic screens. Indicating either
+frequent loss \cite{Davila_Lopez:2008} or sequences that have diverged
+beyond the ability of detection by covariance models \cite{Marz:2008}.
 
@@
 eight are located in the tinamou genome, these are located on the same
-scaffold and are within 2 KB of each other. Implying that SNHG1 is
+scaffold and are within 2 KB of each other. This implies that SNHG1 is
 conserved in the tinamou. Loci with four of the eight snoRNAs can be

### 2014-05-23 paper/supplementary-material.tex
Addressing more Reviewer comments.

@@
 
-The the RNA components of RNase P and RNase MRP also appear to have
+The RNA components of RNase P and RNase MRP also appear to have
 undergone dramatic losses within the bird lineage. RNase P is required
@@
 alligator and turtle genomes within a 3-4 KB locus, yet these have
-largely been lost in the birds. Notable exceptions are five of the
-eight are located in the tinamou genome, these are located on the same
-scaffold and are within 2 KB of each other. This implies that SNHG1 is
+largely been lost in the birds. However, five of the eight snoRNAs are
+located in the tinamou genome. These are located on the same scaffold
+and are within 2 KB of each other. This implies that SNHG1 is
 conserved in the tinamou. Loci with four of the eight snoRNAs can be
 found in zebrafinch, ground-finch, and bald eagle. Still, three of the
-eight are located in the ostrich, crow, and cockoo genomes, again
+eight are located in the ostrich, crow, and cuckoo genomes, again
 within 2 KB of each other on the same scaffolds. This complex pattern
@@
 the ground-finch and bald eagle retain a single SNORD33 and the
-zebrafinch and seriema retain a single SNORD34 (See
+zebrafinch and seriema retain a single SNORD34 (See Supplemental
 Figure~\ref{fig:3}).  In human these snoRNAs are intronic to the host
@@
 rifleman and zebrafinch avian genomes (data not shown). Therefore the
-conservation patterns of the RPL13A host gene and corresponding
-intronic snoRNAs largely mimic each other, each independently supports
-a pattern of loss of the RPL13A gene and the intronic snoRNAs that it
-hosts in the bird genomes.
+RPL13A gene and corresponding intronic snoRNAs show the same
+conservation pattern. This supports a pattern of loss of the RPL13A
+gene and the intronic snoRNAs that it hosts in the bird genomes.
 

### 2014-05-23 paper/supplementary-material.tex
Spell check.

@@
 \emph{S. cerevisiae} \cite{Lestrade:2006}. Of these, 45 snoRNA
-families are conserved in the bird dataset.
+families are conserved in the bird data set.
 Over a third of the apparent losses of the yeast-human conserved
@@
 these genes are chicken-specific. This is consistent with previous
-reports of large number of speciesspecific microRNAs in all animals,
+reports of large number of species specific microRNAs in all animals,
 and supports the view of fast microRNA turnover during animal
@@
 The simplest case are the selenocysteine tRNAs. Here, tRNAscan is
-tuned for specificity and thus misses several occurrances that are
+tuned for specificity and thus misses several occurrences that are
 easily found by \texttt{blastn} with $E\le 10^{-30}$. In some cases
@@
   \includegraphics[height=0.95\textheight]{figures/snoRNA.pdf}
-  \caption[]{Heatmaps showing the prescence/abscence and approximate
+  \caption[]{Heatmaps showing the presence/absence and approximate
     copy number of {\bf miRNA} families on the right and {\bf snoRNA} families on
@@
   \includegraphics[width=0.85\textwidth]{figures/tRNA.pdf}
-  \caption[]{Heatmaps showing the prescence/abscence and approximate
+  \caption[]{Heatmaps showing the presence/absence and approximate
     copy number of {\bf tRNA} families.}\label{fig:5}

### 2014-05-26 paper/manuscript.tex
Affiliations edited

@@
 \email{Paul P.\ Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
-Mario Fasold$^5$
-\email{mario@bierdepot.bioinf.uni-leipzig.de},
+Mario Fasold$^{5,6}$
+\email{mario@bioinf.uni-leipzig.de},
 Sarah W.\ Burge$^3$
@@
 Stephanie Kehr$^5$
-\email{steffi@bierdepot.bioinf.uni-leipzig.de},
+\email{steffi@bioinf.uni-leipzig.de},
 Tammy E.\ Steeves$^1$
@@
 \iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany
+\iid(6) ecSeq Bioinformatics, Brandvorwerkstr.43, D-04275 Leipzig, Germany
 }

### 2014-05-28 paper/supplementary-material.tex
Added common names to NCBI ID table.

@@
     \hline
-\multicolumn{4}{|l|}{Macro-chromosomes:}\\
+\multicolumn{4}{|l|}{{\bf Macro-chromosomes:}}\\
 RNase MRP (RF00030)     & chr4 & 39276202-39276488 & -\\
-\multicolumn{4}{|l|}{Micro-chromosomes:}\\ 
+\multicolumn{4}{|l|}{{\bf Micro-chromosomes:}}\\ 
 U4atac snRNA (RF00618)  & chr7 &	25702714-25702831 & +\\
@@
   \begin{center}
-    \begin{tabular}{|l|l|l|l|}
+    \begin{tabular}{|l|l|l|l|l|}
     \hline
-Species & BioProject ID & SRA ID & Study ID \\
+Name & Species & BioProject ID & SRA ID & Study ID \\
     \hline
-\emph{Chaetura pelagica} & PRJNA210808  & SRA092327 & SRP026688\\
-\emph{Calypte anna} & PRJNA212866 & SRA096094 & SRP028275\\
-\emph{Charadrius vociferus} & PRJNA212867 & SRA096158 & SRP028286\\
-\emph{Corvus brachyrhynchos} & PRJNA212869 & SRA096200 & SRP028317\\
-\emph{Cuculus canorus} & PRJNA212870 & SRA096365 & SRP028349\\
-\emph{Manacus vitellinus} & PRJNA212872 & SRA096507 & SRP028393\\
-\emph{Ophisthocomus hoazin} & PRJNA212873 & SRA096539 & SRP028409\\
-\emph{Picoides pubescens} & PRJNA212874 & SRA097131 & SRP028625\\
-\emph{Struthio camelus} & PRJNA212875 & SRA097407 & SRP028745\\
-\emph{Tinamus guttatus} & PRJNA212876 & SRA097796 & SRP028753\\
-\emph{Acanthisitta chloris} & PRJNA212877 & SRA097960 & SRP028832\\
-\emph{Apaloderma vittatum} & PRJNA212878 & SRA097967 & SRP028834\\
-\emph{Balearica regulorum} & PRJNA212879 & SRA097970 & SRP028839\\
-\emph{Buceros rhinoceros} & PRJNA212887 & SRA097991 & SRP028845\\
-\emph{Antrostomus carolinensis} & PRJNA212888 & SRA098079 & SRP028883\\
-\emph{Cariama cristata} & PRJNA212889 & SRA098089 & SRP028884\\
-\emph{Cathartes aura} & PRJNA212890 & SRA098145 & SRP028913\\
-\emph{Chlamydotis macqueenii} & PRJNA212891 & SRA098203 & SRP028950\\
-\emph{Colius striatus} & PRJNA212892 & SRA098342 & SRP028965\\
-\emph{Eurypyga helias} & PRJNA212893 & SRA098749 & SRP029147\\
-\emph{Fulmarus glacialis} & PRJNA212894 & SRA098806 & SRP029180\\
-\emph{Gavia stellata} & PRJNA212895 & SRA098829 & SRP029187\\
-\emph{Haliaeetus albicilla} & PRJNA212896 & SRA098868 & SRP029203\\
-\emph{Haliaeetus leucocephalus} & PRJNA237821 & SRX475899 & SRP038924\\
-                         & SRX475900   &           &          \\
-                         & SRX475901   &           &          \\
-                         & SRX475902   &           &          \\
-\emph{Leptosomus discolor} & PRJNA212897 & SRA098894 & SRP029206\\
-\emph{Merops nubicus} & PRJNA212898 & SRA099305 & SRP029278\\
-\emph{Mesitornis unicolor} & PRJNA212899 & SRA099409 & SRP029309\\
-\emph{Nestor notabilis} & PRJNA212900 & SRA099410 & SRP029311\\
-\emph{Pelecanus crispus} & PRJNA212901 & SRA099411 & SRP029331\\
-\emph{Phaethon lepturus} & PRJNA212902 & SRA099412 & SRP029342\\
-\emph{Phalacrocorax carbo} & PRJNA212903 & SRA099413 & SRP029344\\
-\emph{Phoenicopterus ruber} & PRJNA212904 & SRA099414 & SRP029345\\
-\emph{Podiceps cristatus} & PRJNA212905 & SRA099415 & SRP029346\\
-\emph{Pterocles gutturalis} & PRJNA212906 & SRA099416 & SRP029347\\
-\emph{Tauraco erythrolophus} & PRJNA212908 & SRA099418 & SRP029348\\
-\emph{Tyto alba} & PRJNA212909 & SRA099419 & SRP029349\\
-\emph{Nipponia nippon} & PRJNA232572 & SRA122361 & SRP035852\\
-\emph{Egretta garzetta} & PRJNA232959 & SRA123137 & SRP035853\\
+Chimney Swift & \emph{Chaetura pelagica} & PRJNA210808  & SRA092327 & SRP026688\\
+Hummingbird & \emph{Calypte anna} & PRJNA212866 & SRA096094 & SRP028275\\
+Plover & \emph{Charadrius vociferus} & PRJNA212867 & SRA096158 & SRP028286\\
+Crow & \emph{Corvus brachyrhynchos} & PRJNA212869 & SRA096200 & SRP028317\\
+Cuckoo & \emph{Cuculus canorus} & PRJNA212870 & SRA096365 & SRP028349\\
+Manakin & \emph{Manacus vitellinus} & PRJNA212872 & SRA096507 & SRP028393\\
+Hoatzin & \emph{Ophisthocomus hoazin} & PRJNA212873 & SRA096539 & SRP028409\\
+Woodpecker & \emph{Picoides pubescens} & PRJNA212874 & SRA097131 & SRP028625\\
+Ostrich & \emph{Struthio camelus} & PRJNA212875 & SRA097407 & SRP028745\\
+Tinamou & \emph{Tinamus guttatus} & PRJNA212876 & SRA097796 & SRP028753\\
+Rifleman & \emph{Acanthisitta chloris} & PRJNA212877 & SRA097960 & SRP028832\\
+Trogon & \emph{Apaloderma vittatum} & PRJNA212878 & SRA097967 & SRP028834\\
+Crane & \emph{Balearica regulorum} & PRJNA212879 & SRA097970 & SRP028839\\
+Rhinoceros Hornbill & \emph{Buceros rhinoceros} & PRJNA212887 & SRA097991 & SRP028845\\
+Nightjar & \emph{Antrostomus carolinensis} & PRJNA212888 & SRA098079 & SRP028883\\
+Seriema & \emph{Cariama cristata} & PRJNA212889 & SRA098089 & SRP028884\\
+Turkey Vulture & \emph{Cathartes aura} & PRJNA212890 & SRA098145 & SRP028913\\
+Bustard & \emph{Chlamydotis macqueenii} & PRJNA212891 & SRA098203 & SRP028950\\
+Mousebird & \emph{Colius striatus} & PRJNA212892 & SRA098342 & SRP028965\\
+Sunbittern & \emph{Eurypyga helias} & PRJNA212893 & SRA098749 & SRP029147\\
+Northern Fulmar & \emph{Fulmarus glacialis} & PRJNA212894 & SRA098806 & SRP029180\\
+Red-throated Loon & \emph{Gavia stellata} & PRJNA212895 & SRA098829 & SRP029187\\
+White-tailed Eagle & \emph{Haliaeetus albicilla} & PRJNA212896 & SRA098868 & SRP029203\\
+Bald Eagle & \emph{Haliaeetus leucocephalus} & PRJNA237821 & SRX475899 & SRP038924\\
+           &              & SRX475900   &           &          \\
+           &              & SRX475901   &           &          \\
+           &              & SRX475902   &           &          \\
+Cuckoo Roller & \emph{Leptosomus discolor} & PRJNA212897 & SRA098894 & SRP029206\\
+Bee-eater & \emph{Merops nubicus} & PRJNA212898 & SRA099305 & SRP029278\\
+Brown Mesite & \emph{Mesitornis unicolor} & PRJNA212899 & SRA099409 & SRP029309\\
+Kea & \emph{Nestor notabilis} & PRJNA212900 & SRA099410 & SRP029311\\
+Pelican & \emph{Pelecanus crispus} & PRJNA212901 & SRA099411 & SRP029331\\
+Tropicbird & \emph{Phaethon lepturus} & PRJNA212902 & SRA099412 & SRP029342\\
+Cormorant & \emph{Phalacrocorax carbo} & PRJNA212903 & SRA099413 & SRP029344\\
+Flamingo & \emph{Phoenicopterus ruber} & PRJNA212904 & SRA099414 & SRP029345\\
+Grebe & \emph{Podiceps cristatus} & PRJNA212905 & SRA099415 & SRP029346\\
+Sandgrouse & \emph{Pterocles gutturalis} & PRJNA212906 & SRA099416 & SRP029347\\
+Turaco & \emph{Tauraco erythrolophus} & PRJNA212908 & SRA099418 & SRP029348\\
+Barn Owl & \emph{Tyto alba} & PRJNA212909 & SRA099419 & SRP029349\\
+Crested Ibis & \emph{Nipponia nippon} & PRJNA232572 & SRA122361 & SRP035852\\
+Little Egret & \emph{Egretta garzetta} & PRJNA232959 & SRA123137 & SRP035853\\
 \hline
@@
   \begin{center}
-    \begin{tabular}{|l|l|l|l|}
+    \begin{tabular}{|l|l|l|l|l|}
     \hline
@@
     \hline
-\multicolumn{4}{|l|}{Genomes released before}\\
+Emperor Penguin & \emph{Aptenodytes forsteri} & PRJNA235982 & SRA129317 & SRP035855\\
+Adelie Penguin & \emph{Pygoscelis adeliae} & PRJNA235983 & SRA129318 & SRP035856\\
+Chicken & \emph{Gallus gallus} & PRJNA13342 & SRA030184 & SRP005856 (galGal4)\\
+Zebra Finch & \emph{Taeniopygia guttata} & PRJNA17289 & SRA010067 & SRP001389\\
+Turkey & \emph{Meleagris gallopavo} & PRJNA42129 & Unknown & Unknown\\
+Budgerigar & \emph{Melopsittacus undulatus} & PRJEB1588 & ERA200248 & ERP002324\\
+Mallard & \emph{Anas platyrhynchos} & PRJNA46621 & SRA010308 & SRP001571\\
+Rock Pigeon & \emph{Columba livia} & PRJNA167554 & SRA054954 & SRP013894\\
+Peregrine Falcon & \emph{Falco peregrinus} & PRJNA159791 & SRA055082 & SRP013939\\
+Medium Ground-finch & \emph{Geospiza fortis} & PRJNA156703 & SRA051234 & SRP011940\\
 \hline
-\emph{Aptenodytes forsteri} & PRJNA235982 & SRA129317 & SRP035855\\
-\emph{Pygoscelis adeliae} & PRJNA235983 & SRA129318 & SRP035856\\
-\emph{Gallus gallus} & PRJNA13342 & SRA030184 & SRP005856 (galGal4)\\
-\emph{Taeniopygia guttata} & PRJNA17289 & SRA010067 & SRP001389\\
-\emph{Meleagris gallopavo} & PRJNA42129 & Unknown & Unknown\\
-\emph{Melopsittacus undulatus} & PRJEB1588 & ERA200248 & ERP002324\\
-\emph{Anas platyrhynchos} & PRJNA46621 & SRA010308 & SRP001571\\
-\emph{Columba livia} & PRJNA167554 & SRA054954 & SRP013894\\
-\emph{Falco peregrinus} & PRJNA159791 & SRA055082 & SRP013939\\
-\emph{Geospiza fortis} & PRJNA156703 & SRA051234 & SRP011940\\
+\multicolumn{5}{|l|}{Outgroups}\\
 \hline
-\multicolumn{4}{|l|}{Outgroups}\\
+Human & \emph{Homo sapiens} & & & HG19/GRCh37\\
+Alligator & \emph{Alligator mississippiensis} & & & allMis1\\
+Green Turtle & \emph{Chelonia mydas} & PRJNA104937 & & \\
 \hline
-\emph{Homo sapiens} & & & HG19/GRCh37\\
-\emph{Alligator mississippiensis} & & & allMis1\\
-\emph{Chelonia mydas} & PRJNA104937 & & \\
+\multicolumn{5}{|l|}{RNA-seq data}\\
 \hline
-\multicolumn{4}{|l|}{RNA-seq data}\\
-\hline
-\emph{Gallus gallus} & PRJNA204941 & NA & NA \\
-\emph{Gallus gallus} & NA & NA & SRP041863 \\
+Chicken & \emph{Gallus gallus} & PRJNA204941 & NA & NA \\
+Chicken & \emph{Gallus gallus} & NA & NA & SRP041863 \\
 \hline

### 2014-05-28 paper/manuscript.tex

@@
 \iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
-\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany
-\iid(6) ecSeq Bioinformatics, Brandvorwerkstr.43, D-04275 Leipzig, Germany
+\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany.
+\iid(6) ecSeq Bioinformatics, Brandvorwerkstr.43, D-04275 Leipzig, Germany.
 }

### 2014-05-29 paper/manuscript.tex
Added Tammy\47s edits.

@@
 these to genuine loss, divergence or missing data.  In particular, we
-show that many of these losses are due to issues with assembling
-microchromosomes using next-generation sequencing methods. These
-combined results illustrate the utility of applying homology-based
-methods for annotating novel vertebrate genomes.
+show that many of these losses are due to the challenges associated
+with assembling Avian microchromosomes. These combined results illustrate the utility of
+applying homology-based methods for annotating novel vertebrate
+genomes.
 \end{abstract}
@@
 \cite{cech2014noncoding}. However, in order to determine the full
-range of ncRNAs across multiple species we would require multiple RNA
+range of ncRNAs across multiple species we require multiple RNA
 fractions (e.g. long and short), in multiple species, in multiple
@@
 still prohibitive in terms of researcher-time and
-finances. Consequently, for this work we have concentrated on
-bioinformatic approaches. However, we do validate the majority of
-these predictions using RNA-seq. Primarily, homology based methods,
-namely covariance models (CMs). These remain state of the art for
-ncRNA bioinformatic analyses
-\cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
+finances. Consequently, in this study we concentrate on bioinformatic
+approaches, primarily homology based methods (i.e. covariance models
+(CMs)). However, we do validate the majority of these predictions
+using RNA-seq. The bioinformatic approaches we use remain state of the art for ncRNA bioinformatic
+analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
 established sensitivity and specificity rates \cite{Freyhult:2007}.
-The CM based approach for annotating ncRNAs in genomes requires
+For example, the CM based approach for annotating ncRNAs in genomes requires
 reliable alignments and consensus secondary structures of
@@
 The conservation of these ``human'' lncRNAs among birds suggests they
-may also be functional in birds. But what these functions are, is not
+may also be functional in birds. But what these functions may be is not
 immediately obvious. For example, PART1 and PCA3 are both described as
@@
 therefore suggest that most or all of these interspersed copies are in
-fact pseudogenes.
+fact pseudogenes (see below).
 
@@
 macrochromosome, and all remaining missing ncRNAs are found on
-microchromosomes (see Supplemental Table ?). A Fisher’s exact test
+microchromosomes (see Supplemental Table~1). A Fisher’s exact test
 showed that there is significantly more missing ncRNAs on
@@
 the major and minor spliceosomes: U1, U2, U4, U5 and U6; and U11, U12,
-U4atac, U5 and U6atac, respectively.\label{fig:1}
+U4atac, U5 and U6atac, respectively.
 
@@
    \includegraphics[width=0.95\textwidth]{figures/figure1.pdf}
+   \caption{}
+   \label{fig:1}
  \end{figure}
@@
 The threshold for determining expression was selected based upon a
-false positive rate of less than 10\%.\label{table:1}
+false positive rate of less than 10\%.
 
@@
 
+\begin{table}
 \begin{tabular}{|r|r|r|r|l|}
@@
 \end{tabular}
+\caption{}
+\label{table:1}
+\end{table}
 

### 2014-05-30 paper/manuscript.tex
Added Maria\47s edits.

@@
 supplement these annotations with predictions from the tRNA annotation
-tool, tRNAscan-SE and microRNAs from miRBase.  We show a number of
+tool, tRNAscan-SE and microRNAs from miRBase.  We show that a number of
 lncRNA-associated loci are conserved between birds and mammals,
@@
 finances. Consequently, in this study we concentrate on bioinformatic
-approaches, primarily homology based methods (i.e. covariance models
+approaches, primarily we use homology-based methods (i.e. covariance models
 (CMs)). However, we do validate the majority of these predictions
@@
 that the CM approaches out-perform alternative methods
-\cite{Freyhult:2007}, although its sensitivity can be limited for
+\cite{Freyhult:2007}, although their sensitivity can be limited for
 rapidly evolving families such as vault RNAs or telomerase RNA
@@
 yet its conservation across the vertebrates certainly implies a
-function (See Figure~\ref{fig:1}). Possibly at the DNA level, however,
-function at the RNA level cannot be ruled out at this stage.
+function (See Figure~\ref{fig:1}). We note that the function for this
+locus may be at the DNA level, however, function at the RNA level
+cannot be ruled out at this stage.
 
@@
 
-In order to determine to what extent the absence of these RNAs from
+In order to determine to what extent the absence of some ncRNAs from
 the infernal-based annotation is caused by sequence divergence beyond
@@
   fide} gene loss. A large fraction of our annotations have been
-confirmed using RNA-seq data, which also showed a 7-fold enriched of
+confirmed using RNA-seq data, which also showed a 7-fold enrichment of
 expression within our annotations relative to unannotated regions.
@@
 that using data from well studied lineages such as mammals can also
-result in quality annotations of sister taxa such as aves.
+result in quality annotations of sister taxa such as Aves.
 
@@
 the fact that the reported functions of some ncRNAs are
-mammal-specific, yet are also found in bird genomes.
+mammal-specific, yet these are also found in bird genomes.
 

### 2014-08-07 paper/manuscript.tex
Updated manuscript to address reviewer comments.

@@
 %\title{Wins and losses: Non-coding RNAs of bird genomes}
-\title{Conservation and losses of avian non-coding RNA loci}
+\title{Conservation and losses non-coding RNA associated loci in avian genomes}
 
@@
 Paul P.\ Gardner\correspondingauthor$^{1,2}$
-\email{Paul P.\ Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
+\email{Paul P.\ Gardner\correspondingauthor - paul.gardner@canterburmotifs/manuscript/y.ac.nz},
 Mario Fasold$^{5,6}$
@@
 supplement these annotations with predictions from the tRNA annotation
-tool, tRNAscan-SE and microRNAs from miRBase.  We show that a number of
-lncRNA-associated loci are conserved between birds and mammals,
-including several intriguing cases where the reported mammalian lncRNA
-function is not conserved in birds.  We also demonstrate extensive
-conservation of classical ncRNAs (e.g., tRNAs) and more recently
-discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds. Furthermore, we
-describe numerous ``losses'' of several RNA families, and attribute
-these to genuine loss, divergence or missing data.  In particular, we
-show that many of these losses are due to the challenges associated
-with assembling Avian microchromosomes. These combined results illustrate the utility of
+tool, tRNAscan-SE and microRNAs from miRBase.  We identify 34
+lncRNA-associated loci that are conserved between birds and mammals
+and validate 12 of these in chicken. These include several intriguing
+cases where the reported mammalian lncRNA function is not conserved in
+birds.  We also demonstrate extensive conservation of classical ncRNAs
+(e.g., tRNAs) and more recently discovered ncRNAs (e.g., snoRNAs and
+miRNAs) in birds. Furthermore, we describe numerous ``losses'' of
+several RNA families, and attribute these to genuine loss, divergence
+or missing data.  In particular, we show that many of these losses are
+due to the challenges associated with assembling Avian
+microchromosomes. These combined results illustrate the utility of
 applying homology-based methods for annotating novel vertebrate
@@
 \cite{birds:14,birds:14a,Huang:2013,Zhan:2013,Shapiro:2013,Howard:2013,Li:2014},
-provides an exciting opportunity to explore ncRNA conservation in
-unprecedented detail.
+provides an exciting opportunity to explore conservation of genomic
+loci that have been associated with ncRNAs in unprecedented detail.
 %{\bf MORE ON BIRD GENOMES. ESP. MACRO \& MICRO
@@
 
-There is substantial gain and loss of lncRNAs and other ncRNA loci
-over evolutionary time
+There is substantial gain and loss of lncRNAs and other ncRNA
+associated loci over evolutionary time
 \cite{Cabili:2011,Kutter:2012,Hoeppner:2012}. It is difficult to
@@
 \cite{Gardner:2005:Nucleic-Acids-Res:15860779}) or due to genuine
-gains and losses. Nevertheless, sequence conservation, generally
-speaking, provides useful evidence for gene and function conservation.
+gains and losses or data missing from the current
+genome assemblies. Nevertheless, sequence conservation, generally speaking,
+provides useful evidence for gene and function conservation.
 
@@
 snoRNAs (summarised in Table~\ref{table:1}). Out of necessity we have
-selected the most charismatic families for further discussion. These
+selected a modest number of families for further discussion. These
 include the lncRNAs that appear to be conserved between Mammals and
@@
 
-When analysing the RNA-domain annotations it is striking that many of
-the lncRNAs with multiple RNA-domains are consistently preserved in
-the birds. The annotations of these domains lie in the same genomic
-region, in the same order as in the mammalian homologs. Thus they
-support a high degree of evolutionary conservation for the entire
-lncRNA. In particular the HOXA11-AS1, PART1, PCA3, RMST, Six3os1,
-SOX2OT and ST7-OT3 lncRNAs have multiple, well conserved RNA-domains
-(See Figure~\ref{fig:1}). The syntenic ordering of these seven lncRNAs
-and the flanking genes are also preserved between the human and
-chicken genomes (data not shown). We illustrate this in detail for the
-HOTAIRM1 lncRNA (See Supplemental Figures~13\&14).
+When analysing the RNA-domain annotations it is striking that the
+order (synteny) of many of the lncRNAs with multiple RNA-domains are
+consistently preserved in the birds. The annotations of these domains
+lie in the same genomic region, in the same order as in the mammalian
+homologs. Thus they support a high degree of evolutionary conservation
+for the entire lncRNA. In particular the HOXA11-AS1, PART1, PCA3,
+RMST, Six3os1, SOX2OT and ST7-OT3 lncRNAs have multiple, well
+conserved RNA-domains (See Figure~\ref{fig:1}). The syntenic ordering
+of these seven lncRNAs and the flanking genes are also preserved
+between the human and chicken genomes (data not shown). We illustrate
+this in detail for the HOTAIRM1 lncRNA (See Supplemental
+Figures~13\&14).
 
